---
title: "Plagiotremus"
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Cameron A.J. Walsh:
      institute:
        - himb
      equal_contributor: yes
  - Randall K. Kosaki:
      institute:
        - pmnm
  - Brian W. Bowen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - pmnm: Papahānaumokuākea Marine National Monument, National Oceanic and Atmospheric Administration
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
link-citations: true
linkReferences: true
tblPrefix: Table
figPrefix:
  - "Fig."
  - "Figs."
chapters: True
chaptersDepth: 1
chapDelim: ""
lofTitle: |
  # Figure legends {-}
lofItemTitle: |
  Fig.
lofItemTemplate: |
  $$lofItemTitle$$ $$i$$$$listItemTitleDelim$$ $$t$$
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
fig_caption: true
---

### Keywords {-}
<div custom-style="mono">blennies, endemism, Hawai&#x02BB;i, phylogeography, phylogenetics</div>

\newpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path="output/figures/",
  dev="svg",
  dpi=300,
  message=FALSE,
  warning=FALSE,
  cache=TRUE,
  cache.path = "cache/",
  cache.comments=FALSE
)

library(tidyverse)
library(pegas)
library(here)
library(Biostrings)
library(msa)
library(fs)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggnewscale)
library(pander)
library(paletteer)
library(patchwork)
library(ggfun)
library(phytools)
library(ggtree)
library(treeio)
source('spp.R')
```

```{r init, include=FALSE}
read_fasta <- function(x,aligned=FALSE) read.dna(x,"fasta",as.matrix=aligned)
write_fasta <- function(x,file) write.dna(x,file,"fasta",nbcol = -1,colsep = "")

# do an english comma-separated list
format_list <- function(l,fmt=function(x) x) {
  l <- fmt(l)
  if(length(l) > 1) l[length(l)] <- str_c("and ",l[length(l)])
  if (length(l) > 2) {
    str_c(l,collapse=", ")
  } else {
    str_c(l,collapse=" ")
  }
}

spp <- Taxa$new('Plagiotremus')


all_samples <- read_csv(here("data","samples.csv"))
original_samples <- all_samples
markers <- c("coi","cytb")

# load marine ecoregions
# TODO: figure out whether shifting longitudes here breaks anything in the map plot
regions <- st_read(here("data","gis","ecoregions","meow_ecos.shp"),quiet = TRUE) %>%
  st_shift_longitude() %>%
  group_by(ECOREGION) %>%
  mutate(middle = max(st_coordinates(st_shift_longitude(st_combine(geometry)))[,'X']) )  %>%
  ungroup()

scounts <- original_samples %>%
  separate_wider_delim(cols = "species",delim = " ",names=c("genus","species")) %>%
  mutate(id = str_glue("{str_sub(genus,1,1)}{str_sub(species,1,2)}")) %>%
  count(id) %>%
  deframe()


# stolen from insect: join two DNAbins together
join_seq <- function(...){
  dots <- list(...)
  nlsts <- length(dots)
  DNA <- any(sapply(dots, class) == "DNAbin")
  if(nlsts == 0) return(NULL)
  if(nlsts == 1) return(dots[[1]])
  islist <- sapply(dots, is.list)
  for(i in which(!islist)){
    tmpattr <- attributes(dots[[i]])
    attributes(dots[[i]]) <- NULL
    dots[[i]] <- list(dots[[i]])
    attributes(dots[[i]]) <- tmpattr
  }
  dots <- lapply(dots, unclass)
  findattr <- function(x){
    names(attributes(x))[sapply(attributes(x), length) == length(x)]
  }
  attrlist <- lapply(dots, findattr)
  ual <- unique(unlist(attrlist, use.names = FALSE))
  validattrs <- ual[sapply(ual, function(e) all(sapply(attrlist, function(g) e %in% g)))]
  validattrs <- validattrs[!validattrs %in% c("names", "class")]
  res <- unlist(dots, recursive = FALSE, use.names = TRUE)
  for(i in validattrs){
    attr(res, i) <- unlist(lapply(dots, attr, i), use.names = FALSE)
  }
  class(res) <- "DNAbin"
  return(res)
}

# format percents and numbers
pc <- function(x,a=0.01) scales::percent_format(accuracy = a)(x)
num <- function(x,a=0.01) scales::label_comma(accuracy=a)(x)

# format a markdown grid table from a data frame or file
print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf, roundr=FALSE, na="") {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

  if (inherits(tbl,"character")) {
    if (id == "") {
      id <- sub(pattern = "(.*?)\\..*$", replacement = "\\1", basename(file))
    }
    tbl <- read.csv(tbl,check.names=FALSE,stringsAsFactors=FALSE)
  } else if (inherits(tbl,"data.frame")) {
    if (id == "") stop("please specify table id")
  } else {
    stop("'tbl' must be a character or a data frame")
  }

  fmt <- function(x,r,a) {
    if (is.numeric(x) & is.numeric(r) & !isFALSE(r))
      round(x,r)
    else if (is.numeric(x))
      num(x,a)
    else
      x
  }

  tbl <- tbl %>%
    select(-any_of(hide)) %>%
    mutate(
      across(
        where(~format_numbers & is.numeric(.x) & !is.integer(.x)),
        ~fmt(.x,roundr,accuracy)
      )
    )


  # can also use rename_with(str_to_sentence)
  names(tbl) <- switch(
    str_to_lower(hdr_case),
    sentence = str_to_sentence(names(tbl)),
    title = str_to_title(names(tbl)),
    upper = str_to_upper(names(tbl)),
    lower = str_to_lower(names(tbl)),
    names(tbl)
  )

  caption <- sprintf("%s {#tbl:%s}",caption,id)

  table_str <- pandoc.table.return(
    tbl,
    keep.line.breaks=TRUE,
    caption=NULL,
    style="multiline",
    split.cells=cell_width,
    split.tables=Inf,
    justify='left',
    missing=na
  )
  table_str <- table_str[table_str != ""]
  cat(paste0(table_str,collapse="\n"),"\n\n",caption,"\n\n",sep="")
}

# little list of our sister species pairs, used for filtering datasets
species_pairs <- list(pgt=c('Plagiotremus goslinei','Plagiotremus tapeinosoma'),per=c('Plagiotremus ewaensis','Plagiotremus rhinorhynchos'))
hawaii_species <- list(pgo='Plagiotremus goslinei',pew='Plagiotremus ewaensis')

# map of various ecoregion realms to color palettes
palette_map <- c(
  'Red Sea' = 'RColorBrewer::Reds',
  'Western Indian Ocean' = 'RColorBrewer::Oranges',
  'Central Indo-Pacific' = 'beyonce::X59',
  'Central Pacific' = 'RColorBrewer::Greens',
  'Hawaii' = 'ggsci::purple_material',
  'Marquesas' = 'Redmonder::dPBIYlBu',
  'Polynesian Province' = 'RColorBrewer::Blues'
)

# make plot tags
plot_tags <- list(str_glue('({LETTERS})'))
```

```{r bold-seqs, dependson="init", include=FALSE}
library(bold)
# remove: Pbr-GBMTG324-16, Pbr-GBMNA15372-19
# remove: Pbr-GBMTG324-16
# bold_seqs_raw <- bold_seqspec(taxon=c("Plagiotremus","Xiphasia","Meiacanthus","Petroscirtes","Aspidontus"))
bold_seqs_raw <- bold_seqspec(taxon=c("Plagiotremus","Xiphasia","Meiacanthus","Omobranchus anolius"))

to_filter <- c('Pbr-GBMTG324-16')

bold_tbl <- bold_seqs_raw %>%
  as_tibble() %>%
  filter(
    nucleotides != "",
    species_name != "",
    !str_detect(species_name,"(sp|cf)\\."),
    !str_detect(processid,'FUT225') # this is a typo, it's a damselfish with the same specific epithet
  ) %>%
  separate_wider_delim(species_name,delim=" ",names=c(NA,"spp"),cols_remove = FALSE) %>%
  mutate(
    id = str_glue("{str_sub(genus_name,1,1)}{str_sub(spp,1,2)}-{processid}"),
    nucleotides = str_replace(nucleotides,"^-+",""),
    nucleotides = str_replace(nucleotides,"-+$",""),
    nucleotides = str_replace(nucleotides,"N+$",""),
    nucleotides = str_replace(nucleotides,"^N+",""),
    source="bold",
    sequenced_coi=1,
    sequenced_cytb=0
  ) %>%
  distinct(id,.keep_all = TRUE) %>%
  filter(str_detect(markercode,'COI')) %>%
  filter(!(id %in% to_filter)) %>%
  select(id,species=species_name,country,locality=exactsite,sequenced_coi,sequenced_cytb,lon,lat,source,nucleotides,bin_uri,processid,genbank_accession,catalognum)

bold_dna <- map2(bold_tbl$id,bold_tbl$nucleotides,~{
  c(str_glue(">{.x}"),.y)
}) %>%
  unlist() %>%
  textConnection() %>%
  read_fasta(aligned=FALSE)

bold_seqs <- list(
  coi = list(
    samples = bold_tbl %>% select(-nucleotides),
    sequences = bold_dna
  )
)

all_samples <- all_samples %>%
  bind_rows(bold_tbl %>% select(-nucleotides)) %>%
  separate_wider_delim(species,delim=' ',names=c('genus','species'),cols_remove = FALSE)
```

```{r local-seqs, dependson="init", include=FALSE}
local_seqs <- markers %>%
  set_names() %>%
  purrr::map( \(marker) {
    seqs <- read_fasta(here("data",str_glue("plagiotremus-{marker}.fasta")),aligned=FALSE)

    # filter metadata to match seqs
    samples <- all_samples %>%
      dplyr::filter(id %in% labels(seqs)) %>%
      arrange(id)

    list(samples = samples, sequences = seqs)
  })
```

```{r alignment, dependson=c("bold-seqs","local-seqs"), include=FALSE}
alignments <- markers %>%
  set_names() %>%
  purrr::map(\(marker) {
    ss <- local_seqs %>%
      pluck(marker,'sequences')
    if (exists(marker,where=bold_seqs)) {
      s2 <- bold_seqs %>%
        pluck(marker,'sequences')
      ss <- join_seq(ss,s2)
    }

    al <- ss %>%
      as.character() %>%
      lapply(paste0,collapse="") %>%
      unlist() %>%
      DNAStringSet() %>%
      msa(.,method="ClustalOmega",gapOpening = 2000,gapExtension = 1000) %>%
      as.DNAbin()

    write_fasta(al,here("cache",str_glue("{marker}_aligned.fasta")))

    covg <- apply(al,2,\(x) (sum(x != 4) / length(x)))
    list(c50 = 0.5, c99 = 0.99, c00 = 0) %>%
      imap(\(cov,idx) {
        p1 <- min(which(covg >= cov))
        p2 <- max(which(covg >= cov))
        ll <- al[,p1:p2]
        write_fasta(ll,here('output','alignments',str_glue('{marker}_{idx}.fasta')))
        return(ll)
      })
  })

sample_data <- markers %>%
  set_names() %>%
  purrr::map(\(marker) {
    all_samples %>%
      filter(id %in% labels(alignments %>% pluck(marker,'c00')))
  })

# cytb_samples <- all_samples %>% filter(id %in% labels(alignments$cytb$c00))
cycount <- sample_data$cytb %>%
  dplyr::count(source) %>%
  deframe()

cocount <- sample_data$coi%>%
  dplyr::count(source) %>%
  deframe()

coi_len <- purrr::map(alignments$coi,ncol)
cytb_len <- purrr::map(alignments$cytb,ncol)
```

```{r haplotypes, include=FALSE, dependson="alignment"}
# map MEOW provinces to kulbicki provinces
province_map <- c(
  'Agulhas' = 'Western Indian Ocean',
  'Red Sea and Gulf of Aden' = 'Red Sea',
  'Western Indian Ocean' = 'Western Indian Ocean',
  'Central Indian Ocean Islands' = 'Western Indian Ocean',
  'South Kuroshio' = 'Central Indo-Pacific',
  'Tropical Northwestern Pacific' = 'Central Indo-Pacific',
  'Northeast Australian Shelf' =  'Central Indo-Pacific',
  'Tropical Southwestern Pacific' = 'Central Pacific',
  'Marshall, Gilbert and Ellis Islands' = 'Central Pacific',
  'Central Polynesia' = 'Central Pacific',
  'Southeast Polynesia' = 'Polynesian Province',
  'Hawaii' = 'Hawaii',
  'Marquesas' = 'Marquesas',
  'Java Transitional' = 'Central Indo-Pacific',
  'Western Coral Triangle' = 'Central Indo-Pacific'
)

kprovince_map <- c(
  'Red Sea and Gulf of Aden' = 'Red Sea',
  'Agulhas' = 'Western Indian Ocean',
  'Western Indian Ocean' = 'Western Indian Ocean',
  'Central Indian Ocean Islands' = 'Western Indian Ocean',
  'South Kuroshio' = 'Central Indo-Pacific',
  'Tropical Northwestern Pacific' = 'Central Indo-Pacific',
  'Northeast Australian Shelf' =  'Central Indo-Pacific',
  'Java Transitional' = 'Central Indo-Pacific',
  'Western Coral Triangle' = 'Central Indo-Pacific',
  'Tropical Southwestern Pacific' = 'Central Pacific',
  'Marshall, Gilbert and Ellis Islands' = 'Central Pacific',
  'Central Polynesia' = 'Central Pacific',
  'Southeast Polynesia' = 'Polynesian Province',
  'Marquesas' = 'Polynesian Province',
  'Hawaii' = 'Hawaii'
)

province_abbrev <- c(
  'Western Indian Ocean' = 'WIO',
  'Western Indian' = 'WIO',
  'Red Sea' = 'RS',
  'Central Indo-Pacific' = 'CIP',
  'Central Pacific' = 'CP',
  'Polynesian' = 'PL',
  'Polynesian Province' = 'PL',
  'Main Hawaiian Islands' = 'MHI',
  'Northwestern Hawaiian Islands' = 'NWHI',
  'Hawaii' = 'HW',
  'Marquesan' = 'MQ',
  'Marquesas' = 'MQ'
)

wrap_fac <- function(f,w=20) {
  factor(
    str_wrap(f,w),
    levels = str_wrap(levels(f),w)
  )
}

plagiotremus_data <- all_samples %>%
  filter(species %in% unlist(species_pairs)) %>%
  drop_na(lon,lat) %>%
  st_as_sf(coords=c('lon','lat'),remove=FALSE,crs=4326) %>%
  st_join(regions,st_within) %>%
  mutate(
    llon = if_else(middle < 0,middle+360,middle),
    ecoregion = case_when(
      ECOREGION == "Hawaii" & nwhi ~ "Northwestern Hawaiian Islands",
      ECOREGION == "Hawaii" & !nwhi ~ "Main Hawaiian Islands",
      str_detect(ECOREGION,regex('red sea',ignore_case = TRUE)) ~ 'Red Sea',
      .default = ECOREGION
    ),
    province = case_when(
      ECOREGION == "Hawaii" & nwhi ~ "Northwestern Hawaiian Islands",
      ECOREGION == "Hawaii" & !nwhi ~ "Main Hawaiian Islands",
      .default = province_map[PROVINCE]
    ),
    prov = province_abbrev[province],
    kulbicki_province = kprovince_map[PROVINCE],
    kprov = province_abbrev[kulbicki_province],
    old_color_category = fct_reorder(
      case_when(
        str_detect(ecoregion,'Hawaii') ~ 'Hawaii',
        str_detect(ecoregion,'Marquesas') ~ 'Marquesas',
        .default = REALM
      ),
      llon
    ),
    color_category = fct_reorder(kulbicki_province,llon)
  )  %>%
  arrange(old_color_category,llon) %>%
  mutate(
    n = row_number(),
    ecoregion = fct_reorder(ecoregion,n),
    province = fct_reorder(province,n),
    kulbicki_province = fct_reorder(kulbicki_province,n),
    prov = fct_reorder(prov,n),
    ecoregion = fct_relevel(ecoregion,c('Main Hawaiian Islands','Northwestern Hawaiian Islands'),after=Inf),
    kprov = fct_relevel(fct_reorder(kprov,n),"HW",after=Inf)
  ) %>%
  as_tibble()

master_palette <- plagiotremus_data %>%
  as_tibble() %>%
  count(color_category,ecoregion) %>%
  group_by(color_category) %>%
  mutate( color = paletteer_d( palette_map[as.character(cur_group()$color_category)],direction=-1)[1:n()], n = n() ) %>%
  ungroup()

haplotypes <- species_pairs %>%
  purrr::map(\(pair) {
    markers %>%
      set_names() %>%
      purrr::map(\(marker) {
        alignment <- alignments %>%
          pluck(marker,'c00')
        # filter by samples in this species pair
        samples <- plagiotremus_data %>%
          filter(
            species %in% pair,
            id %in% labels(alignment)
          )
        alignment <- alignment[samples$id,]
        # calculate per-base coverage
        covg <- apply(alignment,2,\(bases) sum(bases != 4) / length(bases))
        # keep only positions with >= 99% coverage
        alignment <- alignment[,covg >= 0.99]
        samples <- samples %>%
          filter(id %in% labels(alignment))
        hplot <- NULL
        if (nrow(alignment) > 0) {
          haplotypes <- pegas::haplotype(alignment)
          hnet <- haploNet(haplotypes)
          hindex <- attr(haplotypes,'index')
          ht_locations <- haploFreq(alignment,fac=samples$ecoregion)
          palette <- master_palette %>%
            filter(ecoregion %in% colnames(ht_locations)) %>%
            dplyr::slice(match(colnames(ht_locations),ecoregion)) %>%
            select(ecoregion,color) %>%
            deframe() %>%
            set_names(str_wrap(names(.),25))
          bar_legend <- ggplot(data = samples) +
            geom_bar(aes(y = ecoregion, fill = wrap_fac(ecoregion,25))) +
            scale_fill_manual(values = palette,name='Ecoregion') +
            scale_y_discrete(limits=rev) +
            theme_transparent() +
            theme(
              plot.title = element_blank(),
              plot.subtitle = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks.y = element_blank(),
              axis.title = element_blank(),
              panel.grid.major.y = element_blank(),
              panel.grid.minor = element_blank(),
              panel.border = element_blank(),
              panel.grid = element_blank(),
              strip.background = element_rect(fill="grey95",color="darkgrey"),
              legend.title = element_blank()
            ) +
            labs(title = "Legend", subtitle = "Number of samples")  +
            facet_wrap(
              ~kprov,
              scales="free_y",
              ncol=1,
              space="free_y",
              strip.position = "left"
            )

          leg <- get_legend(bar_legend)
          leg <- ggplot() +
            theme_void() +
            annotation_custom(leg) +
            theme(margins = margin(0,0,0,0))
          bar_legend <- bar_legend + theme(legend.position = "none")
          pd <- NULL
          pp <- str_replace_all(pair,'Plagiotremus ','')
          rds_file <- here('data','haplotypes',str_glue('{str_c(pp,collapse="-")}_{marker}_replot.rds'))
          if (file_exists(rds_file)) pd <- readRDS(rds_file)
          # plot(hnet, pie = ht_locations, bg = palette, # keep the same
          #      size=attr(hnet, "freq"), show.mutation = 1, length = 0.5, space = 0.75, # mess with
          #      # labels = F, threshold = 0, fast = F, legend = F)#, xy=pd)
          #      labels = F, threshold = 0, fast = F, legend = F, xy=pd)
          # pd <- replot()
          # saveRDS(pd,rds_file)
          hap_layout <- c(
            # area(top, left, bottom, right)
            area(1,5,100,100),  # plot
            area(64,1.5,99,30), # bar plot
            area(1,1.5,64,15)   # legend
          )

          # get haplotype network plot from base R
          net_plot <- wrap_elements(
            panel = ~plot(
              hnet, pie = ht_locations, bg = palette,
              size=attr(hnet, "freq"), show.mutation = 1,
              labels = F, threshold = 0, fast = F, legend = F, xy=pd
            ),
            clip = FALSE
          )

          joint_plot <- net_plot +
            bar_legend +
            leg +
            plot_layout(design=hap_layout)

          list(
            alignment = alignment,
            samples = samples,
            plot = joint_plot,
            haplotypes = haplotypes,
            network = hnet,
            frequencies = ht_locations
          )
        } else {
          NULL
        }
      })
  })

hap_summary <- function(h) {
  hindex <- attr(h$haplotypes,'index') %>%
    imap(\(idx,hap) {
      tibble(
        haplotype = hap,
        sequence = rownames(h$alignment)[idx]
      )
    }) %>%
    list_rbind() %>%
    left_join(h$samples,by=c(sequence='id'))

  h_summary = hindex %>%
    group_by(haplotype) %>%
    summarise(n = n(), spp = n_distinct(species), src = n_distinct(source), loc = n_distinct(province)) %>%
    ungroup()

  list(
    n = nrow(h$haplotypes),
    details = hindex,
    summary = h_summary,
    shared = h_summary %>%
      filter(n > 1) %>%
      nrow(),
    shared_spp = h_summary %>%
      filter(spp > 1) %>%
      nrow(),
    shared_src = h_summary %>%
      filter(src > 1) %>%
      nrow()
  )
}
```

```{r amovas, include=FALSE,dependson="haplotypes"}
# this goes slow as hell so we're manually caching it for
# convenience sake in rstudio
if (!file_exists(here('cache','amovas.rds'))) {
  amovas <- haplotypes %>%
    imap(\(pair,spp) {
      pair %>%
        imap(\(hap,marker) {
          if (!is.null(hap)) {
            samples <- hap$samples %>%
              add_count(province) %>%
              filter(n >= 5)
            dd <- dist.dna(hap$alignment[samples$id,])
            list(
              amova = amova(dd ~ prov,data=samples,nperm = 9999),
              pairwise = haplotypes::pairPhiST(dd,samples$prov,nperm=9999)
            )
          }
        })
    })
  saveRDS(amovas,here('cache','amovas.rds'))
}
amovas <- readRDS(here('cache','amovas.rds'))
```

```{r dna-distances, include=FALSE, dependson="haplotypes"}
dna_distances <-
  haplotypes %>%
  imap(\(pair,spp) {
    pair %>%
      imap(\(hap,marker) {
        if (!is.null(hap)) {
          alignment <- hap$alignment
          samples <- hap$samples
          distances <- dist.dna(alignment) %>%
            as.matrix() %>%
            as_tibble(rownames='left') %>%
            pivot_longer(-left,names_to = 'right',values_to = 'dist') %>%
            left_join(samples %>% select(id,prov),by=c(left='id')) %>%
            left_join(samples %>% select(id,prov),by=c(right='id'),suffix=c('_left','_right')) %>%
            filter(left != right) %>%
            group_by(across(starts_with('prov'))) %>%
            summarise(dist = mean(dist,na.rm=TRUE)) %>%
            ungroup()  %>%
            mutate(dist = if_else(prov_left == prov_right,0,dist)) %>%
            pivot_wider(names_from = 'prov_right', values_from = 'dist')
          dist_matrix <- distances %>%
            column_to_rownames('prov_left') %>%
            as.matrix()
          diag(dist_matrix) <- NA
          dist_matrix[upper.tri(dist_matrix)] <- NA
          dist_matrix %>%
            as.data.frame() %>%
            filter(!if_all(everything(),is.na)) %>%
            select(where(~!all(is.na(.x))))
        }
      })
  })
```

# Abstract {-}

The Hawaiian archipelago is an endemism hotspot. Up to twenty-five percent of Hawaiian reef fish species are endemic, a higher proportion than anywhere else in the world. However, unlike terrestrial species where high endemism often arises through adaptive radiation (e.g. honeycreepers, fruit flies, silverswords, etc.)—marine endemism in Hawai&#x02BB;i is more evenly- distributed across taxa. Among reef fishes in Hawai&#x02BB;i, sister taxa rarely co-occur, and when they do, they exhibit a higher degree of range size asymmetry than co-occurring sister species elsewhere in the Indo-Pacific. In other words, the sister species of (small range size) Hawaiian endemics are typically widespread in the Indo-Pacific (large range size).

# Introduction {-}

The Hawaiian archipelago is an endemism hotspot, with up to twenty-five percent of its reef fish found nowehere else on earth [@randallReefShoreFishes2007]. However, whereas terrestrial endemism in Hawai&#x02BB;i is often associated with adaptive radiation [e.g., honeycreepers: @freedEvolutionaryEcologyRadiation1987a; fruit flies: @carsonChromosomalSequencesInterisland1983; silverswords: @witterAdaptiveRadiationGenetic1988, etc.], marine endemism is more evenly-distributed across taxa [@kayEndemismEvolutionHawaiian1987; @mundyChecklistFishesHawaiian2005]. Among Hawaiian reef fishes, sister taxa rarely co-occur and when they do, they often exhibit a higher degree of range size asymmetry than co-occurring sister species elsewhere in the Indo-Pacific [@gaboriauInfluenceGeographySpeciation2018]. Whether or not they co-occur, sister species of Hawaiian endemic fishes are often widespread in the Indo-Pacific [@gaboriauInfluenceGeographySpeciation2018; @randallZoogeographyShoreFishes1998; @hobanCirripectesMatatakaroNew2020].

Two examples of endemic reef fishes with widespread sister species are currently recognized within the sabertooth blenny genus `r spp$cite("Plagiotremus")` (Blennioidei: Blenniidae). `r spp$cite("Plagiotremus goslinei")` and `r spp$cite("Plagiotremus ewaensis",TRUE)` are endemic to the Hawaiian Islands and their respective sister taxa, `r spp$cite("Plagiotremus tapeinosoma",TRUE)` and `r spp$cite("Plagiotremus rhinorhynchos",TRUE)` are distributed across the tropical Indo-West-Pacific from the Red Sea to the Marquesas Islands [@smith-vanizSabertoothedBlenniesTribe1976; @randallReefShoreFishes2007]. A phylogenetic analysis of *Plagiotremus* using 22 parsimony-informative morphological characters suggested four species groups and three subgenera [@smith-vanizSabertoothedBlenniesTribe1976]. According to the analysis, the subgenus *Plagiotremus* contains only `r spp$cite("Plagiotremus spilistius")` and `r spp$cite("Plagiotremus iosodon",TRUE)`. `r spp$cite("Plagiotremus celebesensis")` was originally included, but was synonymized with *P. spilistius* in a follow-up study [@smith-vanizSaberToothedBlenniesTribe1987a]. The subgenus *Musgravius*, recognized as the sister to subgenus *Plagiotremus*, contains the *laudandus* group comprising `r spp$cite("Plagiotremus laudandus")`, `r spp$cite("Plagiotremus phenax",TRUE)`, and `r spp$cite("Plagiotremus townsendi",TRUE)` (although the subspecies *Plagiotremus laudandus flavus* may additionally warrant elevation to species) [@hastingsRecognizingDiversityBlennioid2009a; @smith-vanizAggressiveMimicryAspidontus2020]. *Runula*, which is sister to the other subgenera, contains the *rhinorhynchos* group (*P. rhinorhynchos* and *P. ewaensis*) and the *tapeinosoma* group (*P. tapeinosoma*, *P. goslinei*, and `r spp$cite("Plagiotremus azaleus",TRUE)`) [@smith-vanizSabertoothedBlenniesTribe1976]. Although Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976; -@smith-vanizSaberToothedBlenniesTribe1987a] supported the Hawaiian endemic/widespread sister species relationships, he did not rule out the possibility that putative endemics comprised isolated populations of their widespread sister species. Furthermore, he could not unambiguously resolve the interrelationships of subgenera within *Plagiotremus*, nor the placement of *Plagiotremus* relative to other genera in the tribe Nemophini. Although morphology-based phylogenetic analyses have provided some taxonomic clarity within this group, molecular approaches may enable the further resolution of uncertain relationships as well as help place endemic species within a broader biogeographic context.

Few studies have included *Plagiotremus* in molecular-phylogenetic analyses, and none have considered species endemic to Hawai&#x02BB;i. *Plagiotremus rhinorhynchos*, *P. tapeinosoma*, *P. azaleus*, and *P. laudandus* are found in several multilocus analyses of blennioids [@linPhylogenyBiogeographyShallow2013; @casewellEvolutionFangsVenom2017; @hundtMultilocusMolecularPhylogeny2014; @hundtExtremeDentitionDoes2018]. Most recently, Smith-Vaniz et al. [-@smith-vanizAggressiveMimicryAspidontus2020] compared cytochrome c oxidase subunit I (COI) sequences from five species of *Plagiotremus* (*P. tapeinosoma*, *P. laudandus*, *P. flavus*, *P. townsendi*, and *P. phenax*) using a neighbor-joining tree, calling for the reintegration of *P. flavus* as a subspecies under *P. laudandus*.

Another important question relating to endemic reef fishes in Hawai&#x02BB;i is their geographic origin. Due to the volcanic nature of the Hawaiian archipelago, all species necessarily originated elsewhere in the Pacific. For reef fishes, there are two prevailing hypotheses regarding their colonization of the Hawaiian Islands: i) Hawaiian species originate in the western Pacific, ariving through rare dispersal events via the Kuroshio current [@houriganMidoceanIsolationEvolution1987a]; ii) Hawaiian species originate in the central/southern Pacific, arriving via stepping-stone dispersal through the Line Islands [@goslineInshoreFishFauna1955]. In both cases, Johnston Atoll occupies a potentially key position as the nearest reef habitat to the Hawaiian Archipelago, although its specific role in all cases remains unresolved [@gaitherHighConnectivityDeepwater2011; @dibattistaPhylogeographyPacificBlueline2011; @lobelJohnstonAtollReef2020; @andrewsPhylogeographicAnalysesSubmesophotic2014; @riveraSubtlePopulationGenetic2004] and while many endemic Hawaiian reef fishes are known to occur at there, no records yet exist for either Hawaiian *Plagiotremus* species [@randallAnnotatedChecklistFishes1985]. Phylogeographic studies have shown support for both hypotheses, with species such as limpets (*Cellana* spp.) [@birdDiversificationSympatricBroadcastspawning2011] and butterflyfishes (*Chaetodon miliaris*, *C. fremblii*) [@craigOriginsAgesPopulation2010] arriving via the western route and combtooth blennies (*Cirripectes vanderbilti*) [@hobanCirripectesMatatakaroNew2020] and *Etelis* spp. snappers [@andrewsPhylogeographicAnalysesSubmesophotic2014] colonizing via the southern route.

To resolve the taxonomy of *Plagiotremus* at various levels, we combined sequence data from hundreds of new *Plagiotremus* samples with sequences from a public database to create a molecular phylogeny based on two mitochondrial markers. With data from nine putative *Plagiotremus* species and three other genera, this study presents the most taxonomically-comprehensive molecular phylogeny of *Plagiotremus* to-date. This allowed for comparisons with the morphology-based phylogeny at multiple levels. Furthermore, and of particular interest in Hawai&#x02BB;i, this study is the first to present molecular sequence data from the endemic Hawaiian *Plagiotremus* species. If both *P. ewaensis* and *P. goslinei* are truly distinct from their widespread sister species *P. rhinorhynchos* and *P. tapeinosoma*, each should be considerably diverged from one another at the molecular level.

# Materials and Methods {-}

## Specimen collection {-}

We collected `r nrow(original_samples)` individuals of *Plagiotremus tapeinosoma*, *P. rhinorhynchos*, *P. goslinei*, and *P. ewaensis* from Hawai&#x02BB;i and abroad. Specimens were collected by pole spear (captured among the spread-out tines of a three-prong) and/or hand net and a portion of the caudal fin was preserved in salt-saturated DMSO solution [@seutinPreservationAvianBlood2011] for subsequent molecular analysis. *Plagiotremus rhinorhynchos* and *P. tapeinosoma* were collected across their ranges in the Indo-Pacific, from the Red sea to the Marquesas Islands. *Plagiotremus ewaensis* and *P. goslinei* were collected from 13 locations in the Main (MHI) and Northwestern Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_collection]). All specimens from the NWHI were collected under permits PMNM-2007-032, PMNM-2008-046, PMNM-2009-032L, PMNM-2009-044, PMNM-2011-025, and PMNM-2012-045, issued by the National Oceanic and Atmospheric Administration, the U.S. Fish and Wildlife Service, and the State of Hawaii Division of Aquatic Resources to BWB. Fish sampling protocols were approved by the University of Hawai&#x02BB;i at Manoa Institutional Animal Care & Use Committee under approval nos. 09-753 and 15-2271.

## DNA extraction and sequencing {-}

Tissue samples were brought to the Hawai&#x02BB;i Institute of Marine Biology (HIMB) for processing. We extracted genomic DNA using a modified version of the HotSHOT protocol [@meekerMethodIsolationPCRready2007] or the Omega EZNA-96^&reg;^ Tissue DNA Kit (Omega Bio-Tek, Georgia, USA) following manufacturer's protocols. Specimens were amplified for a region of the mitochondrial cytochrome oxidase subunit I (COI) gene using the primers FishF1 (5'--TCAACCAACCACAAAGACATTGGCAC--3') and FishR1 (5'--TAGACTTCTGGGTGGCCAAAGAATCA--3') following the chemistry and thermocycler settings of Ward et al. [-@wardDNABarcodingAustralia2005]. Select specimens of *Plagiotremus goslinei* and *P. tapeinosoma* were also amplified for a region of cytochrome B (cytb) using the custom-designed forward primer BlnF3 (5'--TCAGTHTGATGAAACTTTGGCTC--3') and the reverse primer cyb7 (5'--AATAGGAAGTATCATTCGGGTTTGAT--3') [@taberletUnusualMitochondrialDNA1992]. PCR reactions for cytb were carried out in 15μl volumes consisting of 7.5μl BioMix Red™ (Bioline Inc., Massachusetts, USA), 0.3μl of each 10μM primer, 1.5μl BSA (New England Biolabs, Massachusetts, USA), 1μl DNA, and filled to volume with PCR-grade ultrapure water. Reactions were performed under the following conditions: 95°C for 2 m followed by 35 cycles of denaturation at 94°C for 30 s, annealing for 50°C for 45 s, and extension at 72°C for 45 s, with a final extension of 10 min at 72°C. We visualized PCR products for all markers using 1.5-2% agarose gels stained with GelRed^&reg;^ Nucleic Acid Gel Stain (Biotium, California, USA) and purified final products by incubating them with 0.75 units of exonuclease and 0.5 units of shrimp alkaline phosphatase (ExoSAP; USB, Ohio, USA) per PCR product for 30 min at 37°C followed by 15 min at 85°C. DNA sequencing was performed using fluorescently-labeled dideoxy terminators on an ABI 3730XL Genetic Analyzer (Applied Biosystems, California, USA) at the University of Hawai&#x02BB;i Advanced Studies of Genomics, Proteomics and Bioinformatics sequencing facility.

## Additional sequence acquisition {-}

We downloaded all available COI sequence data in the Barcode of Life Database (BoLD) for the genera *Plagiotremus*, *Xiphasia* and *Meiacanthus*. *Omobranchus* has been identified as an appropriate outgroup to the tribe Nemophini in the multi-locus molecular phylogenies discussed above, so a COI sequence from *Omobranchus anolius* with identification and location information was downloaded from BoLD and used as the outgroup in our phylogenetic analyses. <!--Three *Plagiotremus* sequences pulled from BoLD were excluded in downstream analyses due to unclear origins or mislabeling.-->


## Sequence analysis {-}

### Phylogenetic analyses {-}

All data analyses were performed using `R` 4.4.1 [@rcoreteamLanguageEnvironmentStatistical2024]. Sequence data was read using the `R` packages `Biostrings` [@aboyounBiostringsEfficientManipulation2017] and `ape` [@paradisApeEnvironmentModern2019] and aligned using the Clustal Omega algorithm [@sieversFastScalableGeneration2011] implemented in the package `msa` [@bodenhoferMsaPackageMultiple2015]. Since we used sequences from various sources, their lengths varied and not all loci were present in the final alignment. To compare the effects of missing data on tree node support, three versions of the sequence alignment were generated: one that excluded sites missing in more than one individual (hereafter referred to as the 99% coverage dataset), one that only included sites sequenced in >50% of individuals (>50% coverage dataset), and one with all sequences in the alignment (the full dataset). After filtering, identical sequences were merged and taxonomic and geographic distribution of merged sequences were recorded.

We conducted a Bayesian phylogenetic reconstruction using `MrBayes` v3.2.7 [@ronquistMrBayes32Efficient2012], running three independent searches for 20 million generations each, saving trees every 1,000 generations and discarding the first 10% as burn-in. We used `PartitionFinder` 2 with greedy algorithm to select the best model(s) of nucleotide substitution partitioned by codon position, based on corrected Akaike Information Criterion (AICc) [@lanfearPartitionFinder2New2017; @lanfearPartitionFinderCombinedSelection2012; @guindonNewAlgorithmsMethods2010]. We verified model parameter and Markov-chain convergence using `Tracer` v1.7.2 [@rambautPosteriorSummarizationBayesian2018], assuring effective sample sizes &gt;200 for all parameters. Consensus trees were interpreted and plotted in R using the `ggtree` package [@yuGgtreePackageVisualization2017]. As a compromise between potential under- and over-filtering, we used the &gt;50% coverage dataset to examine phylogenetic relationships within *Plagiotremus*.

### Intraspecific phylogeography {-}

To examine intraspecific relationships within *P. tapeinosoma*, *P. goslinei*, *P. rhinorhynchos*, and *P. ewaensis*, we constructed haplotype networks using all sequences from each species pair. Sequences were interpreted and aligned using the same procedures as for phylogenetic analyses and only loci sequenced in all samples were used (i.e., no sites with missing data were kept). Haplotype frequencies were calculated, arranged into a maximum-parsimony network [@templetonCladisticAnalysisPhenotypic1992], and plotted using the package `pegas` [@paradisPegasPackagePopulation2010]. Sampling locations were grouped by color into marine biogeographic provinces as defined for reef fishes by Kulbicki et al. [-@kulbickiGlobalBiogeographyReef2013], with the Hawaiian province subdivided into Main and Northwestern Hawaiian Islands and the Marquesas separated from the Polynesian province. We examined genetic differentiation and diversity among provinces using the K2P distance [@kimuraSimpleMethodEstimating1980] and analysis of molecular variance (AMOVA), as well as calculating haplotype diversity (h), nucleotide diversity ($\mathrm{\uppi}$), and Tajima's D (as implemented in `pegas`). In addition to overall AMOVA analyses, we calculated pairwise $\upPhi_\mathrm{st}$ using the package `haplotypes` [@aktasHaplotypesManipulatingDNA2025] to examine degree of differentiation among provinces. Statistical significance was calculated using permutation tests (with 9,999 permutations) and *p*-values were adjusted for multiple comparisons according to the modified false discovery rate [B-Y FDR, @benjaminiControlFalseDiscovery2001; @narumBonferroniLessConservative2006].

# Results {-}

```{r alignment-summary, dependson="alignment", include=FALSE}
coi_covg <- apply(alignments$coi$c00,2,\(x) (sum(x != 4) / length(x)))
cytb_covg <- apply(alignments$cytb$c00,2,\(x) (sum(x != 4) / length(x)))
sample_seqs <- sample_data %>%
  purrr::map(\(x) x %>%
        count(source) %>%
        deframe() %>%
        as.list()
  )
```

## Overall sequencing results {-}

We generated `r sample_seqs$coi$sequenced` new COI sequences and retrieved `r sample_seqs$coi$bold` sequences from BOLD ([@tbl:tbl_collection;@tbl:tbl_supp_bold_seqs]). The final full COI alignment was `r coi_len$c00` base pairs long (50% coverage: `r coi_len$c50` bp, 99% coverage: `r coi_len$c99` bp) ([@fig:fig_supp_covg]). We generated `r sample_seqs$cytb$sequenced` new cytb sequences, retaining the longest alignment with sites present in all samples (`r cytb_len$c00` bp). New sequences generated for this study have been deposited in GenBank under accession numbers [XXXX](https://www.ncbi.nlm.nih.gov/nuccore/)--[XXXX](https://www.ncbi.nlm.nih.gov/nuccore/).

## Phylogenetic reconstruction {-}

<!--
models:
codon 1: GTR+I+G
codon 2: HKY+I
codon 3: GTR+G
-->

```{r include=FALSE, dependson="alignment"}
samples <- sample_data$coi
alignment <- alignments$coi$c50[samples$id,]

dd <- dist.dna(alignment) %>%
  as.matrix()

az_sp <- samples %>%
  filter(species == 'Plagiotremus azaleus') %>%
  pull(id)
gt_sp <- samples %>%
  filter(species %in% c('Plagiotremus tapeinosoma','Plagiotremus goslinei')) %>%
  pull(id)

paz_div <- mean(dd[az_sp,gt_sp])
```

According to `PartitionFinder` 2, the best-fit partitioning scheme for the 50%-coverage COI alignment was GTR+I+$\upGamma$, HKY+I, and GTR+$\upGamma$ for codons 1--3 respectively. Best-fit models for other filtering levels can be found in [@tbl:tbl_supp_models]. All genus-level nodes (*Xiphasia*, *Meiacanthus*, and *Plagiotremus*) in the Bayesian consensus tree had $\geq99%$ posterior probability based on the 50% coverage and full datasets ([@fig:fig_tree_genera[]{}A; @fig:fig_tree_genera[]{}B]). In the 99% coverage dataset, the split between *Xiphasia* and *Meiacanthus/Plagiotremus* had posterior support of ~90% ([@fig:fig_tree_genera[]{}C]). All three trees placed *Meiacanthus* as sister to *Plagiotremus*.

All *Plagiotremus* species currently recognized as valid (with the exception of the *tapeinosoma* group) resolved as monophyletic with high ($\geq99\%$) node support ([@fig:fig_tree_plagiotremus]). *Plagiotremus ewaensis* and *P. rhinorhynchos* comprised the sister group to all other species. *Plagiotremus townsendi* and *phenax* fell within distinct clades but *P. laudandus* and *flavus*, their sister group, formed an unresolved polytomy (*P. flavus* is recognized as a subspecies of *P. laudandus*). The *tapeinosoma* group, containing *P. azaleus*, *P. tapeinosoma*, and *P. goslinei*, was less clearly-resolved. *Plagiotremus tapeinosoma* was polyphyletic, with samples from the Marquesas forming a clade basal to the group and the remaining samples comprising a polytomy of mixed *P. goslinei* and *tapeinosoma* sequences. *Plagiotremus azaleus* occurred on a long branch within the polytomy, but made up a well-suported monophyletic group diverged by an average of `r scales::percent(paz_div,0.01)` (K2P distance).

## Phylogeography {-}

### *Plagiotremus goslinei* and *P. tapeinosoma* {-}

#### COI {-}
```{r include=FALSE, dependson="haplotypes"}
hap <- haplotypes$pgt$coi
h <- hap_summary(hap)
top_3 <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(haplotype)
top_3_n <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(n)
max_hap <- h$summary %>%
  arrange(desc(n)) %>%
  dplyr::slice(4:n()) %>%
  pull(n) %>%
  max()


hap1 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[1]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()

hap2 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[2]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()
```

There were `r h$n` unique COI haplotypes among the `r nrow(hap$alignment)` sequences generated for the *P. goslinei*/*tapeinosoma* species pair, with `r h$shared` haplotypes that were shared in at least two individuals, `r h$shared_spp` shared across species, and `r h$shared_src` shared between original and BoLD data sources ([@fig:fig_hap_coi_pgt]). There were `r length(seg.sites(hap$alignment))` segregating sites, overall haplotype diversity (h) was `r num(hap.div(hap$alignment),0.001)` and nucleotide diversity ($\mathrm{\uppi}$) was `r num(nuc.div(hap$alignment),0.001)`. Haplotype diversity was highest in the Western Indian Ocean (WIO) and Central Indo-Pacific (CIP) (WIO: h = 0.926; CIP: h = 0.952; [@tbl:tbl_stats_coi_pgt]) and lowest in the Marquesas (h = 0.474). Nucleotide diversity was also highest in the WIO ($\mathrm{\uppi}$ = 0.0056) and CIP ($\mathrm{\uppi}$ = 0.008), but was lowest in the Red Sea ($\mathrm{\uppi}$ = 0.001). Tajima's D was negative and significant in Hawai&#x02BB;i, the Marquesas, and the Polynesian province, which may indicate past population expansion or selection in these areas.

```{r include=FALSE, dependson="alignment"}
alignment <- haplotypes$pgt$coi$alignment
samples <- haplotypes$pgt$coi$samples

dd <- dist.dna(alignment) %>%
  as.matrix()

m_sp <- samples %>%
  filter(province == 'Marquesas') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas') %>%
  pull(id)

marq_div_all <- mean(dd[m_sp,ip_sp])

m_sp <- samples %>%
  filter(province == 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)

marq_div_pta <- mean(dd[m_sp,ip_sp])
```

Three main haplotypes represented `r format_list(top_3_n)` individuals across all biogeographic provinces ([@fig:fig_hap_coi_pgt]). The bulk of *P. goslinei* individuals (`r hap1$hi` specimens from both MHI and NWHI) shared a single haplotype with `r hap1$ip` *P. tapeinosoma* specimens from the Central Indo-Pacific, Central Pacific, and Polynesian provinces. The most common *P. tapeinosoma* haplotype represented `r hap2$ip` individuals spanning the Indo-Pacific realm from South Africa and the Red Sea to the Society Islands (every Indo-Pacific province except the Marquesas). `r english::Words(hap2$hi)` *P. goslinei* individual`r ifelse(hap2$hi > 1,'s','')` from the Northwestern Hawaiian Islands also shared this haplotype. The third most frequent haplotype almost entirely comprised individuals from the Marquesas, with the exception of one from the Society Islands. The remaining haplotypes all represented `r english::words(max_hap)` or fewer individuals each. The two most common haplotypes differed from one another by a single mutation and represented both species. A third haplotype shared across species contained a single individual of each and differed from the main Hawaiian haplotype by one mutation. Two *P. goslinei* individuals were more closely related to the main *P. tapeinosoma* haplotype than to the rest of the Hawaiian haplotypes, and two *P. tapeinosoma* (from the Central Indo-Pacific and Central Pacific) were more closely related to the main *P. goslinei* haplotype. The remaining haplotypes of both species, with the the exception of Marquesan haplotypes, were not more than a few steps diverged from either of the two most common haplotypes. All individuals from the Marquesas (here presented in its own biogeographic province) either shared or were linked to the third most frequent haplotype, and all were separated from the others of either species by 5--9 mutational steps (`r scales::percent(marq_div_all,0.01)` sequence divergence including both species, `r scales::percent(marq_div_pta,0.01)` for *P. tapeinosoma* only).

#### cytb {-}

```{r include=FALSE, dependson="haplotypes"}
hap <- haplotypes$pgt$cytb
h <- hap_summary(hap)
top_3 <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(haplotype)
top_3_n <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(n)
max_hap <- h$summary %>%
  arrange(desc(n)) %>%
  dplyr::slice(4:n()) %>%
  pull(n) %>%
  max()


hap1 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[1]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()

hap2 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[2]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()
```
There were `r h$n` unique cytb haplotypes among the `r nrow(hap$alignment)` sequences generated for the *P. goslinei*/*tapeinosoma* species pair, with `r h$shared` haplotypes that were shared in at least two individuals and `r h$shared_spp` shared across species ([@fig:fig_hap_cytb_pgt]). There were `r length(seg.sites(hap$alignment))` segregating sites, overall haplotype diversity (h) was `r num(hap.div(hap$alignment),0.001)` and nucleotide diversity ($\mathrm{\uppi}$) was `r num(nuc.div(hap$alignment),0.001)`. Summary statistics by province can be found in [@tbl:tbl_stats_cytb_pgt].

As in the COI dataset, the plurality of Hawaiian specimens (`r hap1$hi` individuals) were represented by a single haplotype, which also included `r english::words(hap1$ip)` *P. tapeinosoma* (from the Central Pacific and Polynesian provinces) ([@fig:fig_hap_cytb_pgt]). Unlike COI, *P. tapeinosoma* cytb haplotypes were not shared as broadly across provinces, with the Red Sea, Western Indian Ocean (WIO, and Marquesas) largely represented by unique haplotypes (although there was one haplotype shared by specimens from the WIO to Polynesia). However, singleton haplotypes from distant provinces were frequently diverged by only one or two mutations. There were no *P. goslinei* individuals represented by primarily *P. tapeinosoma* haplotypes, but two were more closely related to diverged haplotypes from the WIO or the Marquesas. The majority of cytb haplotypes were only separated by 1--3 mutational steps across both species, but there were two notably divergent haplogroups in the network: one representing all Marquesan individuals (as for COI) and one representing 7 from the WIO. The WIO haplogroup was diverged from the main *P. goslinei* haplotype by 5 mutations, with the Marquesan group a minimum of 12--18 steps further.

### *Plagiotremus ewaensis* and *P. rhinorhynchos* {-}

# Discussion {-}

This study of COI gene sequences from 327 Nemophini fishes supported the monophyly of most genera and species considered. *Meiacanthus* was identified as the sister genus to *Plagiotremus*, with *Xiphasia* as the sister to this two-genus group. Almost all species within *Plagiotremus* exhibited well-supported reciprocal monophyly with the exception being the group containing *P. azaleus*, *P. tapeinosoma*, and *P. goslinei*. Within this group, *P. azaleus* is very clearly diverged while *P. tapeinosoma* from the Marqueseas are more distinct from the rest of *P. tapeinosoma* than *P. goslinei* is as these latter two groups share several haplotypes.

Only one previously-published phylogeny has supported *Meiacanthus* as the sister genus to *Plagiotremus* [@linPhylogenyBiogeographyShallow2013]. All subsequent phylogenies found *Xiphasia* to be the sister genus of *Plagiotremus* while the position of *Meiacanthus* within Nemophini has varied [@hundtMultilocusMolecularPhylogeny2014; @casewellEvolutionFangsVenom2017; @hundtExtremeDentitionDoes2018]. These later differences have been attributed by some authors to be due to increased taxonomic coverage within *Meiacanthus* and other Nemophini genera [@hundtMultilocusMolecularPhylogeny2014]. While this COI phylogeny is limited by a lack of genetic resolution, it is plausible that the increased taxonomic resolution of this study led to the inference of different relationships. Relative to the previous multi-locus phylogenies, this study considered three times more *Plagiotremus* species and included considerable geographic variation by analyzing individuals from as many locations as possible across all genera and species. However, interpretation of the genus-level relationships observed here should only be considered speculative as they were not well supported overall. Increasing both the genetic and taxonomic resolution in future phylogenetic studies may further resolve these relationships.

In attempting to resolve the phylogenetic relationships between the "species groups" he defined within *Plagiotremus*, Smith-Vaniz [@smith-vanizSabertoothedBlenniesTribe1976] noted that:

"Rigid application of the principles of phylogenetic (cladistic) systematics dictates that the *laudandus* and *tapeinosoma* species groups be accorded a sister-group relationship because they share the greatest number of apomorphic characters…However…I believe their occurrence in both groups is a case of parallel evolution."

The results of this genetic study, however, support the "rigid cladistic" interpretation. The clearest divergence within *Plagiotremus* was between the "*rhinorhynchos* species group" and all others. Long internal branches separated these well-supported groups. This topology indicates that the previously-recognized subgenus Runula is polyphyletic as its constituent "*tapeinosoma* species group" is in fact more closely related to the "*laudandus* species group" (subgenus Musgravius) than to the "*rhinorhynchos* species group" that was also placed in Runula by Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976]. Other molecular phylogenies also support this alternative arrangement of "species groups" within *Plagiotremus*. Similar to this study, Hundt et al. [-@hundtMultilocusMolecularPhylogeny2014] found *P. tapeinsoma* to be more closely-related to *P. laudandus* than *P. rhinorhynchos* (although these were the only three *Plagiotremus* species considered in this study), and all phylogenetic analyses that have included *P. rhinorhynchos* put it on a rather long branch relative to other species within Nemophini [@linPhylogenyBiogeographyShallow2013; @hundtMultilocusMolecularPhylogeny2014; @hundtExtremeDentitionDoes2018]. These results suggest that the subgenus Runula should be dissolved. Further analyses including samples from the "*spilistius* species group" not considered here would provide a more complete picture of the relationships of all *Plagiotremus* species.

Both Hawaiian endemic species within *Plagiotremus* were considered by Smith-Vaniz to potentially only warrant subspecies recognition. The reciprocal monophyly of the Hawaiian endemic *P. ewaensis* and the widespread *P. rhinorhynchos* in this study validates the species-level status of the former. While Smith-Vaniz distinguished *P. ewaensis* and *P. rhinorhynchos* based on their color patterns and significantly different number of incisors, he more generally based his decision to uphold species-level recognition for both Hawaiian endemics based on the precedent set by Gosline and Brock [-@goslineHandbookHawaiianFishes1960]—the first of whom was honored by the naming of *P. goslinei*, and the second of whom initially described *P. ewaensis*. The arguments of Gosline and Brock were that: (1) there was no biological reason not to recognize the Hawaiian endemics as full species as there was no "intergradation" between them and their widespread relatives (i.e. intermediate or mixed forms), and (2) they were the most distinctive sub-groups in the Indo-West-Pacific. In light of the results of this study, the arguments of Gosline and Brock no longer hold for *P. goslinei*.

Regarding the first argument concerning intergradation, *P. goslinei* and *P. tapeinosoma* share multiple haplotypes. In more recent discussion of (full) species delimitation (i.e. in the era of molecular phylogenies) Smith-Vaniz et al. [@smith-vanizAggressiveMimicryAspidontus2020] argue (regarding *P. laudandus*/*flavus*) that:

"Shared haplotypes are characteristic of populations making up a single species but are not proof two populations are the same species. The exception would be when phenotypes are consistently different and the boundaries well-defined; in that case two populations sharing haplotypes may represent recently split species and/or established species with some low degree of hybridization".

While shared haplotypes between *P. goslinei* and *P. tapeinosoma* alone may not be sufficient to synonymize them (as they may still be diverging if truly isolated), the fact that two *P. goslinei* samples were more closely-related to *P. tapeinosoma* and two *P. tapeinsoma* samples from Kiribati and Palau were more closely-related to the main *P. goslinei* haplotype is evidence of recent migration between the two groups with supposedly different ranges. This potential recent mixing along with the presence of shared haplotypes imply that they are not demographically or evolutionarily distinct (yet, at least).

Regarding the second argument of Hawaiian endemics being the most distinct in the Indo-West-Pacific, *P. tapeinosoma* from the Marquesas is considerably more diverged from the rest of *P. tapeinosoma* than *P. goslinei* is, despite not also being considered a distinct species. Smith-Vaniz [@smith-vanizSabertoothedBlenniesTribe1976] recognized that Marquesan *P. tapeinosoma* had meristic values (fish trait counts/measurements) "at least as low as those of *goslinei*" that potentially warranted them subspecies status. Therefore, according to the criteria specified by previous authors to justify their own species-level taxonomic recognition of Hawaiian endemics [@goslineHandbookHawaiianFishes1960; @smith-vanizSabertoothedBlenniesTribe1976], *P. goslinei* should likely not be considered a distinct species from *P. tapeinosoma*. On the other hand, *P. tapeinosoma* samples on literal opposite sides of the planet shared the same COI haplotype while *P. goslinei* is almost completely diverged from the rest of the "*tapeinosoma* species group". Therefore, joint study of how phenotypic and genetic variation is partitioned geographically within the geographic "*tapeinosoma* species group" lineages (wider Indo-Pacific, Hawaiian, Marquesas, and Eastern Pacific) should be carried out to clarify their interrelationships and taxonomic statuses. This would likely result in *P. tapeinosoma* from the Marquesas being formally described as a distinct species given its complete genetic divergence from *P. tapeinosoma* outside of Polynesia and its meristic distinctness observed by Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976].

Although *P. azaleus* (in the Eastern Pacific) is considerably diverged from all other "*tapeinosoma* species group" lineages, the initial description of *P. goslinei* [@strasburgNotesBlennioidFishes1956] suggested that it may be a geographical intermediate of *P. azaleus* and *P. tapeinosoma*. The placement of *P. azaleus* as the sister group to *P. tapeinosoma* and *P. goslinei* in this study was not well supported, and other trees nested it within mixed groups of *P. tapeinosoma* and *P. goslinei*. Furthermore, although *P. azaleus* was closest to a *P. tapeinosoma* haplotype in the network, only one to two more steps separated it from multiple *P. goslinei* clusters—which is very few relative to its 43 steps from the rest of the network. Therefore, further genetic study of these lineages could determine whether *P. azaleus* originally colonized the Eastern Pacific through Hawai&#x02BB;i (sharing an ancestor with *P. goslinei* there as well) and contribute to the growing literature about whether isolated islands with high levels of endemism export biodiversity to other regions [@bowenOriginsTropicalMarine2013]. On the other hand, both the *P. azaleus* and Marquesan *P. tapeinosoma* haplotypes were linked to the rest of *P. tapeinosoma* through Indian Ocean haplotypes (rather than those from the Pacific which are much closer). This may indicate that the divergence of these lineages dates back to the closure of the Tethys Sea (that divided the tropical Atlantic and Indo-Pacific realms) as observed in many other reef fish groups [@floeterAtlanticReefFish2008].

The taxonomy and phylogeography of fangblennies was elucidated in this study by including more species than ever before in a molecular phylogeny of *Plagiotremus*. Although only one of the Hawaiian lineages seems to merit full species status, further study of the phenotypic and genetic variation within the "*tapeinosoma*" sub-group could lead to the description of distinct species in other regions, as well as provide further insight into how endemic marine species evolve in some of the world's most isolated archipelagos.

# Figures and Tables {-}

```{r fig-map, fig.cap='Map of *Plagiotremus* collection sites across the Indo-Pacific. Background shading indicates biogeographic realm, yellow: Western Indo-Pacific, green: Central Indo-Pacific, red: Eastern Indo-Pacific, orange: Tropical East Pacific.', results="asis", fig.width=8, fig.height=10, dependson="init"}
# define our fake antemeridian
ante_meridian <- 165

# define a long & slim polygon that overlaps the meridian line & set its CRS to match
# that of world
offset <- 180 - ante_meridian
polygon <- st_polygon(x = list(rbind(
  c(-0.0001 - offset, 90),
  c(0 - offset, 90),
  c(0 - offset, -90),
  c(-0.0001 - offset, -90),
  c(-0.0001 - offset, 90)
))) %>%
  st_sfc() %>%
  st_set_crs(4326)


# ocean background
offset <- ante_meridian+180

# our target crs transformation. maybe we did this twice but I'll leave it since it works
target_crs <- st_crs(str_glue("+proj=moll +x_0=0 +y_0=0 +lat_0=0 +lon_0={ante_meridian}"))


# realms we want to show on the map
map_regions <- regions %>%
  filter(str_detect(REALM,fixed("pacific",ignore_case = TRUE)) & !str_detect(REALM,fixed("temperate",ignore_case = TRUE)))

# Convert to polygon geometries, correct geometry errors, project to PCS,
# snap vertices, merge polygons by REALM, project back to original CRS
# this gets rid of the vertical line at 180, at least in theory
map_regions <- suppressWarnings(
  map_regions %>%
    st_cast("POLYGON") %>%
    st_make_valid() %>%
    st_transform(target_crs) %>%
    st_snap(x = ., y = ., tolerance = 165) %>%
    summarise(geometry = st_union(geometry), .by = REALM) %>%
    st_transform(4326) %>%
    st_shift_longitude() %>%
    st_transform(target_crs)
)

# create world map without overlaps and weird artifacts
worldrn <- suppressWarnings(
  ne_countries(scale = "medium", returnclass = "sf") %>%
    st_make_valid() %>%
    st_difference(polygon) %>%
    st_transform(crs = target_crs)
)


# make the ocean background fill
jitter = 0.01
lat <- c(seq(from=-90,to=90,by=0.25))
lon <- c(rep(offset-jitter,length(lat)),rep(offset+jitter,length(lat)))

oceans <- cbind(lon, c(lat, rev(lat))) %>%
  st_linestring() %>%
  st_cast("POLYGON") %>%
  st_sfc(crs=4326) %>%
  st_shift_longitude() %>%
  st_transform(crs=paste0("+proj=moll +lon_0=",ante_meridian," +lat_0=0 +x_0=0 +y_0=0"))

# create point data
plagiotremus_data <- all_samples %>%
  filter(species %in% unlist(species_pairs)) %>%
  filter(source == 'sequenced') %>%
  drop_na(lat,lon) %>%
  st_as_sf(coords=c('lon','lat'),remove=FALSE,crs=4326) %>%
  dplyr::rename(Species=species) %>%
  st_transform(target_crs) %>%
  st_jitter()

# get unique species
spp <- unique(plagiotremus_data$Species)

# make shape scale
shapos <- c(21:24) %>%
  set_names(spp)
# make fill scale
fillers <- paletteer_d('wesanderson::Darjeeling1',length(spp)) %>%
  set_names(spp)

# make two world plots with the different species pairs
plotz <- species_pairs %>%
  purrr::map(\(pair) {
    # get only species in pair
    pd <- plagiotremus_data %>%
      filter(Species %in% pair)

    # make the pllot
    ggplot() +
      geom_sf(data = oceans, fill="lightblue", alpha=0.3) +
      geom_sf(data = worldrn, aes(group = admin), fill="lightgrey") +
      geom_sf(data = map_regions, aes(fill = REALM),show.legend = FALSE, alpha = 0.3, color="grey77") +
      scale_fill_paletteer_d('waRhol::skull_76_1',guide="none") +
      new_scale_fill() +
      geom_sf(data = pd,  aes(fill=Species, shape=Species), size=2, color="black" ) +
      scale_fill_manual(values = fillers, guide = guide_legend(title=NULL)) +
      scale_shape_manual(values=shapos, guide = guide_legend(title=NULL)) +
      theme_bw() +
      theme(
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom",
        legend.key = element_blank(),
        legend.text = element_text(size=12,face='italic')
      )
  })

# reduce the plots with `/` so they're stacked on top of one another
plotz %>%
  purrr::reduce(`/`) +
  theme(plot.tag = element_text(face='bold',margin = margin(r=4,b=4)))
cat("{#fig:fig_map}")
```

\newpage

```{r fig-hap-coi-pgt, fig.cap='COI haplotype network of *Plagiotremus tapeinosoma* and *P. goslinei*. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.', results="asis", fig.width=16, fig.height=10, dependson="haplotypes"}
haplotypes$pgt$coi$plot
cat("{#fig:fig_hap_coi_pgt}")
```

\newpage

```{r fig-hap-cytb-pgt, fig.cap='cytb haplotype network of *Plagiotremus tapeinosoma* and *P. goslinei*. Abbreviations as in [@fig:fig_hap_coi_pgt].', results="asis", fig.width=16, fig.height=10, dependson="haplotypes"}
haplotypes$pgt$cytb$plot
cat("{#fig:fig_hap_cytb_pgt}")
```

\newpage

```{r fig-hap-coi-per, fig.cap='COI haplotype network of *Plagiotremus rhinorhynchos* and *P. ewaensis*. Abbreviations as in [@fig:fig_hap_coi_pgt].', results='asis', fig.width=16, fig.height=10, dependson='haplotypes'}
haplotypes$per$coi$plot
cat("{#fig:fig_hap_coi_per}")
```

\newpage

```{r trees, results="asis", dependson="alignment"}
# read complete COI tree and join sample data
tree_covg <- names(alignments$coi)

trees <- tree_covg %>%
  sort() %>%
  set_names() %>%
  purrr::map(\(covg) {
    tree <- read.mrbayes(here('data','trees',str_glue('coi_{covg}.nex.con.tre'))) %>%
      left_join(all_samples,by=c(label='id'))

    # drop all but one Omobranchus tip to simplify things
    to_drop <- tree %>%
      as_tibble() %>%
      filter(
        genus == 'Omobranchus',
        label != 'Oan-AMS266-08'
      ) %>%
      pull(label)

    tree <- tree %>%
      drop.tip(to_drop)

    # figure out our one omobranchus outgroup node
    outgroup <- tree %>%
      as_tibble() %>%
      filter(genus == 'Omobranchus') %>%
      pull(node)
    if (length(outgroup) > 1) {
      outgroup <- getMRCA(tree@phylo,outgroup)
    }

    # reroot the tree on the omobranchus node
    tree@phylo <- reroot(
      tree@phylo,
      outgroup,
      position = tree@phylo$edge.length[which(tree@phylo$edge[,2]==outgroup)]/2
    )

    # probs get set to NA when tree is rerooted
    # here we're doing essentially what figtree does
    tree <- tree %>%
      mutate(
        prob = case_when(
          !isTip & is.na(prob) ~ "1",
          .default = prob
        ),
        prob_cut = cut(as.numeric(prob),c(-Inf,0.9,0.99,Inf),labels=c('<0.9','≥0.9','≥0.99'))
      )

    # create groups to color the tree sections
    genus_groups <- tree %>%
      as_tibble() %>%
      filter(genus != 'Omobranchus') %>%
      split(.$genus) %>%
      imap(\(grp,genus) {
        unique(grp$label)
      })
    tree <- tree %>%
      mutate(lbl = case_when(
        genus == 'Omobranchus' ~ 'Omobranchus',
        .default = ''
      )) %>%
      groupOTU(genus_groups,'Genus')
    # make the groups be a factor
    attr(tree@phylo,'Genus') <- factor(attr(tree@phylo,'Genus'),levels=names(genus_groups))

    tree
  })
```

```{r fig-tree-genera, fig.cap='COI Bayesian consensus trees for *Plagiotremus* and closely-related genera. (A) Full tree based on 50% coverage alignment. (B) Genus-level tree based on full alignment (all sequences). (C) Genus-level tree based on 99% coverage alignment.', results="asis", fig.width=8, fig.height=5, dependson="trees"}
# full tree
ft <- 'c50'
# subtrees
st <- grep('c50',names(trees),invert = TRUE,value = TRUE)

# calculate MRCA of genera for clade labeling, excluding Omobranchus
genera <- trees %>%
  purrr::map(\(tree) {
    tree %>%
      as_tibble() %>%
      filter(genus != 'Omobranchus') %>%
      split(.$genus) %>%
      imap(\(grp,genus) {
        if (nrow(grp) > 1) {
          node <- getMRCA(tree@phylo,grp$node)
        } else {
          node <- grp$node
        }
        tibble(id=node,text=genus)
      }) %>%
      list_rbind()
  })

onodes <- trees %>%
  purrr::map(\(tree) {
    tree %>%
      as_tibble() %>%
      filter(genus == 'Omobranchus') %>%
      pull(node)
  })

# gcol <- c(paletteer_d('lisa::MaxErnst',3),'#000000')
# gcol <- c(paletteer_d('lisa::M_C_Escher',3),'#000000')
gcol <- c(set_names(paletteer_d('lisa::WassilyKandinsky_1',3),genera$c00$text),'0'='#000000')

gnodes <- genera[[ft]] %>%
  pull(id) %>%
  combn(2) %>%
  apply(MARGIN=2,FUN=\(nodes) {
    getMRCA(trees[[ft]]@phylo,nodes)
  }) %>%
  unique()

pointcol <- c('≥0.99' = 'black','≥0.9' = 'darkgrey')

# plot the tree
full_tree <- ggtree(trees %>% pluck(ft),aes(color=Genus),size=0.7) +
  geom_cladelab(mapping=aes(node=id,label=text,color=Genus),data=genera %>% pluck(ft), align=FALSE, fontface = "bold.italic", offset = 0.005)  +
  geom_cladelabel(onodes %>% pluck(ft),'bolditalic(Omobranchus)', offset = 0.005, parse=TRUE, size=3) +
  geom_point2(aes(fill=prob_cut,subset = node %in% gnodes),shape=21,size=2,show.legend = TRUE) +
  scale_y_reverse() +
  geom_rootedge(0.005) +
  scale_color_manual(values=gcol,guide='none') +
  scale_fill_manual(values=pointcol,drop=FALSE,name='Node support\n(posterior probability)') +
  theme(
    legend.position = 'inside',
    legend.justification = 'left',
    legend.position.inside = c(0,0.1),
    legend.background = element_blank()
  ) +
  xlim(0,0.55)


subtrees <- trees[st] %>%
  imap(\(tree,covg) {
    gg <- genera %>%
      pluck(covg)

    gnodes <- gg %>%
      pull(id) %>%
      combn(2) %>%
      apply(MARGIN=2,FUN=\(nodes) {
        getMRCA(tree@phylo,nodes)
      }) %>%
      unique()

    gg$id %>%
      purrr::reduce(\(g,node) g %>% collapse(node=node),.init=ggtree(tree,aes(color=Genus))) +
      geom_cladelab(mapping=aes(node=id,label=text,color=Genus),data=gg, align=FALSE, fontface = "bold.italic", offset = 0.001)  +
      geom_point2(aes(fill=prob_cut,subset = node %in% gnodes),shape=21,size=2) +
      geom_cladelabel(onodes %>% pluck(covg),'bolditalic(Omobranchus)', parse=TRUE) +
      geom_point2(aes(subset = (node %in% gg$id)),shape=18,size=3) +
      geom_rootedge(0.005) +
      xlim(NA,0.3) +
      scale_color_manual(values=gcol,guide='none')  +
      scale_fill_manual(values=pointcol,guide='none') +
      scale_y_reverse() +
      xlim(0,0.28)
  })

# plot layout
layout <- '
  11122
  11133
'
# build the plot
c(full_tree,subtrees) %>%
  purrr::reduce(`+`) +
  plot_layout(design=layout) &
  plot_annotation(tag_levels = plot_tags) &
  theme(plot.tag = element_text(face='bold'))

cat("{#fig:fig_tree_genera}")
```

\newpage

```{r fig-tree-plagiotremus, fig.cap='COI Bayesian consensus tree of relationships within *Plagiotremus* based on 50% coverage alignment. Dashed magenta edges indicate individuals collected in Hawai&#x02BB;i. Node labels indicate Bayesian posterior probability.', results="asis", fig.width=9, fig.height=7, dependson="trees"}
# drop non-Plagiotremus tips
to_drop <- trees$c50 %>%
  as_tibble() %>%
  filter(genus != 'Plagiotremus') %>%
  pull(label)

# filter the tree
tree <- trees$c50 %>%
  drop.tip(to_drop)
# save the tip labels
tips <- tree@phylo$tip.label

hap <- haplotypes %>%
  pluck('pgt','coi')

# get haplotypes
# generate haplotype index map
hindex <- attr(hap$haplotypes,'index') %>%
  imap(\(idx,h) {
    tibble(
      haplotype = h,
      sequence = rownames(hap$alignment)[idx]
    )
  }) %>%
  list_rbind()

# join metadata to tree
# and try to define some geographic range info
tree <- tree %>%
  left_join(hindex,by=c(label='sequence')) %>%
  left_join(hap$samples,by=c(label='id'))

# get tips with marquesan haplotypes
marquesan_tips <- tree %>%
  as_tibble() %>%
  group_by(haplotype) %>%
  filter(province == 'Marquesas') %>%
  ungroup() %>%
  pull(label)

# get root node and all children for marquesan haplotypes
marquesan_node <- getMRCA(tree@phylo,marquesan_tips)
marquesas <- ggtree:::getSubtree(tree@phylo,marquesan_node)

# make nice-looking clades we can label
clades <- tree %>%
  as_tibble() %>%
  drop_na(label) %>%
  mutate(
    grp = case_when(
      str_detect(species,'rhinorhynchos') ~ 'P. rhinorhynchos',
      str_detect(species,'ewaensis') ~ 'P. ewaensis',
      str_detect(species,'ewaensis') ~ 'P. ewaensis',
      str_detect(species,'flavus|laudandus') ~ 'P. laudandus/flavus',
      str_detect(species,'townsendi') ~ 'P. townsendi',
      str_detect(species,'phenax') ~ 'P. phenax',
      str_detect(species,'azaleus') ~ 'P. azaleus',
      str_detect(species,'tapeinosoma|goslinei') ~ 'P. tapeinosoma/goslinei',
      .default = ''
    )
  ) %>%
  mutate(
    grp = case_when(
      node %in% marquesas & !is.na(label) ~ 'P. cf. tapeinosoma\n(Marquesas)',
      .default = grp
    ),
  ) %>%
  split(.$grp) %>%
  imap(\(grp,clade) {
    if (nrow(grp) > 1) {
      node <- getMRCA(tree@phylo,grp$node)
    } else {
      node <- grp$node
    }
    tibble(id=node,text=clade)
  }) %>%
  list_rbind() %>%
  mutate(
    offset = case_when(
      str_detect(text,'azaleus') ~ -0.001,
      str_detect(text,'goslinei') ~ -0.08,
      .default = 0
    )
  )

# get the rode node, because we're gonna set its probability to 1
rootnode <- getMRCA(tree@phylo,tree@phylo$tip.label)

# update some of the internal data for the tree
tree <- tree %>%
  mutate(
    prob = case_when(
      node == rootnode ~ 1,
      .default = as.numeric(prob)
    ),
    posterior = cut(
      prob,
      c(-Inf,0.60,0.75,0.999,Inf),
      labels=c("<0.60","0.60-0.75","0.75-0.95","1")
    )
  )

# figure out which nodes to put points on
nodepoints <- clades %>%
  pull(id) %>%
  combn(2) %>%
  apply(MARGIN=2,FUN=\(nodes) {
    getMRCA(tree@phylo,nodes)
  }) %>%
  unique()

# make colors for our node points
clr_map <- c("<0.60"="white","0.60-0.75"="gray75","0.75-0.95"="gray40","1"="black")

tree <- groupOTU(tree, tree@phylo$tip.label[grep("Pew|Pgo", tree@phylo$tip.label, invert = TRUE)])

ggtree(tree,aes(color=group,linetype=group)) +
  scale_y_reverse() +
  scale_color_manual(values = c('#C82B84','black'),guide='none') +
  scale_linetype_manual(values=c(2,1), guide='none') +
  geom_cladelab(mapping=aes(node=id,label=text,offset=offset),data=clades, parse=TRUE,fontface='bold.italic', barsize=1) +
  geom_rootedge(0.005) +
  geom_nodelab(aes(subset=(node %in% nodepoints),label=round(prob,2)),hjust=0,geom='text',fontface='bold',size=3) +#nudge_x = -0.003, nudge_y=1.1) +
  theme(legend.position = c(0.1,0.25)) +
  xlim(NA,0.25) +
  geom_treescale(y=-135,x=0,offset=1)
cat("{#fig:fig_tree_plagiotremus}")
```

\newpage

```{r tbl-collection,results="asis",dependson="init"}
tbl_collection <- original_samples %>%
  left_join(plagiotremus_data,by="id",suffix=c("","_sequenced")) %>%
  group_by(species) %>%
  summarise(
    Range = ifelse(cur_group()$species %in% hawaii_species,'Hawai&#x02BB;i','Indo-Pacific'),
    N = n(),
    `COI sequences` = sum(sequenced_coi),
    `cytb sequences` = sum(sequenced_cytb)
  ) %>%
  mutate(species = str_glue("*{species}*")) %>%
  dplyr::rename(Species = species)
print_table(tbl_collection,"tbl_collection","Specimen collection summary.",hdr_case=NA,accuracy=1)
```

\newpage

```{r tbl-stats-coi-pgt, results="asis", dependson="haplotypes"}
hap <- haplotypes$pgt$coi

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `$\\mathrm{\\uppi}$` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    sig = tajima.test(hap$alignment[id,])$Pval.beta < 0.05,
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(D = case_when(
    sig ~ str_glue('**{round(D,3)}**'),
    is.nan(D) ~ NA,
    .default = as.character(round(D,3))
  )) %>%
  arrange(llon) %>%
  select(-c(sig,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_coi_pgt","COI population genetic summary statistics for *P. goslinei*/*tapeinosoma* across marine biogeographic provinces (n = number of samples, S = number of segregating sites, H = number of unique haplotypes, h = haplotype diversity, $\\mathrm{\\uppi}$ = nucleotide diversity, D = Tajima's D; results in bold are statistically significant).",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-stats-cytb-pgt, results="asis", dependson="haplotypes"}
hap <- haplotypes$pgt$cytb

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `$\\mathrm{\\uppi}$` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    sig = tajima.test(hap$alignment[id,])$Pval.beta < 0.05,
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(D = case_when(
    sig ~ str_glue('**{round(D,3)}**'),
    is.nan(D) ~ NA,
    .default = as.character(round(D,3))
  )) %>%
  arrange(llon) %>%
  select(-c(sig,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_cytb_pgt","cytb population genetic summary statistics for *P. goslinei*/*tapeinosoma* across marine biogeographic provinces. Column names as in [@tbl:tbl_stats_coi_pgt].",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-stats-coi-per, results="asis", dependson="haplotypes"}
hap <- haplotypes$per$coi

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `$\\mathrm{\\uppi}$` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    sig = tajima.test(hap$alignment[id,])$Pval.beta < 0.05,
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(D = case_when(
    sig ~ str_glue('**{round(D,3)}**'),
    is.nan(D) ~ NA,
    .default = as.character(round(D,3))
  )) %>%
  arrange(llon) %>%
  select(-c(sig,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_coi_per","COI population genetic summary statistics for *P. ewaensis*/*rhinorhynchos* across marine biogeographic province. Column names as in [@tbl:tbl_stats_coi_pgt].",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r amova-tbl, include=FALSE}
amova_tbl <- function(mova,roundr=3,adjust=FALSE, alpha=0.05, upper_p = FALSE) {
  phi <- mova$PhiST
  pval <- mova$p

  if (adjust != FALSE & adjust %in% p.adjust.methods) {
    pval[lower.tri(pval)] <- p.adjust(pval[lower.tri(pval)],adjust)
  }

  if (upper_p) {
    phi[upper.tri(phi)] <- t(pval)[upper.tri(t(pval))]
  }

  sig_idx <- !is.na(pval) & pval < alpha
  phi <- round(phi,roundr)
  sig_phi <- phi[sig_idx]
  mode(phi) <- 'character'
  phi[sig_idx] <- str_glue('**{sig_phi}**')
  as.data.frame(phi)
}
```

```{r tbl-amova-pgt-coi, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$pgt$coi$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_pgt_coi","Pairwise COI $\\upPhi_\\mathrm{st}$ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where N &gt; 5) for *Plagiotremus goslinei*/*P. tapeinosoma*. Significant comparisons appear in bold. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-amova-pgt-cytb, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$pgt$cytb$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_pgt_cytb","Pairwise cytb $\\upPhi_\\mathrm{st}$ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where N &gt; 5) for *Plagiotremus goslinei*/*P. tapeinosoma*. Significant comparisons appear in bold.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-amova-per-coi, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$per$coi$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_per_coi","Pairwise COI $\\upPhi_\\mathrm{st}$ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where N &gt; 5) for *Plagiotremus ewaensis*/*P. rhinorhynchos*. Significant comparisons appear in bold.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

# References {-}

<!-- show main bibliography -->
<div class="multi-refs"></div>

\newpage

# Supplemental material {label="S"}

```{r fig-supp-covg, fig.cap='COI alignment coverage by dataset filtering level.', results="asis", fig.width=6, fig.height=7, results="asis", dependson="alignment"}
covg_data <- alignments %>%
  pluck('coi') %>%
  imap(\(alignment,ds) {
    covg <- apply(alignment,2,\(bases) sum(bases != 4) / length(bases))
    tibble(pos = seq_along(covg), coverage = covg, dataset = ds)
  }) %>%
  list_rbind() %>%
  mutate(dataset = fct_reorder(dataset,-parse_number(dataset)))

verts <- covg_data %>%
  group_by(dataset) %>%
  summarise(min = min(pos), max = max(pos), first_cov=dplyr::first(coverage), last_cov=dplyr::last(coverage))
ggplot(covg_data) +
  geom_area(aes(x = pos, y = coverage, fill=dataset),alpha=0.7, stat="identity",position="identity") +
  geom_line(aes(x = pos, y = coverage, color = dataset)) +
  geom_segment(aes(x=min,xend=min,y=0,yend=first_cov,color=dataset),data=verts) +
  geom_segment(aes(x=max,xend=max,y=0,yend=last_cov,color=dataset),data=verts)  +
  facet_wrap(~dataset,ncol=1) +
  labs(x='Nucleotide position',y='Coverage (%)') +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    panel.grid = element_line(color="#f0f0f0")
  ) +
  guides(color='none',fill = guide_legend(override.aes = list(alpha=1))) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_paletteer_d('tvthemes::Pearl',direction = -1,labels=c('99%','50%','Full dataset'),name='Filtering') +
  scale_color_paletteer_d('tvthemes::Pearl',direction = -1,labels=c('99%','50%','Full dataset'))
cat("{#fig:fig_supp_covg}")
```

\newpage

```{r tbl-supp-bold-seqs,results="asis",dependson="bold-seqs"}
bold_seq_tbl <- bold_seqs$coi$samples %>%
  select(species,processid,bin_uri,genbank_accession,country,catalognum) %>%
  dplyr::rename(
    Species=species,
    `Process ID`=processid,
    `BIN`=bin_uri,
    `GenBank Accession`=genbank_accession,
    Country=country,
    `Catalog Number`=catalognum
  ) %>%
  arrange(Species)
print_table(bold_seq_tbl,'tbl_supp_bold_seqs','Summary of external sequences used in this study.',hdr_case=NA)
```

\newpage

```{r tbl-supp-alignment-models, results="asis", dependson="init"}
print_table(here('data','tables','models.csv'),'tbl_supp_models','Summary of best-fit paritioning models by COI alignment coverage')
```

\newpage

```{r tbl-supp-dist-pgt-coi, results="asis", dependson="dna-distances"}
print_table(dna_distances$pgt$coi,id="tbl_supp_dist_pgt_coi","Mean COI Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus goslinei*/*P. tapeinosoma*. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-supp-dist-pgt-cytb, results="asis", dependson="dna-distances"}
print_table(dna_distances$pgt$cytb,id="tbl_supp_dist_pgt_cytb","Mean cytb Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus goslinei*/*P. tapeinosoma*.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-supp-dist-per-coi, results="asis", dependson="dna-distances"}
print_table(dna_distances$per$coi,id="tbl_supp_dist_per_coi","Mean COI Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus ewaensis*/*P. rhynorhynchos*.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```