---
title: "Origins and relationships of endemic Hawaiian saber-toothed blenny genus *Plagiotremus* (Blennioidei: Blenniidae)"
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Cameron A.J. Walsh:
      institute:
        - himb
      equal_contributor: yes
  - Randall K. Kosaki:
      institute:
        - pmnm
  - Brian W. Bowen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - pmnm: Papah&#257;naumoku&#257;kea Marine National Monument, National Oceanic and Atmospheric Administration
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
link-citations: true
linkReferences: true
tblPrefix: Table
figPrefix:
  - "Figure"
  - "Figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
lofTitle: |
  # Figure legends {-}
lofItemTitle: |
  Fig.
lofItemTemplate: |
  $$lofItemTitle$$ $$i$$$$listItemTitleDelim$$ $$t$$
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
fig_caption: true
---

### Keywords {-}
[blennies, endemism, Hawai&#x02BB;i, phylogeography, phylogenetics]{custom-style="mono"}

\newpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path="output/figures/",
  dev="svg",
  dpi=600,
  message=FALSE,
  warning=FALSE,
  cache=TRUE,
  cache.path = "cache/",
  cache.comments=FALSE
)

library(tidyverse)
library(pegas)
library(here)
library(Biostrings)
library(msa)
library(fs)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggnewscale)
library(pander)
library(paletteer)
library(patchwork)
library(ggfun)
library(phytools)
library(ggtree)
library(treeio)
source('spp.R')
```

```{r init, include=FALSE}
read_fasta <- function(x,aligned=FALSE) read.dna(x,"fasta",as.matrix=aligned)
write_fasta <- function(x,file) write.dna(x,file,"fasta",nbcol = -1,colsep = "")

spp <- Taxa$new('Plagiotremus')


all_samples <- read_csv(here("data","samples.csv"))
original_samples <- all_samples
markers <- c("coi","cytb")

# load marine ecoregions
# TODO: figure out whether shifting longitudes here breaks anything in the map plot
regions <- st_read(here("data","gis","ecoregions","meow_ecos.shp"),quiet = TRUE) %>%
  st_shift_longitude() %>%
  group_by(ECOREGION) %>%
  mutate(middle = max(st_coordinates(st_shift_longitude(st_combine(geometry)))[,'X']) )  %>%
  ungroup()

scounts <- original_samples %>%
  separate_wider_delim(cols = "species",delim = " ",names=c("genus","species")) %>%
  mutate(id = str_glue("{str_sub(genus,1,1)}{str_sub(species,1,2)}")) %>%
  count(id) %>%
  deframe()


# stolen from insect: join two DNAbins together
join_seq <- function(...){
  dots <- list(...)
  nlsts <- length(dots)
  DNA <- any(sapply(dots, class) == "DNAbin")
  if(nlsts == 0) return(NULL)
  if(nlsts == 1) return(dots[[1]])
  islist <- sapply(dots, is.list)
  for(i in which(!islist)){
    tmpattr <- attributes(dots[[i]])
    attributes(dots[[i]]) <- NULL
    dots[[i]] <- list(dots[[i]])
    attributes(dots[[i]]) <- tmpattr
  }
  dots <- lapply(dots, unclass)
  findattr <- function(x){
    names(attributes(x))[sapply(attributes(x), length) == length(x)]
  }
  attrlist <- lapply(dots, findattr)
  ual <- unique(unlist(attrlist, use.names = FALSE))
  validattrs <- ual[sapply(ual, function(e) all(sapply(attrlist, function(g) e %in% g)))]
  validattrs <- validattrs[!validattrs %in% c("names", "class")]
  res <- unlist(dots, recursive = FALSE, use.names = TRUE)
  for(i in validattrs){
    attr(res, i) <- unlist(lapply(dots, attr, i), use.names = FALSE)
  }
  class(res) <- "DNAbin"
  return(res)
}

# format percents and numbers
pc <- function(x,a=0.01) scales::percent_format(accuracy = a)(x)
num <- function(x,a=0.01) scales::label_comma(accuracy=a)(x)

# do a sentence case transformation, optionally preserving some case
sentence_case <- function(s) {
  str_match_all(s,'\\{(.+?)\\}') %>%
    imap_chr(\(matches,i) {
      matches %>%
        array_branch(1) %>%
        purrr::reduce(\(s,r) {
          str_replace_all(s,fixed(r[1],ignore_case = TRUE),r[2])
        },.init = str_to_sentence(s[i]))
    })
}

# format a markdown grid table from a data frame or file
print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf, roundr=FALSE, na="") {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

  if (inherits(tbl,"character")) {
    if (id == "") {
      id <- sub(pattern = "(.*?)\\..*$", replacement = "\\1", basename(file))
    }
    tbl <- read.csv(tbl,check.names=FALSE,stringsAsFactors=FALSE)
  } else if (inherits(tbl,"data.frame")) {
    if (id == "") stop("please specify table id")
  } else {
    stop("'tbl' must be a character or a data frame")
  }

  fmt <- function(x,r,a) {
    if (is.numeric(x) & is.numeric(r) & !isFALSE(r))
      round(x,r)
    else if (is.numeric(x))
      num(x,a)
    else
      x
  }

  tbl <- tbl %>%
    select(-any_of(hide)) %>%
    mutate(
      across(
        where(~format_numbers & is.numeric(.x) & !is.integer(.x)),
        ~fmt(.x,roundr,accuracy)
      )
    )


  # can also use rename_with(str_to_sentence)
  names(tbl) <- switch(
    str_to_lower(hdr_case),
    sentence = sentence_case(names(tbl)),
    title = str_to_title(names(tbl)),
    upper = str_to_upper(names(tbl)),
    lower = str_to_lower(names(tbl)),
    names(tbl)
  )

  caption <- sprintf("%s {#tbl:%s}",caption,id)

  table_str <- pandoc.table.return(
    tbl,
    keep.line.breaks=TRUE,
    caption=NULL,
    style="multiline",
    split.cells=cell_width,
    split.tables=Inf,
    justify='left',
    missing=na
  )
  table_str <- table_str[table_str != ""]
  cat(paste0(table_str,collapse="\n"),"\n\n",caption,"\n\n",sep="")
}

# little list of our sister species pairs, used for filtering datasets
species_pairs <- list(pgt=c('Plagiotremus goslinei','Plagiotremus tapeinosoma'),per=c('Plagiotremus ewaensis','Plagiotremus rhinorhynchos'))
hawaii_species <- list(pgo='Plagiotremus goslinei',pew='Plagiotremus ewaensis')

# map of various ecoregion realms to color palettes
palette_map <- c(
  'Red Sea' = 'RColorBrewer::Reds',
  'Western Indian Ocean' = 'RColorBrewer::Oranges',
  'Central Indo-Pacific' = 'beyonce::X59',
  'Central Indo Pacific' = 'beyonce::X59',
  'Central Pacific' = 'RColorBrewer::Greens',
  'Hawaii' = 'ggsci::purple_material',
  'Marquesas' = 'Redmonder::dPBIYlBu',
  'Polynesian Province' = 'RColorBrewer::Blues',
  'Tropical East Pacific' = 'beyonce::X34'
)

# make plot tags
plot_tags <- list(str_glue('({letters})'))
```

```{r util, include=FALSE}
# do an english comma-separated list
format_list <- function(l,fmt=function(x) x,word="and") {
  l <- fmt(l)
  if(length(l) > 1) l[length(l)] <- str_c(str_c(word," "),l[length(l)])
  if (length(l) > 2) {
    str_c(l,collapse=", ")
  } else {
    str_c(l,collapse=" ")
  }
}

# get elements of a named vector minus certain ones
range_minus <- function(x,minus='') {
  x[setdiff(names(x),minus)]
}

# format a p-value nicely
pval <- function(p) {
  case_when(
    p < 0.001 ~ "*p* \\< 0.001",
    p < 0.01 ~ "*p* \\< 0.01",
    TRUE ~ as.character(str_glue("*p* = {num(p,0.01)}"))
  )
}

# get p-value from amova and format it
amova_pval <- function(a) {
  pval(a$varcomp$P.value[1])
}

# get phist from amova
phist <- function(a) {
  getPhi(a$varcomp$sigma2)[1,1]
}

# format a range of numbers
format_range <- function(vals,roundr=FALSE,middle='',rev=FALSE) {
  f <- range(vals,na.rm=TRUE)
  if (rev) f <- rev(f)
  if (!isFALSE(roundr)) f <- round(f,roundr)
  if (middle == '') {
    str_glue("{f[1]}--{f[2]}")
  } else {
    str_glue("{f[1]} &le; {middle} &le; {f[2]}")
  }
}
```

```{r bold-seqs, dependson="init", include=FALSE}
library(bold)
# remove: Pbr-GBMTG324-16, Pbr-GBMNA15372-19
# remove: Pbr-GBMTG324-16
# bold_seqs_raw <- bold_seqspec(taxon=c("Plagiotremus","Xiphasia","Meiacanthus","Petroscirtes","Aspidontus"))
bold_seqs_raw <- bold_seqspec(taxon=c("Plagiotremus","Xiphasia","Meiacanthus","Omobranchus anolius"))

to_filter <- c('Pbr-GBMTG324-16')

bold_tbl <- bold_seqs_raw %>%
  as_tibble() %>%
  filter(
    nucleotides != "",
    species_name != "",
    !str_detect(species_name,"(sp|cf)\\."),
    !str_detect(processid,'FUT225') # this is a typo, it's a damselfish with the same specific epithet
  ) %>%
  separate_wider_delim(species_name,delim=" ",names=c(NA,"spp"),cols_remove = FALSE) %>%
  mutate(
    id = str_glue("{str_sub(genus_name,1,1)}{str_sub(spp,1,2)}-{processid}"),
    nucleotides = str_replace(nucleotides,"^-+",""),
    nucleotides = str_replace(nucleotides,"-+$",""),
    nucleotides = str_replace(nucleotides,"N+$",""),
    nucleotides = str_replace(nucleotides,"^N+",""),
    source="bold",
    sequenced_coi=1,
    sequenced_cytb=0
  ) %>%
  distinct(id,.keep_all = TRUE) %>%
  filter(str_detect(markercode,'COI')) %>%
  filter(!(id %in% to_filter)) %>%
  select(id,species=species_name,country,locality=exactsite,sequenced_coi,sequenced_cytb,lon,lat,source,nucleotides,bin_uri,processid,genbank_accession,catalognum)

bold_dna <- map2(bold_tbl$id,bold_tbl$nucleotides,~{
  c(str_glue(">{.x}"),.y)
}) %>%
  unlist() %>%
  textConnection() %>%
  read_fasta(aligned=FALSE)

bold_seqs <- list(
  coi = list(
    samples = bold_tbl %>% select(-nucleotides),
    sequences = bold_dna
  )
)

all_samples <- all_samples %>%
  bind_rows(bold_tbl %>% select(-nucleotides)) %>%
  separate_wider_delim(species,delim=' ',names=c('genus','species'),cols_remove = FALSE)
```

```{r local-seqs, dependson="init", include=FALSE}
local_seqs <- markers %>%
  set_names() %>%
  purrr::map( \(marker) {
    seqs <- read_fasta(here("data",str_glue("plagiotremus-{marker}.fasta")),aligned=FALSE)

    # filter metadata to match seqs
    samples <- all_samples %>%
      dplyr::filter(id %in% labels(seqs)) %>%
      arrange(id)

    list(samples = samples, sequences = seqs)
  })
```

```{r alignment, dependson=c("bold-seqs","local-seqs"), include=FALSE}
alignments <- markers %>%
  set_names() %>%
  purrr::map(\(marker) {
    ss <- local_seqs %>%
      pluck(marker,'sequences')
    if (exists(marker,where=bold_seqs)) {
      s2 <- bold_seqs %>%
        pluck(marker,'sequences')
      ss <- join_seq(ss,s2)
    }

    al <- ss %>%
      as.character() %>%
      lapply(paste0,collapse="") %>%
      unlist() %>%
      DNAStringSet() %>%
      msa(.,method="ClustalOmega",gapOpening = 2000,gapExtension = 1000) %>%
      as.DNAbin()

    write_fasta(al,here("cache",str_glue("{marker}_aligned.fasta")))

    covg <- apply(al,2,\(x) (sum(x != 4) / length(x)))
    list(c50 = 0.5, c99 = 0.99, c00 = 0) %>%
      imap(\(cov,idx) {
        p1 <- min(which(covg >= cov))
        p2 <- max(which(covg >= cov))
        ll <- al[,p1:p2]
        write_fasta(ll,here('output','alignments',str_glue('{marker}_{idx}.fasta')))
        return(ll)
      })
  })

sample_data <- markers %>%
  set_names() %>%
  purrr::map(\(marker) {
    all_samples %>%
      filter(id %in% labels(alignments %>% pluck(marker,'c00')))
  })

# cytb_samples <- all_samples %>% filter(id %in% labels(alignments$cytb$c00))
cycount <- sample_data$cytb %>%
  dplyr::count(source) %>%
  deframe()

cocount <- sample_data$coi%>%
  dplyr::count(source) %>%
  deframe()

coi_len <- purrr::map(alignments$coi,ncol)
cytb_len <- purrr::map(alignments$cytb,ncol)
```

```{r haplotypes, include=FALSE, dependson="alignment"}
# map MEOW provinces to kulbicki provinces
province_map <- c(
  'Agulhas' = 'Western Indian Ocean',
  'Red Sea and Gulf of Aden' = 'Red Sea',
  'Western Indian Ocean' = 'Western Indian Ocean',
  'Central Indian Ocean Islands' = 'Western Indian Ocean',
  'South Kuroshio' = 'Central Indo-Pacific',
  'Tropical Northwestern Pacific' = 'Central Indo-Pacific',
  'Northeast Australian Shelf' =  'Central Indo-Pacific',
  'Tropical Southwestern Pacific' = 'Central Pacific',
  'Marshall, Gilbert and Ellis Islands' = 'Central Pacific',
  'Central Polynesia' = 'Central Pacific',
  'Southeast Polynesia' = 'Polynesian Province',
  'Hawaii' = 'Hawaii',
  'Marquesas' = 'Marquesas',
  'Java Transitional' = 'Central Indo-Pacific',
  'Western Coral Triangle' = 'Central Indo-Pacific'
)

kprovince_map <- c(
  'Red Sea and Gulf of Aden' = 'Red Sea',
  'Agulhas' = 'Western Indian Ocean',
  'Western Indian Ocean' = 'Western Indian Ocean',
  'Central Indian Ocean Islands' = 'Western Indian Ocean',
  'South Kuroshio' = 'Central Indo-Pacific',
  'Tropical Northwestern Pacific' = 'Central Indo-Pacific',
  'Northeast Australian Shelf' =  'Central Indo-Pacific',
  'Java Transitional' = 'Central Indo-Pacific',
  'Western Coral Triangle' = 'Central Indo-Pacific',
  'Tropical Southwestern Pacific' = 'Central Pacific',
  'Marshall, Gilbert and Ellis Islands' = 'Central Pacific',
  'Central Polynesia' = 'Central Pacific',
  'Southeast Polynesia' = 'Polynesian Province',
  'Marquesas' = 'Polynesian Province',
  'Hawaii' = 'Hawaii'
)

province_abbrev <- c(
  'Western Indian Ocean' = 'WIO',
  'Western Indian' = 'WIO',
  'Red Sea' = 'RS',
  'Central Indo-Pacific' = 'CIP',
  'Central Indo Pacific' = 'CIP',
  'Central Pacific' = 'CP',
  'Polynesian' = 'PL',
  'Polynesian Province' = 'PL',
  'Main Hawaiian Islands' = 'MHI',
  'Northwestern Hawaiian Islands' = 'NWHI',
  'Hawaii' = 'HW',
  'Marquesan' = 'MQ',
  'Marquesas' = 'MQ',
  'Tropical East Pacific' = 'TEP'
)

wrap_fac <- function(f,w=20) {
  factor(
    str_wrap(f,w),
    levels = str_wrap(levels(f),w)
  )
}

plagiotremus_data <- all_samples %>%
  filter(species %in% unlist(species_pairs)) %>%
  drop_na(lon,lat) %>%
  st_as_sf(coords=c('lon','lat'),remove=FALSE,crs=4326) %>%
  st_join(regions,st_within) %>%
  mutate(
    llon = if_else(middle < 0,middle+360,middle),
    ecoregion = case_when(
      ECOREGION == "Hawaii" & nwhi ~ "Northwestern Hawaiian Islands",
      ECOREGION == "Hawaii" & !nwhi ~ "Main Hawaiian Islands",
      str_detect(ECOREGION,regex('red sea',ignore_case = TRUE)) ~ 'Red Sea',
      .default = ECOREGION
    ),
    province = case_when(
      ECOREGION == "Hawaii" & nwhi ~ "Northwestern Hawaiian Islands",
      ECOREGION == "Hawaii" & !nwhi ~ "Main Hawaiian Islands",
      .default = province_map[PROVINCE]
    ),
    prov = province_abbrev[province],
    kulbicki_province = kprovince_map[PROVINCE],
    kprov = province_abbrev[kulbicki_province],
    old_color_category = fct_reorder(
      case_when(
        str_detect(ecoregion,'Hawaii') ~ 'Hawaii',
        str_detect(ecoregion,'Marquesas') ~ 'Marquesas',
        .default = REALM
      ),
      llon
    ),
    color_category = fct_reorder(kulbicki_province,llon)
  )  %>%
  arrange(old_color_category,llon) %>%
  mutate(
    n = row_number(),
    ecoregion = fct_reorder(ecoregion,n),
    province = fct_reorder(province,n),
    kulbicki_province = fct_reorder(kulbicki_province,n),
    prov = fct_reorder(prov,n),
    ecoregion = fct_relevel(ecoregion,c('Main Hawaiian Islands','Northwestern Hawaiian Islands'),after=Inf),
    kprov = fct_relevel(fct_reorder(kprov,n),"HW",after=Inf)
  ) %>%
  as_tibble()

master_palette <- plagiotremus_data %>%
  as_tibble() %>%
  count(color_category,ecoregion) %>%
  group_by(color_category) %>%
  mutate( color = paletteer_d( palette_map[as.character(cur_group()$color_category)],direction=-1)[1:n()], n = n() ) %>%
  ungroup()

# add on cheeky rhinorhynchos/ewaensis-only "pairs" when calculating this bit
hpairs <- c(species_pairs,list(prh='Plagiotremus rhinorhynchos'),list(pew='Plagiotremus ewaensis'))
haplotypes <- hpairs %>%
  purrr::map(\(pair) {
    markers %>%
      set_names() %>%
      purrr::map(\(marker) {
        alignment <- alignments %>%
          pluck(marker,'c00')
        # filter by samples in this species pair
        samples <- plagiotremus_data %>%
          filter(
            species %in% pair,
            id %in% labels(alignment)
          )
        alignment <- alignment[samples$id,]
        # calculate per-base coverage
        covg <- apply(alignment,2,\(bases) sum(bases != 4) / length(bases))
        # keep only positions with >= 99% coverage
        alignment <- alignment[,covg >= 0.99]
        samples <- samples %>%
          filter(id %in% labels(alignment))
        hplot <- NULL
        if (nrow(alignment) > 0) {
          haplotypes <- pegas::haplotype(alignment)
          hnet <- haploNet(haplotypes)
          hindex <- attr(haplotypes,'index')
          ht_locations <- haploFreq(alignment,fac=samples$ecoregion)
          palette <- master_palette %>%
            filter(ecoregion %in% colnames(ht_locations)) %>%
            dplyr::slice(match(colnames(ht_locations),ecoregion)) %>%
            select(ecoregion,color) %>%
            deframe() %>%
            set_names(str_wrap(names(.),25))
          bar_legend <- ggplot(data = samples) +
            geom_bar(aes(y = ecoregion, fill = wrap_fac(ecoregion,25))) +
            scale_fill_manual(values = palette,name='Ecoregion') +
            scale_y_discrete(limits=rev) +
            theme_transparent() +
            theme(
              plot.title = element_blank(),
              plot.subtitle = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks.y = element_blank(),
              axis.title = element_blank(),
              panel.grid.major.y = element_blank(),
              panel.grid.minor = element_blank(),
              panel.border = element_blank(),
              panel.grid = element_blank(),
              strip.background = element_rect(fill="grey95",color="darkgrey"),
              legend.title = element_blank()
            ) +
            labs(title = "Legend", subtitle = "Number of samples")  +
            facet_wrap(
              ~kprov,
              scales="free_y",
              ncol=1,
              space="free_y",
              strip.position = "left"
            )

          leg <- get_legend(bar_legend)
          leg <- ggplot() +
            theme_void() +
            annotation_custom(leg) +
            theme(margins = margin(0,0,0,0))
          bar_legend <- bar_legend + theme(legend.position = "none")
          pd <- NULL
          pp <- str_replace_all(pair,'Plagiotremus ','')
          rds_file <- here('data','haplotypes',str_glue('{str_c(pp,collapse="-")}_{marker}_replot.rds'))
          if (file_exists(rds_file)) pd <- readRDS(rds_file)
          # plot(hnet, pie = ht_locations, bg = palette, # keep the same
          #      size=attr(hnet, "freq"), show.mutation = 1, length = 0.5, space = 0.75, # mess with
          #      # labels = F, threshold = 0, fast = F, legend = F)#, xy=pd)
          #      labels = F, threshold = 0, fast = F, legend = F, xy=pd)
          # pd <- replot()
          # saveRDS(pd,rds_file)
          hap_layout <- c(
            # area(top, left, bottom, right)
            area(1,5,100,100),  # plot
            area(64,1.5,99,30), # bar plot
            area(1,1.5,64,15)   # legend
          )

          # get haplotype network plot from base R
          net_plot <- wrap_elements(
            panel = ~plot(
              hnet, pie = ht_locations, bg = palette,
              size=attr(hnet, "freq"), show.mutation = 1,
              labels = F, threshold = 0, fast = F, legend = F, xy=pd
            ),
            clip = FALSE
          )

          joint_plot <- net_plot +
            bar_legend +
            leg +
            plot_layout(design=hap_layout)

          list(
            alignment = alignment,
            samples = samples,
            plot = joint_plot,
            haplotypes = haplotypes,
            network = hnet,
            frequencies = ht_locations
          )
        } else {
          NULL
        }
      })
  })

hap_summary <- function(h) {
  hindex <- attr(h$haplotypes,'index') %>%
    imap(\(idx,hap) {
      tibble(
        haplotype = hap,
        sequence = rownames(h$alignment)[idx]
      )
    }) %>%
    list_rbind() %>%
    left_join(h$samples,by=c(sequence='id'))

  h_summary = hindex %>%
    group_by(haplotype) %>%
    summarise(n = n(), spp = n_distinct(species), src = n_distinct(source), loc = n_distinct(province)) %>%
    ungroup()

  list(
    n = nrow(h$haplotypes),
    details = hindex,
    summary = h_summary,
    shared = h_summary %>%
      filter(n > 1) %>%
      nrow(),
    shared_spp = h_summary %>%
      filter(spp > 1) %>%
      nrow(),
    shared_src = h_summary %>%
      filter(src > 1) %>%
      nrow()
  )
}
```

```{r amovas, include=FALSE,dependson="haplotypes"}
# this goes slow as hell so we're manually caching it for
# convenience sake in rstudio
if (!file_exists(here('cache','amovas.rds'))) {
  amovas <- haplotypes %>%
    purrr::map(\(pair) {
      pair %>%
        purrr::map(\(hap) {
          if (!is.null(hap)) {
            samples <- hap$samples %>%
              add_count(province,name='p') %>%
              filter(p >= 5) %>%
              arrange(prov)
            dd <- dist.dna(hap$alignment[samples$id,])
            list(
              amova = amova(dd ~ prov,data=samples,nperm = 9999),
              pairwise = haplotypes::pairPhiST(dd,samples$prov,nperm=9999,showprogbar=FALSE)
            )
          }
        })
    })
  saveRDS(amovas,here('cache','amovas.rds'))
}
amovas <- readRDS(here('cache','amovas.rds'))
```

```{r dna-distances, include=FALSE, dependson="haplotypes"}
dna_distances <-
  haplotypes %>%
  imap(\(pair,spp) {
    pair %>%
      imap(\(hap,marker) {
        if (!is.null(hap)) {
          alignment <- hap$alignment
          samples <- hap$samples
          distances <- dist.dna(alignment) %>%
            as.matrix() %>%
            as_tibble(rownames='left') %>%
            pivot_longer(-left,names_to = 'right',values_to = 'dist') %>%
            left_join(samples %>% select(id,prov),by=c(left='id')) %>%
            left_join(samples %>% select(id,prov),by=c(right='id'),suffix=c('_left','_right')) %>%
            filter(left != right) %>%
            group_by(across(starts_with('prov'))) %>%
            summarise(dist = mean(dist,na.rm=TRUE)) %>%
            ungroup()  %>%
            mutate(dist = if_else(prov_left == prov_right,0,dist)) %>%
            pivot_wider(names_from = 'prov_right', values_from = 'dist')
          dist_matrix <- distances %>%
            column_to_rownames('prov_left') %>%
            as.matrix()
          mat <- dist_matrix
          diag(mat) <- NA
          mat[upper.tri(mat)] <- NA
          list(
            matrix = dist_matrix,
            tbl = mat %>%
            as.data.frame() %>%
            filter(!if_all(everything(),is.na)) %>%
            select(where(~!all(is.na(.x))))
          )
        }
      })
  })
```

# Abstract {-}

The Hawaiian archipelago is an endemism hotspot. Up to twenty-five percent of Hawaiian reef fish species are endemic, a higher proportion than anywhere else in the world. However, unlike terrestrial species where high endemism often arises through adaptive radiation (e.g. honeycreepers, fruit flies, silverswords, etc.)---marine endemism in Hawai&#x02BB;i is more evenly-distributed across taxa. Among reef fishes in Hawai&#x02BB;i, sister taxa rarely co-occur, and when they do, they exhibit a higher degree of range size asymmetry than co-occurring sister species elsewhere in the Indo-Pacific. In other words, the sister species of (small range size) Hawaiian endemics are typically widespread in the Indo-Pacific (large range size).

# Introduction {-}

The Hawaiian archipelago is a marine endemism hotspot, with up to twenty-five percent of its reef fishes found nowehere else on earth [@randallReefShoreFishes2007]. However, whereas terrestrial endemism in Hawai&#x02BB;i is often associated with adaptive radiation [e.g., honeycreepers: @freedEvolutionaryEcologyRadiation1987; fruit flies: @carsonChromosomalSequencesInterisland1983; silverswords: @witterAdaptiveRadiationGenetic1988, etc.], marine endemism is more evenly-distributed across taxa [@kayEndemismEvolutionHawaiian1987; @mundyChecklistFishesHawaiian2005]. Among Hawaiian reef fishes, sister taxa rarely co-occur and when they do, they often exhibit a higher degree of range size asymmetry than co-occurring sister species elsewhere in the Indo-Pacific [@gaboriauInfluenceGeographySpeciation2018]. Whether or not they co-occur, sister species of Hawaiian endemic fishes are often widespread in the Indo-Pacific [@gaboriauInfluenceGeographySpeciation2018; @randallZoogeographyShoreFishes1998; @hobanCirripectesMatatakaroNew2020].

Two examples of endemic reef fishes with widespread sister species are currently recognized within the sabertooth blenny genus `r spp$cite("Plagiotremus")` (Blennioidei: Blenniidae). `r spp$cite("Plagiotremus goslinei")` ([@fig:fig_plagiotremus[]{}a]) and `r spp$cite("Plagiotremus ewaensis",TRUE)` ([@fig:fig_plagiotremus[]{}b]) are endemic to the Hawaiian Islands and their respective sister taxa, `r spp$cite("Plagiotremus tapeinosoma",TRUE)` and `r spp$cite("Plagiotremus rhinorhynchos",TRUE)` are distributed across the tropical Indo-West-Pacific from the Red Sea to the Marquesas Islands [@smith-vanizSabertoothedBlenniesTribe1976; @randallReefShoreFishes2007]. A phylogenetic analysis of *Plagiotremus* using 22 parsimony-informative morphological characters suggested four species groups and three subgenera [@smith-vanizSabertoothedBlenniesTribe1976]. According to the analysis, the subgenus *Plagiotremus* contains<!-- only--> `r spp$cite("Plagiotremus spilistius")` and `r spp$cite("Plagiotremus iosodon",TRUE)`<!--. `r spp$cite("Plagiotremus celebesensis")` was originally included, but was synonymized with *P. spilistius* in a follow-up study [@smith-vanizSaberToothedBlenniesTribe1987].--> The subgenus *Musgravius*, recognized as the sister to subgenus *Plagiotremus*, contains the *laudandus* group comprising `r spp$cite("Plagiotremus laudandus")`, `r spp$cite("Plagiotremus phenax",TRUE)`, and `r spp$cite("Plagiotremus townsendi",TRUE)` [@hastingsRecognizingDiversityBlennioid2009a; @smith-vanizAggressiveMimicryAspidontus2020]. The names `r spp$cite("Plagiotremus flavus")` and *P. laudandus flavus* are currently considered junior synonyms of *P. laudandus* [@smith-vanizAggressiveMimicryAspidontus2020; @frickeEschmeyersCatalogFishes2025]. Subgenus *Runula*, which is sister to the other subgenera, contains the *rhinorhynchos* group (*P. rhinorhynchos* and *P. ewaensis*) and the *tapeinosoma* group (*P. tapeinosoma*, *P. goslinei*, and `r spp$cite("Plagiotremus azaleus",TRUE)`) [@smith-vanizSabertoothedBlenniesTribe1976]. Although Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976; -@smith-vanizSaberToothedBlenniesTribe1987] supported the Hawaiian endemic/widespread sister species relationships, he did not rule out the possibility that putative endemics comprised isolated populations of their widespread sister species. Furthermore, he could not unambiguously resolve the interrelationships of subgenera within *Plagiotremus*, nor the placement of *Plagiotremus* relative to other genera in the tribe Nemophini. Although morphology-based phylogenetic analyses have provided some taxonomic clarity within this group, molecular approaches may enable the further resolution of uncertain relationships as well as help place endemic species within a broader biogeographic context.

Few studies have included *Plagiotremus* in molecular-phylogenetic analyses, and none have considered species endemic to Hawai&#x02BB;i. *Plagiotremus rhinorhynchos*, *P. tapeinosoma*, *P. azaleus*, and *P. laudandus* are found in several multilocus analyses of blennioids [@linPhylogenyBiogeographyShallow2013; @casewellEvolutionFangsVenom2017; @hundtMultilocusMolecularPhylogeny2014; @hundtExtremeDentitionDoes2018]. Most recently, Smith-Vaniz et al. [-@smith-vanizAggressiveMimicryAspidontus2020] compared cytochrome c oxidase subunit I (COI) sequences from five species of *Plagiotremus* (*P. tapeinosoma*, *P. laudandus*, *P. flavus*, *P. townsendi*, and *P. phenax*) using a neighbor-joining tree, synonymizing *P. flavus* under *P. laudandus*.

Another important question relating to endemic reef fishes in Hawai&#x02BB;i is their geographic origin. Due to the volcanic nature of the Hawaiian archipelago, all species necessarily originated elsewhere in the Pacific. For reef fishes, there are two prevailing hypotheses regarding their colonization of the Hawaiian Islands: i) Hawaiian species originate in the western Pacific, ariving through rare dispersal events via the Kuroshio current [@houriganMidoceanIsolationEvolution1987a]; ii) Hawaiian species originate in the central/southern Pacific, arriving via stepping-stone dispersal through the Line Islands [@goslineInshoreFishFauna1955]. In both cases, Johnston Atoll occupies a potentially key position as the nearest reef habitat to the Hawaiian Archipelago, although its specific role in all cases remains unresolved [@gaitherHighConnectivityDeepwater2011; @dibattistaPhylogeographyPacificBlueline2011; @lobelJohnstonAtollReef2020; @andrewsPhylogeographicAnalysesSubmesophotic2014; @riveraSubtlePopulationGenetic2004] and while many endemic Hawaiian reef fishes are known to occur there [@randallAnnotatedChecklistFishes1985], no records yet exist for either Hawaiian *Plagiotremus* species. Phylogeographic studies have shown support for both hypotheses, with species such as limpets (*Cellana* spp.) [@birdDiversificationSympatricBroadcastspawning2011] and butterflyfishes (*Chaetodon miliaris*, *C. fremblii*) [@craigOriginsAgesPopulation2010] arriving via the western route and combtooth blennies (*Cirripectes vanderbilti*) [@hobanCirripectesMatatakaroNew2020] and *Etelis* spp. snappers [@andrewsPhylogeographicAnalysesSubmesophotic2014] colonizing via the southern route.

Here, we investigate the status of *Plagiotremus* at various levels, combining sequence data from hundreds of new *Plagiotremus* samples with sequences from a public database to create a molecular phylogeny based on a segment of cytochrome oxidase subunit I (COI). With data from nine putative *Plagiotremus* species and three other genera, this study presents the most taxonomically-comprehensive molecular phylogeny of *Plagiotremus* to-date, which allows for comparisons with the original morphological phylogeny. Furthermore, and of particular interest in Hawai&#x02BB;i, this study is the first to present molecular sequence data from the endemic Hawaiian *Plagiotremus* species. If both *P. ewaensis* and *P. goslinei* are truly distinct from their widespread sister species *P. rhinorhynchos* and *P. tapeinosoma*, each should be considerably diverged from one another at the molecular level. Additionally, we use COI and cytochrome B (cytb) to investigate the phylogeography of *Plagiotremus tapeinosoma* and *P. rhinorhynchos* in the context of their relationship with their respective Hawaiian endemic sister species, with the aim of understanding their geographic origins in the Indo-Pacific.

# Materials and Methods {-}

## Specimen collection {-}

We collected `r nrow(original_samples)` individuals of *Plagiotremus tapeinosoma*, *P. rhinorhynchos*, *P. goslinei*, and *P. ewaensis* from Hawai&#x02BB;i and abroad. Specimens were collected by pole spear (captured among the spread-out tines of a three-prong tip) and/or hand net and a portion of the caudal fin was preserved in 70% ethanol or salt-saturated DMSO solution [@seutinPreservationAvianBlood2011] for subsequent molecular analysis. *Plagiotremus rhinorhynchos* and *P. tapeinosoma* were collected across their ranges in the Indo-Pacific, from the Red sea to the Marquesas Islands. *Plagiotremus ewaensis* and *P. goslinei* were collected from 13 locations in the Main (MHI) and Northwestern Hawaiian Islands (NWHI) ([@fig:fig_map[]{}a; @fig:fig_map[]{}b; @tbl:tbl_collection; @tbl:tbl_supp_seqs]). All specimens from the NWHI were collected under permits PMNM-2007-032, PMNM-2008-046, PMNM-2009-032L, PMNM-2009-044, PMNM-2011-025, and PMNM-2012-045, issued by the National Oceanic and Atmospheric Administration, the U.S. Fish and Wildlife Service, and the State of Hawaii Division of Aquatic Resources to BWB. Specimen sampling and handling protocols were approved by the University of Hawai&#x02BB;i at M&#257;noa Institutional Animal Care & Use Committee under approval nos. 09-753 and 15-2271.

## DNA extraction and sequencing {-}

Tissue samples were brought to the Hawai&#x02BB;i Institute of Marine Biology (HIMB) for processing. We extracted genomic DNA using a modified version of the HotSHOT protocol [@meekerMethodIsolationPCRready2007] or the Omega EZNA-96^&reg;^ Tissue DNA Kit (Omega Bio-Tek, Georgia, USA) following manufacturer's protocols. Specimens were amplified for a region of the mitochondrial cytochrome oxidase subunit I (COI) gene using the primers FishF1 (5'--TCAACCAACCACAAAGACATTGGCAC--3') and FishR1 (5'--TAGACTTCTGGGTGGCCAAAGAATCA--3') following the chemistry and thermocycler settings of Ward et al. [-@wardDNABarcodingAustralias2005]. <!--Select specimens of *Plagiotremus goslinei* and *P. tapeinosoma*-->Specimens were also amplified for a region of cytochrome B (cytb) using the custom-designed forward primer BlnF3 (5'--TCAGTHTGATGAAACTTTGGCTC--3') and the reverse primer cyb7 (5'--AATAGGAAGTATCATTCGGGTTTGAT--3') [@taberletUnusualMitochondrialDNA1992]. PCR reactions for cytb were carried out in 15&#x3BC;l volumes consisting of 7.5&#x3BC;l BioMix Red^&trade;^ (Bioline Inc., Massachusetts, USA), 0.3&#x3BC;l of each 10&#x3BC;M primer, 1.5&#x3BC;l BSA (New England Biolabs, Massachusetts, USA), 1&#x3BC;l DNA, and filled to volume with PCR-grade ultrapure water. Reactions were performed under the following conditions: 95&deg;C for 2 m followed by 35 cycles of denaturation at 94&deg;C for 30 s, annealing for 50&deg;C for 45 s, and extension at 72&deg;C for 45 s, with a final extension of 10 min at 72&deg;C. We visualized PCR products for all markers using 1.5--2% agarose gels stained with GelRed^&reg;^ Nucleic Acid Gel Stain (Biotium, California, USA) and purified final products by incubating them with 0.75 units of exonuclease and 0.5 units of shrimp alkaline phosphatase (ExoSAP; USB, Ohio, USA) per PCR product for 30 min at 37&deg;C followed by 15 min at 85&deg;C. DNA sequencing was performed using fluorescently-labeled dideoxy terminators on an ABI 3730XL Genetic Analyzer (Applied Biosystems, California, USA) at the University of Hawai&#x02BB;i Advanced Studies of Genomics, Proteomics and Bioinformatics sequencing facility.

## Additional sequence acquisition {-}

We downloaded all available COI sequence data in the Barcode of Life Database (BoLD) for the genera *Plagiotremus*, *Xiphasia* and *Meiacanthus*. Detailed metadata for external sequences can be found in [@tbl:tbl_supp_bold_seqs]. *Omobranchus* has been identified as an appropriate outgroup to the tribe Nemophini in the multi-locus molecular phylogenies discussed above, so a COI sequence from *Omobranchus anolius* with identification and location information was downloaded from BoLD and used as the outgroup in our phylogenetic analyses. <!--Three *Plagiotremus* sequences pulled from BoLD were excluded in downstream analyses due to unclear origins or mislabeling.-->


## Sequence analysis {-}

### Phylogenetic analyses {-}

All data analyses were performed using `R` 4.4.1 [@rcoreteamLanguageEnvironmentStatistical2024]. Sequence data was read using the `R` packages `Biostrings` [@aboyounBiostringsEfficientManipulation2017] and `ape` [@paradisApeEnvironmentModern2019] and aligned using the Clustal Omega algorithm [@sieversFastScalableGeneration2011] implemented in the package `msa` [@bodenhoferMsaPackageMultiple2015]. Since we used sequences from various sources, their lengths varied and not all loci were present in the final alignment. To compare the effects of missing data on tree node support, three versions of the sequence alignment were generated: one that excluded sites missing in more than one individual (hereafter referred to as the 99% coverage dataset), one that only included sites sequenced in >50% of individuals (>50% coverage dataset), and one with all sequences in the alignment (the full dataset). After filtering, identical sequences were merged<!-- and taxonomic and geographic distribution of merged sequences were recorded-->.

We conducted a Bayesian phylogenetic reconstruction using `MrBayes` v3.2.7 [@ronquistMrBayes32Efficient2012], running three independent searches for 20 million generations each, saving trees every 1,000 generations and discarding the first 10% as burn-in. Since COI is a protein-coding gene, we used `PartitionFinder` 2 with greedy algorithm to select the best model(s) of nucleotide substitution partitioned by codon position, based on corrected Akaike Information Criterion (AICc) [@lanfearPartitionFinder2New2017; @lanfearPartitionFinderCombinedSelection2012; @guindonNewAlgorithmsMethods2010]. We verified model parameter and Markov-chain convergence using `Tracer` v1.7.2 [@rambautPosteriorSummarizationBayesian2018], assuring effective sample sizes &gt;200 for all parameters. Consensus trees were interpreted and plotted in R using the package `ggtree` [@yuGgtreePackageVisualization2017]. As a compromise between potential under- and over-filtering, we used the &gt;50% coverage dataset to examine phylogenetic relationships within *Plagiotremus*.

<!--
### Divergence time estimation {-}

We used `BEAST` v2.6.7 [@bouckaertBEAST2Software2014; @drummondBEASTBayesianEvolutionary2007] to estimate divergence times among *Plagiotremus* species. We used the 99% coverage COI alignment filtered to include four individuals of each *Plagiotremus* species plus *Meiacanthus atrodorsalis*, with *P. tapeinosoma* split into two groups comprising Marquesan individuals and those from across the broader Indo-Pacific. We allowed `BEAST` to iterate over all possible sequence evolution models and choose the model with the highest likelihood given the data, as implemented in the bModelTest package [@bouckaertBModelTestBayesianPhylogenetic2017]. We ran four independent searches for 50 million generations each, saving trees every 1,000 generations and discarding the first 10% as burn-in. We chose a clock rate prior of 1.2% per million years based on previous estimates for marine bony fishes [@berminghamFishBiogeographyMolecular1997; @hunePhylogeographicAnalysisThyrsites2021; @fontesPhylogeographyWidelyDistributed2025], a relaxed uncorrelated lognormal clock, and a yule model with uniform birth rate tree priors. Monophyletic MRCA priors were set on *Plagiotremus* and *Meiacanthus*. MCMC convergence was verified with `Tracer`, logs and trees were combined using `logcombiner` v2.6.7, and final maximum clade credibility trees were generated using `treeannotator` v2.6.3 `BEAST` [@heledLookingTreesForest2013].
-->

### Intraspecific phylogeography {-}

To examine intraspecific relationships within the *Plagiotremus goslinei/tapeinosoma* and *P. ewaensis/rhinorhynchos* species pairs, we constructed haplotype networks using all available sequences from each pair. Sequences were interpreted and aligned using the same procedures as for phylogenetic analyses and we used the longest alignment with data from all individuals (the 99% coverage dataset, as described above). Haplotype frequencies were calculated, arranged into maximum-parsimony networks [@templetonCladisticAnalysisPhenotypic1992], and plotted using the package `pegas` [@paradisPegasPackagePopulation2010]. Sampling locations were grouped by marine biogeographic provinces as defined for reef fishes by Kulbicki et al. [-@kulbickiGlobalBiogeographyReef2013], and further subdivided for visualization into marine ecoregions [@spaldingMarineEcoregionsWorld2007], with the Hawaiian province split into Main and Northwestern Hawaiian Islands and the Marquesas separated from the Polynesian province. We examined genetic differentiation and diversity among biogeographic provinces using the K2P distance [@kimuraSimpleMethodEstimating1980] and analysis of molecular variance (AMOVA), as well as calculating haplotype diversity (h), nucleotide diversity (&#x03C0;), and Tajima's D (as implemented in `pegas`). In addition to overall AMOVA, we calculated pairwise &#x03A6;~st~ using the package `haplotypes` [@aktasHaplotypesManipulatingDNA2025] to examine degree of differentiation among provinces. Statistical significance was calculated using permutation tests (with 9,999 permutations) and *p*-values were adjusted for multiple comparisons according to the modified false discovery rate [B-Y FDR, @benjaminiControlFalseDiscovery2001; @narumBonferroniLessConservative2006].

We used `BEAST` v2.6.7 [@bouckaertBEAST2Software2014; @drummondBEASTBayesianEvolutionary2007] to investicate the geographic origins of the *tapeinosoma* species group. We set biogeographic province (*sensu* Kulbicki) as the factor to reconstruct (with *P. azaleus* individuals assigned to the Tropical East Pacific province) and modeled potential geographic origins with a symmetric discrete-trait analysis model. We allowed `BEAST` to iterate over possible sequence evolution models and choose the one with the highest likelihood given the data, as implemented in the `bModelTest` package [@bouckaertBModelTestBayesianPhylogenetic2017]. We ran four independent searches for 50 million generations each, saving trees every 1,000 generations and discarding the first 10% as burn-in. We used a relaxed uncorrelated lognormal clock model with a rate prior of 1.2% sequence divergence per million years (between lineages) based on previous estimates for COI in marine bony fishes [@berminghamFishBiogeographyMolecular1997; @hunePhylogeographicAnalysisThyrsites2021; @fontesPhylogeographyWidelyDistributed2025], and a yule tree model with uniform birth rate priors. A monophyletic MRCA prior was set on *P. azaleus*. Logs and trees were combined using `LogCombiner` v2.6.7, MCMC convergence was verified with `Tracer` (as for `MrBayes` analyses, ESS &gt; 200), and final maximum clade credibility trees were generated using `TreeAnnotator` v2.6.3 [@heledLookingTreesForest2013].

To further examine variation within the *tapeinosoma* species group, we assembled morphological data from the available literature [@smith-vanizSabertoothedBlenniesTribe1976; @smith-vanizSaberToothedBlenniesTribe1987] for *Plagiotremus azaleus*, *P. goslinei*, and *P. tapeinosoma* across their distributional ranges. We assessed distinctness of meristic character frequency distributions using chi-square tests.


# Results {-}

```{r alignment-summary, dependson="alignment", include=FALSE}
coi_covg <- apply(alignments$coi$c00,2,\(x) (sum(x != 4) / length(x)))
cytb_covg <- apply(alignments$cytb$c00,2,\(x) (sum(x != 4) / length(x)))
sample_seqs <- sample_data %>%
  purrr::map(\(x) x %>%
        count(source) %>%
        deframe() %>%
        as.list()
  )
```

## Overall sequencing results {-}

We generated `r sample_seqs$coi$sequenced` new COI sequences and retrieved `r sample_seqs$coi$bold` sequences from BOLD ([@tbl:tbl_collection;@tbl:tbl_supp_bold_seqs]). The final full COI alignment was `r coi_len$c00` base pairs long (50% coverage: `r coi_len$c50` bp, 99% coverage: `r coi_len$c99` bp) ([@fig:fig_supp_covg]). Despite several iterations of primer design and troubleshooting, we were only able to successfully amplify cytb for *Plagiotremus goslinei* and *P. tapeinosoma*. We generated `r sample_seqs$cytb$sequenced` new cytb sequences, retaining the longest alignment with sites present in all samples (`r cytb_len$c00` bp). New sequences generated for this study have been deposited in GenBank under accession numbers [PX905124](https://www.ncbi.nlm.nih.gov/nuccore/PX905124)--[PX905357](https://www.ncbi.nlm.nih.gov/nuccore/PX905357) (COI) and [PX907028](https://www.ncbi.nlm.nih.gov/nuccore/PX907028)--[PX907188](https://www.ncbi.nlm.nih.gov/nuccore/PX907188) (cytb).

## Phylogenetic reconstruction {-}

```{r include=FALSE, dependson="alignment"}
samples <- sample_data$coi
alignment <- alignments$coi$c50[samples$id,]

dd <- dist.dna(alignment) %>%
  as.matrix()

az_sp <- samples %>%
  filter(species == 'Plagiotremus azaleus') %>%
  pull(id)
gt_sp <- samples %>%
  filter(species %in% c('Plagiotremus tapeinosoma','Plagiotremus goslinei')) %>%
  pull(id)

paz_div <- mean(dd[az_sp,gt_sp])
```

According to `PartitionFinder` 2, the best-fit partitioning scheme for the 50%-coverage COI alignment was GTR+I+&#x0393;, HKY+I, and GTR+&#x0393; for codons 1--3 respectively. Best-fit models for other filtering levels can be found in [@tbl:tbl_supp_models]. All genus-level nodes (*Xiphasia*, *Meiacanthus*, and *Plagiotremus*) in the Bayesian consensus tree had &ge;99% posterior probability based on the 50% coverage and full datasets ([@fig:fig_tree_genera[]{}a; @fig:fig_tree_genera[]{}b]). In the 99% coverage dataset, the split between *Xiphasia* and *Meiacanthus/Plagiotremus* had posterior support of ~90% ([@fig:fig_tree_genera[]{}c]). All three trees placed *Meiacanthus* as sister to *Plagiotremus*.

All *Plagiotremus* species currently recognized as valid (with the exception of the *tapeinosoma* group) resolved as monophyletic with high (&ge;99%) node support ([@fig:fig_tree_plagiotremus]). *Plagiotremus ewaensis* and *P. rhinorhynchos* comprised the sister group to all other species. *Plagiotremus townsendi* and *phenax* fell within distinct clades but *P. laudandus* and *flavus*, their sister group, formed an unresolved polytomy (as mentioned previously, *P. flavus* is considered a junior synonym of *P. laudandus*, [@smith-vanizAggressiveMimicryAspidontus2020; @frickeEschmeyersCatalogFishes2025]). The *tapeinosoma* group, containing *P. azaleus*, *P. tapeinosoma*, and *P. goslinei*, was less clearly-resolved. *Plagiotremus tapeinosoma* was polyphyletic, with samples from the Marquesas forming a clade basal to the group and the remaining samples comprising a polytomy of mixed *P. goslinei* and *tapeinosoma* sequences. *Plagiotremus azaleus* occurred on a long branch within the polytomy, but made up a well-suported monophyletic group diverged by an average of `r pc(paz_div,0.01)` (K2P distance).

## Phylogeography {-}

### *Plagiotremus goslinei* and *P. tapeinosoma* {-}

```{r include=FALSE, dependson="haplotypes"}
hap <- haplotypes$pgt$coi
h <- hap_summary(hap)
top_3 <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(haplotype)
top_3_n <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(n)
max_hap <- h$summary %>%
  arrange(desc(n)) %>%
  dplyr::slice(4:n()) %>%
  pull(n) %>%
  max()


hap1 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[1]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()

hap2 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[2]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()
```
#### COI {-}

There were `r h$n` unique COI haplotypes among the `r nrow(hap$alignment)` sequences generated for the *P. goslinei*/*tapeinosoma* species pair, with `r h$shared` haplotypes that were shared in at least two individuals, `r h$shared_spp` shared across species, and `r h$shared_src` shared between original and BoLD data sources ([@fig:fig_hap_coi_pgt]). There were `r length(seg.sites(hap$alignment))` segregating sites, overall haplotype diversity (h) was `r num(hap.div(hap$alignment),0.001)` and nucleotide diversity (&#x03c0;) was `r num(nuc.div(hap$alignment),0.001)`. Haplotype diversity was highest in the Western Indian Ocean (WIO) and Central Indo-Pacific (CIP) (WIO: h = 0.926; CIP: h = 0.952; [@tbl:tbl_stats_coi_pgt]) and lowest in the Marquesas (h = 0.474). Nucleotide diversity was also highest in the WIO (&#x03c0; = 0.006) and CIP (&#x03c0; = 0.008), but was lowest in the Red Sea (&#x03c0; = 0.001). Tajima's D was negative and significant in Hawai&#x02BB;i, the Marquesas, and the Polynesian province, which may indicate past population expansion or selection in these areas ([@tbl:tbl_stats_coi_pgt]).

```{r include=FALSE, dependson="alignment"}
alignment <- haplotypes$pgt$coi$alignment
samples <- haplotypes$pgt$coi$samples

dd <- dist.dna(alignment) %>%
  as.matrix()

m_sp <- samples %>%
  filter(province == 'Marquesas') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas') %>%
  pull(id)

marq_div_all <- mean(dd[m_sp,ip_sp])

m_sp <- samples %>%
  filter(province == 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)

marq_div_pta <- mean(dd[m_sp,ip_sp])
```

Three main haplotypes represented `r format_list(top_3_n)` individuals across all biogeographic provinces ([@fig:fig_hap_coi_pgt]). The bulk of *P. goslinei* individuals (`r hap1$hi` specimens from both MHI and NWHI) shared a single haplotype with `r hap1$ip` *P. tapeinosoma* specimens from the Central Indo-Pacific, Central Pacific, and Polynesian provinces. The most common *P. tapeinosoma* haplotype represented `r hap2$ip` individuals spanning the Indo-Pacific realm from South Africa and the Red Sea to the Society Islands (every Indo-Pacific province). `r english::Words(hap2$hi)` *P. goslinei* individual`r ifelse(hap2$hi > 1,'s','')` from the Northwestern Hawaiian Islands also shared this haplotype. The third most frequent haplotype almost entirely comprised individuals from the Marquesas, with the exception of one from the Society Islands. Remaining haplotypes represented `r english::words(max_hap)` or fewer individuals each. The two most common haplotypes differed from one another by a single mutation and represented both species. A third haplotype shared across species contained a single individual of each and differed from the main Hawaiian haplotype by one mutation. Two *P. goslinei* individuals were more closely related to the main *P. tapeinosoma* haplotype than to the rest of the Hawaiian haplotypes, and two *P. tapeinosoma* (from the Central Indo-Pacific and Central Pacific) were more closely related to the main *P. goslinei* haplotype. The remaining haplotypes of both species, with the the exception of Marquesan haplotypes, were not more than a few steps diverged from either of the two most common haplotypes. All individuals from the Marquesas either shared or were linked to the third most frequent haplotype, and all were separated from the others of either species by 5--9 mutational steps (mean `r pc(marq_div_all,0.01)` sequence divergence including both species, `r pc(marq_div_pta,0.01)` for *P. tapeinosoma* only). Mean pairwise sequence divergence among provinces can be found in [@tbl:tbl_supp_dist_pgt_coi].

```{r amova-coi-pgt, include=FALSE, dependson="amovas"}
model <- amovas$pgt$coi$amova

# get pairwise PhiST and make it a full matrix
phi <- as.matrix(as.dist(amovas$pgt$coi$pairwise$PhiST))
# set diagonals to NA
diag(phi) <- NA

# get pairwise adjusted p-values and make them a full matrix
# with NA diagonals
pp <- amovas$pgt$coi$pairwise$p
pp[lower.tri(pp)] <- p.adjust(pp[lower.tri(pp)],"BY")
pp <- as.matrix(as.dist(pp))
diag(pp) <- NA
```
COI AMOVA analyses showed significant overall variation among biogeographic provinces (&#x03A6;~st~ = `r round(phist(model),3)`, `r amova_pval(model)`), with multiple significant pairwise comparisons among provinces ([@tbl:tbl_amova_pgt_coi]). The WIO was significantly differentiated from all provinces except the Red Sea (&#x03A6;~st~: `r format_range(range_minus(phi['WIO',],'RS'),3)`), to which it is adjacent. In contrast and despite greater geographic distance, the Red Sea was significantly differentiated only from Hawai&#x02BB;i and the Marquesas (&#x03A6;~st~: `r format_range(phi['RS',c('MHI','NWHI','MQ')],3)`, *p* < 0.001). *Plagiotremus goslinei* (MHI and NWHI) was significantly differentiated from *P. tapeinosoma* (&#x03A6;~st~: `r format_range(c(range_minus(phi['MHI',],'NWHI'),range_minus(phi['NWHI',],'MHI')),3)`, p < 0.001). However, the strongest genetic differentiation was not between putatitve species or the most geographically distant provinces, but between the Marquesas and the rest of the Indo-Pacific (&#x03A6;~st~: `r format_range(phi['MQ',],3)`, *p* < 0.001; [@tbl:tbl_amova_pgt_coi]).

```{r include=FALSE, dependson="haplotypes"}
hap <- haplotypes$pgt$cytb
h <- hap_summary(hap)
top_3 <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(haplotype)
top_3_n <- h$summary %>% arrange(desc(n)) %>% dplyr::slice(1:3) %>% pull(n)
max_hap <- h$summary %>%
  arrange(desc(n)) %>%
  dplyr::slice(4:n()) %>%
  pull(n) %>%
  max()


hap1 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[1]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()

hap2 <- h$details %>%
  mutate(hi = if_else(str_detect(province,'Hawaii'),'hi','ip')) %>%
  count(haplotype,hi) %>%
  filter(haplotype == top_3[2]) %>%
  select(-haplotype) %>%
  deframe() %>%
  as.list()
```
#### cytb {-}

There were `r h$n` unique cytb haplotypes among the `r nrow(hap$alignment)` sequences generated for the *P. goslinei*/*tapeinosoma* species pair, with `r h$shared` haplotypes that were shared in at least two individuals and `r h$shared_spp` shared across species ([@fig:fig_hap_cytb_pgt]). There were `r length(seg.sites(hap$alignment))` segregating sites, overall haplotype diversity (h) was `r num(hap.div(hap$alignment),0.001)` and nucleotide diversity (&#x03c0;) was `r num(nuc.div(hap$alignment),0.001)`. In contrast to COI, there were many low-frequency haplotypes, which resulted in higher haplotype diversity across all provinces. It was highest in the WIO (h = 0.918), CIP (h = 1), and Central Pacific (CP, h = 0.939) and lowest in Hawai&#x02BB;i (mean h = 0.722) ([@tbl:tbl_stats_cytb_pgt]). Nucleotide diversity was highest in the WIO (&#x03c0; = 0.007) and lowest in Hawai&#x02BB;i (mean &#x03c0; = 0.003) and the Red Sea (&#x03c0; = 0.002). Tajima's D was negative and significant in the Red Sea and Northwestern Hawaiian Islands ([@tbl:tbl_stats_cytb_pgt]).

```{r include=FALSE, dependson="alignment"}
alignment <- haplotypes$pgt$cytb$alignment
samples <- haplotypes$pgt$cytb$samples

dd <- dist.dna(alignment) %>%
  as.matrix()

m_sp <- samples %>%
  filter(province == 'Marquesas') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas') %>%
  pull(id)

marq_div_all <- mean(dd[m_sp,ip_sp])

m_sp <- samples %>%
  filter(province == 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)
ip_sp <- samples %>%
  filter(province != 'Marquesas' & species == 'Plagiotremus tapeinosoma') %>%
  pull(id)

marq_div_pta <- mean(dd[m_sp,ip_sp])
```

As in the COI dataset, the plurality of Hawaiian specimens (`r hap1$hi` individuals) shared a single haplotype, which also included `r english::words(hap1$ip)` *P. tapeinosoma* (from the Central Pacific and Polynesian provinces) ([@fig:fig_hap_cytb_pgt]). Unlike COI, *P. tapeinosoma* cytb haplotypes were not shared as broadly across provinces, with the Red Sea, Western Indian Ocean (WIO), and Marquesas largely represented by unique haplotypes (although one haplotype was shared by specimens from the WIO to Polynesia). However, singleton haplotypes from distant provinces were commonly diverged by only one or two mutations. There were no *P. goslinei* individuals represented by primarily *P. tapeinosoma* haplotypes, but two were more closely related to diverged haplotypes from the WIO or the Marquesas than to most other conspecifics. Most cytb haplotypes were only separated by 1--3 mutational steps across both species, but there were two notably divergent haplogroups in the network: one representing Marquesan individuals (as for COI) and one representing 7 from the WIO. The WIO haplogroup was diverged from the main *P. goslinei* haplotype by 5 mutations, with the Marquesan group a minimum of 12--18 steps further (mean `r pc(marq_div_all,0.01)` sequence divergence including both species, `r pc(marq_div_pta,0.01)` for *P. tapeinosoma* only). Mean pairwise sequence divergence among provinces can be found in [@tbl:tbl_supp_dist_pgt_cytb].

```{r amova-cytb-pgt, include=FALSE, dependson="amovas"}
model <- amovas$pgt$cytb$amova

# get pairwise PhiST and make it a full matrix
phi <- as.matrix(as.dist(amovas$pgt$cytb$pairwise$PhiST))
# set diagonals to NA
diag(phi) <- NA

# get pairwise adjusted p-values and make them a full matrix
# with NA diagonals
pp <- amovas$pgt$cytb$pairwise$p
pp[lower.tri(pp)] <- p.adjust(pp[lower.tri(pp)],"BY")
pp <- as.matrix(as.dist(pp))
diag(pp) <- NA
```

In the AMOVA analysis of the cytb alignment, overall variation among biogeographic provinces was significant (&#x03A6;~st~ = `r round(phist(model),3)`, `r amova_pval(model)`). This dataset had more significant pairwise comparisons than COI ([@tbl:tbl_amova_pgt_cytb]), with non-significant differentiation detected only between the CP and PL and MHI and NWHI. However, as in the COI dataset, the highest significant &#x03A6;~st~ values resulted from comparisons with the Marquesas (&#x03A6;~st~: `r format_range(phi['MQ',],3)`, *p* < 0.001), and not across species or between far-flung provinces ([@tbl:tbl_amova_pgt_cytb]).

### *Plagiotremus ewaensis* and *P. rhinorhynchos* {-}

```{r include=FALSE, dependson="haplotypes"}
hap <- haplotypes$per$coi
h <- hap_summary(hap)


r_hap <- h$details %>%
  filter(species == 'Plagiotremus rhinorhynchos') %>%
  count(haplotype) %>%
  slice_max(n)

e_hap <- h$details %>%
  filter(species == 'Plagiotremus ewaensis') %>%
  count(haplotype) %>%
  arrange(desc(n))
  # filter(haplotype == top_3[2]) %>%
  # select(-haplotype) %>%
  # deframe() %>%
  # as.list()
```

```{r include=FALSE, dependson="alignment"}
alignment <- haplotypes$per$coi$alignment
samples <- haplotypes$per$coi$samples

dd <- dist.dna(alignment) %>%
  as.matrix()

r_sp <- samples %>%
  filter(species == 'Plagiotremus rhinorhynchos') %>%
  pull(id)
e_sp <- samples %>%
  filter(species == 'Plagiotremus ewaensis') %>%
  pull(id)

per_div <- mean(dd[r_sp,e_sp])
```
#### COI {-}

There were `r h$n` unique COI haplotypes among the `r nrow(hap$alignment)` sequences generated for the *P. ewaensis*/*rhinorhynchos* species pair (`r length(e_sp)` *P. ewaensis*, `r length(r_sp)` *P. rhinorhynchos*), with `r h$shared` haplotypes that were shared in at least two individuals and `r h$shared_src` shared between original and BoLD data sources ([@fig:fig_hap_coi_per]). The two species shared no haplotypes in common, and were separated by a minimum of 16 mutational steps (mean `r pc(per_div,0.01)` sequence divergence). There were `r length(seg.sites(hap$alignment))` segregating sites, overall haplotype diversity (h) was `r num(hap.div(hap$alignment),0.001)` and nucleotide diversity (&#x03c0;) was `r num(nuc.div(hap$alignment),0.001)`. Within *P. ewaensis*, there were `r length(seg.sites(hap$alignment[e_sp,]))` segregating sites, haplotype diversity was `r num(hap.div(hap$alignment[e_sp,]),0.001)`, and nucleotide diversity was `r num(nuc.div(hap$alignment[e_sp,]),0.001)`. Within *P. rhinorhynchos*, there were `r length(seg.sites(hap$alignment[r_sp,]))` segregating sites, haplotype diversity was `r num(hap.div(hap$alignment[r_sp,]),0.001)`, and nucleotide diversity was `r num(nuc.div(hap$alignment[r_sp,]),0.001)`. For *P. ewaensis*, haplotype diversity was similar in the MHI and NWHI (MHI: h = 0.821; NWHI: h = 0.85) and nucleotide diversity was equal (&#x03c0; = 0.004) ([@tbl:tbl_stats_coi_per]). For *P. rhinorhynchos*, haplotype and nucleotide diversity were highest in the CP (h = 0.909, &#x03c0; = 0.004) and lowest in Red Sea (h = 0, &#x03c0; = 0). Tajima's D was negative and significant only in the CP ([@tbl:tbl_stats_coi_per]). Mean pairwise sequence divergence among provinces can be found in [@tbl:tbl_supp_dist_per_coi].

Most *P. rhinorhynchos* individuals (n = `r r_hap$n`) shared a single haplotype across all biogeographic provinces (except Hawai&#x02BB;i, where the species does not occur) and all but one marine ecoregion. The remaining *P. rhinorhynchos* haplotypes were singletons. Most were separated from the most frequent haplotype by a single step, but there was divergence of up to 4 steps. Within *P. ewaensis*, three haplotypes represented `r format_list(e_hap$n[1:3])` individuals, with the rest having `r e_hap$n[4]` or fewer. All non-singleton *ewaensis* haplotypes spanned both the MHI and NWHI. No Hawai&#x02BB;i haplotype was diverged from another by more than 2 steps.

```{r amova-coi-per, include=FALSE, dependson="amovas"}
model <- amovas$per$coi$amova
prh_model <- amovas$prh$coi$amova
pew_model <- amovas$pew$coi$amova

# get pairwise PhiST and make it a full matrix
phi <- as.matrix(as.dist(amovas$prh$coi$pairwise$PhiST))
# set diagonals to NA
diag(phi) <- NA

# get pairwise adjusted p-values and make them a full matrix
# with NA diagonals
pp <- amovas$prh$coi$pairwise$p
pp[lower.tri(pp)] <- p.adjust(pp[lower.tri(pp)],"BY")
pp <- as.matrix(as.dist(pp))
diag(pp) <- NA

# get missing provinces from pairwise comparisons
pm <- rownames(amovas$pgt$coi$pairwise$PhiST) %>%
  setdiff(rownames(phi)) %>%
  set_names() %>%
  range_minus(c('NWHI','MHI'))
```

There was no significant differentation within either *P. ewaensis* (&#x03A6;~st~ = `r round(phist(pew_model),3)`, `r amova_pval(pew_model)`) or *P. rhinorhynchos* (&#x03A6;~st~ = `r round(phist(prh_model),3)`, `r amova_pval(prh_model)`). No pairwise comparisons within *P. rhinorhynchos* were significant (&#x03A6;~st~: `r format_range(phi,3)`, `r format_range(pp,3,'*p*')`), although due to low sample sizes and/or sampling limitations, we did not include `r format_list(pm,word='or')` in our comparisons ([@tbl:tbl_amova_prh_coi]).

### Ancestral range reconstruction of *tapeinosoma* species group {-}

```{r beast-tree, include=FALSE, dependson='init'}
# read the beast tree and do some tweaking
beast_tree <- read.beast(here('data','trees','tapeinosoma_biogeography.tre')) %>%
  mutate(
    location = str_replace_all(location,'([a-z])([A-Z])','\\1 \\2'),
    location = str_replace_all(location,'Indo Pacific','Indo-Pacific'),
    location.set = purrr::map(location.set,\(l) {
      l <- str_replace_all(l,'([a-z])([A-Z])','\\1 \\2')
      str_replace_all(l,'Indo Pacific','Indo-Pacific')
    }),
    location.set.prob = purrr::map(location.set.prob,\(p) as.numeric(p)),
    location.prob = as.numeric(location.prob)
  )

# just hard-code these since we know what they are
provinces <- c(
  'Red Sea','Western Indian Ocean','Central Indo-Pacific','Central Pacific','Polynesian Province',
  'Northwestern Hawaiian Islands','Main Hawaiian Islands','Marquesas','Tropical East Pacific'
)

# these are just exported subclades that make it easier to figure out root nodes of interest
paz_clade <- read.nexus(here('data','trees','paz_clade.tre'))$tip.label
paz <- getMRCA(beast_tree@phylo,paz_clade)
pta_clade <- read.nexus(here('data','trees','pta_clade.tre'))$tip.label
pta <- getMRCA(beast_tree@phylo,pta_clade)
pgo_clade <- read.nexus(here('data','trees','pgo_clade.tre'))$tip.label
pgo <- getMRCA(beast_tree@phylo,pgo_clade)
mq_clade <- read.nexus(here('data','trees','mq_clade.tre'))$tip.label
mq <- getMRCA(beast_tree@phylo,mq_clade)
rn <- rootnode(beast_tree@phylo)
both <- getMRCA(beast_tree@phylo,c(pta_clade,pgo_clade))

mq_both <- getMRCA(beast_tree@phylo,c(pta_clade,pgo_clade,mq_clade))

node_labels <- c(
  'atop(italic(tapeinosoma),species~group)',
  'italic(P.~azaleus)',
  'atop(paste("\\"",italic(P.~tapeinosoma),"\\""),"(Indo-Pacific)")',
  'atop(paste("\\"",italic(P.~tapeinosoma),"\\""),"(Marquesas)")',
  'paste("\\"",italic(P.~goslinei),"\\"")',
  'italic(P.~goslinei)+italic(tapeinosoma)',
  'atop(italic(P.~goslinei)+italic(tapeinosoma),(incl.~Marquesas))'
)
node_id <- c('root','paz','pta','pta_mq','pgo','pta_pgo','pta_all')
root_map <- tibble(
  node = c(rn,paz,pta,mq,pgo,both,mq_both),
  group = factor(node_labels,levels=node_labels),
  id = node_id
)

clade_nodes <- root_map$node

beast_tree <- beast_tree %>%
  mutate(location = factor(location,levels=provinces))

ancestral_states <- beast_tree %>%
  as_tibble() %>%
  filter(node %in% clade_nodes) %>%
  # filter(!isTip(beast_tree,node)) %>%
  select(node,location.set,location.set.prob) %>%
  mutate(location.set.prob = purrr::map(location.set.prob,as.numeric)) %>%
  # select(node,location,location.prob) %>%
  # mutate(location.prob = as.numeric(location.prob)) %>%
  unnest(-node) %>%
  dplyr::rename(location=2,prob=3) %>%
  mutate(location = factor(location, levels=provinces)) %>%
  # mutate(location = str_replace_all(location,'([a-z])([A-Z])','\\1 \\2')) %>%
  pivot_wider(names_from = 'location', values_from = 'prob', names_expand = TRUE) %>%
  mutate(across(-node,~replace_na(.x,0)))

root_table <- ancestral_states %>%
  left_join(root_map,by='node') %>%
  select(-node) %>%
  select(group,everything())

bst <- beast_tree %>%
  as_tibble() %>%
  filter(node %in% root_map$node)

beast_info <- root_map %>%
  pull(id) %>%
  set_names() %>%
  purrr::map(\(d) {
    nn <- root_map %>%
      filter(id == d) %>%
      pull(node)
    nn <- bst %>%
      filter(node == nn)

    oss <- tibble(og = unlist(nn$location.set), p = as.numeric(unlist(nn$location.set.prob))) %>%
      arrange(desc(p))

    list(
      age = nn %>%
        pull(height) %>%
        as.numeric() %>%
        round(1),
      hpd = nn %>%
        pull(CAheight_0.95_HPD) %>%
        unlist() %>%
        as.numeric() %>%
        round(1),
      origin = nn %>%
        pull(location) %>%
        as.character(),
      origin_prob = nn %>%
        pull(location.prob) %>%
        as.numeric() %>%
        round(3),
      origin_set = oss$og,
      origin_set_prob = oss$p
    )
  })


root_diff <- round(beast_info$root$origin_prob / beast_info$root$origin_set_prob[2],1)
```

The maximum clade credibility tree produced by the `BEAST` analysis matched our Bayesian phylogeny, with the exception that trees output by `BEAST` necessarily produce bifurcating topologies and *Plagiotremus azaleus* was placed as sister to *P. tapeinosoma* and *P. goslinei* (which given their degree of differentiation is likely to better reflect their true relationship). The Marquesan subgroup of *P. tapeinosoma* was also well-supported ([@fig:fig_tree_ancestral_range]). In this analysis, *Plagiotremus azaleus* and *P. tapeinosoma*/*goslinei* shared a common ancestor ca `r beast_info$root$age` million years ago (Ma) (95% HPD: `r format_range(beast_info$root$hpd,rev=TRUE)` Ma), with a `r pc(beast_info$root$origin_prob,0.1)` likelihood of originating in the `r beast_info$root$origin`. The Marquesan subgroup of *P. tapeinosoma* diverged ca `r beast_info$pta_all$age` Ma (95% HPD: `r format_range(beast_info$pta_all$hpd,rev=TRUE)` Ma) in the `r beast_info$pta_all$origin` (`r pc(beast_info$pta_all$origin_prob,0.1)` likelihood), and the remaining individuals of *P. tapeinosoma* and *goslinei*  diverged ca `r beast_info$pta_pgo$age` Ma (95% HPD: `r format_range(beast_info$pta_pgo$hpd,rev=TRUE)` Ma) in the `r beast_info$pta_pgo$origin` (`r pc(beast_info$pta_pgo$origin_prob,0.1)` likelihood) ([@fig:fig_tree_ancestral_range]). *Plagiotremus tapeinosoma* from the Indo-Pacific had a `r pc(beast_info$pta$origin_prob,0.1)` likelihood of originating in the `r beast_info$pta$origin`, *P. goslinei* (and related *P. tapeinosoma* haplotypes) had a `r pc(beast_info$pgo$origin_prob,0.1)` likelihood of originating in the `r beast_info$pgo$origin`, and the Marquesan group had a `r pc(beast_info$pta_mq$origin_prob,0.1)` likelihood of originating in the `r beast_info$pta_mq$origin`. <!--Aside from the origin of the group as a whole-->In general, likelihoods of ancestral geographic origin were low and not strongly skewed toward a specific province. However, the `r beast_info$root$origin` origin of the *tapeinosoma* group as a whole (`r pc(beast_info$root$origin_prob,0.1)`) was `r root_diff` times more likely than the next-most probable province (`r beast_info$root$origin_set[2]`, `r pc(beast_info$root$origin_set_prob[2],0.1)`) ([@fig:fig_supp_origin_prob]).

### Morphological variation of *tapeinosoma* species group {-}

```{r meristics, include=FALSE, dependson='init'}
meristics <- read_csv("data/meristics.csv") %>%
  mutate(across(where(is.numeric),~replace_na(.x,0))) %>%
  select(-ends_with('_mean'),-ends_with('_n')) %>%
  pivot_longer(-c(species,location),names_to = 'measurement',values_to='count') %>%
  separate_wider_delim(cols='measurement',delim='_',names = c('measurement','bin')) %>%
  mutate(species = case_when(
    location == 'Marquesas Is.' ~ 'tapeinosoma_mq',
    .default = species
  )) %>%
  group_by(species,measurement,location,bin) %>%
  group_map(\(data,grp) {
    tibble(species=grp$species,measurement=grp$measurement,location=grp$location,size=as.numeric(rep(grp$bin,data$count)))
  }) %>%
  list_rbind()

x2 <- meristics %>%
  split(.$measurement) %>%
  purrr::map(\(dd) {
    chisq.test(table(dd$size,dd$species))
  })

csq <- function(test,roundr=2) {
  x2 <- round(test$statistic,roundr)
  str_glue("&#x03C7;^2^ = {x2}, {pval(test$p.value)}")
}

mcounts <- meristics %>%
  count(species,measurement) %>%
  group_by(species) %>%
  slice_max(n) %>%
  select(species,n) %>%
  deframe() %>%
  as.list()

```

We recorded morphological characters for `r mcounts$azaleus` *P. azaleus*, `r mcounts$goslinei` *P. goslinei*, `r mcounts$tapeinosoma` Indo-Pacific *P. tapeinosoma*, and `r mcounts$tapeinosoma_mq` Marquesan *P. tapeinosoma* specimens. While detailed counts/measurements for individual specimens could not be retrieved from the published literature, we were able to compile geographic variation in frequencies of meristic characters across the species' distributional ranges. We assembled frequency tables for three meristic characters: anal-fin rays ([@fig:fig_supp_meristics[]{}a]), total dorsal-fin elements ([@fig:fig_supp_meristics[]{}b]), and total vertebrae ([@fig:fig_supp_meristics[]{}c]). Frequencies were significantly different across species/groups for all characters (anal-fin rays: `r csq(x2$anal)`; dorsal-fin elements: `r csq(x2$dorsal)`; vertebrae: `r csq(x2$vert)`). Across all three characters, *P. azaleus* (in the Tropical East Pacific) consistently had the lowest mean counts and Indo-Pacific *P. tapeinosoma* the highest; *Plagiotremus goslinei* and Marquesan *P. tapeinosoma* had mean counts intermediate between the East Pacific and Indo-Pacific widespread species ([@fig:fig_supp_meristics; @fig:fig_supp_meristic_boxes]).

# Discussion {-}

## Phylogenetic relationships of *Plagiotremus* {-}

Our phylogenetic analysis of `r nrow(alignments$coi$c50)` COI sequences from the tribe Nemophini supported the monophyly of most genera and species considered. We found *Xiphasia* as the sister to *Plagiotremus* and *Meiacanthus* ([@fig:fig_tree_genera]). Almost all species within *Plagiotremus*, with the exception of the group comprising *P. azaleus*, *P. tapeinosoma*, and *P. goslinei* (the *tapeinosoma* group *sensu* Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976]), were monophyletic with strong node support.<!-- The remaining *P. tapeinosoma* and *goslinei* specimens, which share haplotypes in both COI and cytb, comprised a large polytomy with no distinct phylogenetic signal of species or geography. *Plagiotremus azaleus* branched from within this polytomy, but formed a deeply-diverged and well-supported clade ([@fig:fig_tree_plagiotremus]). *Plagiotremus tapeinosoma* specimens from the Marquesas formed a clade basal to the *tapeinosoma*/*goslinei* group.-->

Only one previously-published phylogeny has supported *Meiacanthus* as the sister genus to *Plagiotremus* [@linPhylogenyBiogeographyShallow2013]. All subsequent phylogenies found *Xiphasia* to be the sister genus of *Plagiotremus* while the position of *Meiacanthus* has varied [@hundtMultilocusMolecularPhylogeny2014; @casewellEvolutionFangsVenom2017; @hundtExtremeDentitionDoes2018]. These later differences have been attributed by some authors to increased taxonomic coverage <!--within *Meiacanthus* and other Nemophini genera -->[@hundtMultilocusMolecularPhylogeny2014]. While our dataset was limited to a single marker, taxonomic coverage was greater than in previous studies, and this this may have resulted in the different relationships we observed. Relative to previous multi-locus phylogenies, we considered three times more *Plagiotremus* species and included considerable geographic variation by analyzing individuals from as many locations as possible across all genera and species. However, the genus-level relationships we recovered should be accepted with caution, given our reliance on a single mitochondrial marker. Increasing both the genetic and taxonomic resolution should help to further resolve these relationships in future studies.

In attempting to resolve the phylogenetic relationships between species groups within *Plagiotremus*, Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976] remarked that "Rigid application of...cladistic..systematics dictates that the *laudandus* and *tapeinosoma* species groups be accorded a sister-group relationship because they share the greatest number of apomorphic characters...[but]...I believe their occurrence in both groups is a case of parallel evolution." Our molecular analyses support the "rigid cladistic" interpretation, with *laudandus* and *tapeinosoma* as sister groups. The deepest divergence within *Plagiotremus* was between the *rhinorhynchos* species group and all others, rendering the subgenus *Runula* (as defined by Smith-Vaniz) polyphyletic. Other studies have also supported this arrangement. Hundt et al. [-@hundtMultilocusMolecularPhylogeny2014] found *P. tapeinsoma* to be more closely-related to *P. laudandus* than *P. rhinorhynchos* (although these were the only three *Plagiotremus* species considered), and all analyses that have included *P. rhinorhynchos* have placed it on a long branch relative to other species within Nemophini [@linPhylogenyBiogeographyShallow2013; @hundtMultilocusMolecularPhylogeny2014; @hundtExtremeDentitionDoes2018]. These results suggest that the subgenus *Runula* is invalid as circumscribed. We were unable to obtain sequences from the *spilistius* species group, but their inclusion in future studies may provide a more complete picture of the relationships within *Plagiotremus*.

```{r mq-diff, includ=FALSE}
phi <- as.matrix(as.dist(amovas$pgt$coi$pairwise$PhiST))
diag(phi) <- NA
hi <- mean(phi[c('NWHI','MHI'),],na.rm=T)
mq <- mean(phi['MQ',],na.rm=T)

phi_diff <- round(mq / hi,1)

dd <- dna_distances$pgt$coi$matrix
diag(dd) <- NA
hid <- mean(dd[c('NWHI','MHI'),],na.rm=T)
mqd <- mean(dd['MQ',],na.rm=T)

dist_diff <- round(mqd / hid,1)
```
### Status of Hawaiian endemics {-}

Both Hawaiian endemic *Plagiotremus* species were considered by Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976] to potentially warrant only subspecies recognition. Strong support for the monophyly of the Hawaiian endemic *P. ewaensis* and the widespread *P. rhinorhynchos* validates the status of the former. On the other hand, we were unable to resolve *P. goslinei* as phylogenetically distinct from *P. tapeinosoma*. Additionally, while there was genetic differentiation between the two putative species, they shared haplotypes in both COI and cytb and the degree to which *P. goslinei* was differentiated from other populations in the species pair was, on average, half that of the Marquesan population. While shared haplotypes alone may not be sufficient to synonymize the two species, the additional fact that two *P. goslinei* specimens were more closely related to *P. tapeinosoma* and two *P. tapeinsoma* specimens (from the Line Islands and Palau) were more closely related to the main *P. goslinei* haplotype is at least evidence of recent and/or ongoing migration between the two groups. Shared haplotypes and the potential for recent mixing suggest that they may be neither demographically nor evolutionarily distinct. However, *P. goslinei* is morphologically distinct from *P. tapeinosoma* across the characters we examined. Thus, we conclude that there is a distinct species in the Hawaiian Archipelgo, but it is either recently diverged or subject to ongoing exchange with the broader Indo-Pacific. Increased molecular coverage may help resolve this relationship in future studies.

```{r marquesas, include=FALSE, dependson='amovas'}
coi_phi <- as.matrix(as.dist(amovas$pgt$coi$pairwise$PhiST))
cytb_phi <- as.matrix(as.dist(amovas$pgt$cytb$pairwise$PhiST))
diag(coi_phi) <- NA
diag(cytb_phi) <- NA

coi_dist <- dna_distances$pgt$coi$matrix
cytb_dist <- dna_distances$pgt$cytb$matrix
diag(coi_dist) <- NA
diag(cytb_dist) <- NA

phi_mq_coi <- mean(coi_phi['MQ',],na.rm=T)
phi_mq_cytb <- mean(cytb_phi['MQ',],na.rm=T)
dist_mq_coi <- mean(coi_dist['MQ',],na.rm=T)
dist_mq_cytb <- mean(cytb_dist['MQ',],na.rm=T)
```

## Origins and phylogeography of Hawaiian endemics {-}

### *tapeinosoma* species group {-}

Due to the uncertain phylogenetic status of *Plagiotremus goslinei* relative to *P. tapeinosoma*, the *tapeinosoma* species group warrants closer investigation. In contrast to the apparent admixture of most Pacific and Hawaiian members of this group, Marquesan *P. tapeinosoma* specimens occurred within a separate and well-supported clade and were genetically diverged from all other populations (mean COI &#x03A6;~st~: `r round(phi_mq_coi,3)`, K2P: `r pc(dist_mq_coi,0.01)`, mean cytb &#x03A6;~st~: `r round(phi_mq_cytb,3)`, K2P: `r pc(dist_mq_cytb,0.01)`). Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976] noted that *P. tapeinosoma* from the Marquesas had meristic values "at least as low as those of *goslinei*," and we found that three key characters (anal-fin rays, total dorsal-fin elements, and total vertebrae) showed disjunct frequency distributions across the four groups (*azaleus*, *goslinei*, Indo-Pacific *tapeinosoma*, Marquesan *tapeinosoma*). Therefore, further study of how phenotypic and genetic variation is partitioned geographically within the *tapeinosoma* species group (wider Indo-Pacific, Hawai&#x02BB;i, and the Marquesas) is warranted, with particular attention to the Marquesan population. Given its morphological and genetic divergence, it likely warrants elevation to species.

*Plagiotremus azaleus* (in the Eastern Pacific) is deeply diverged from other *tapeinosoma* group lineages. Our `MrBayes` analysis recovered an ambiguous placement within the *goslinei/tapeinosoma* polytomy, but the `BEAST` tree was in agreement with the original assessment by Smith-Vaniz [-@smith-vanizSabertoothedBlenniesTribe1976], placing it basal to *goslinei* and *tapeinosoma*. Given its degree of divergence and geographic distribution, the latter placement is more likely to be correct and the former an artifact of the analysis. Ancestral range reconstruction suggests an East Pacific origin for the *tapeinosoma* group as a whole, with *P. goslinei/tapeinosoma* originating in the Central Pacific. This is supported by both genetic and morphological analyses. While there was genetic structure among biogeographic provinces in *P. tapeinosoma*, the occurrence of a single haplotype across the entire range (except the East Pacific and Marquesas) indicates a highly dispersive life history with rare dispersals across barriers leading to divergence and/or speciation. With regard to morphology, *P. azaleus* had the lowest meristic counts, Indo-Pacific *P. tapeinosoma* the highest, and *P. goslinei* and Marquesan *P. tapeinosoma* intermediate. In combination with population genetic and ancestral range analyses, this presents the possibility of an ancient East Pacific origin, with rare dispersal to the Central Pacific giving rise to the Marquesan, Hawaiian, and Indo-Pacific clades/haplogroups. Due to the isolation of Hawai&#x02BB;i (by distance) and the Marquesas (primarily by the configuration of oceanographic currents), these groups underwent only slight morphological divergence from their ancestral population. In contrast, *P. tapeinosoma* dispersed throughout the rest of the Indo-Pacific and experienced significant increases in meristic character counts.

<!--, but the initial description of *P. goslinei* [@strasburgNotesBlennioidFishes1956] suggested that it was intermediate between *P. azaleus* and *P. tapeinosoma*. The placement of *P. azaleus* as the sister group to *P. tapeinosoma* and *P. goslinei* in this study was not supported by the `MrBayes` tree, as it nested within mixed groups of *P. tapeinosoma* and *P. goslinei*. However, it occurred on a long internal branch (and was supported in the `BEAST` tree). Thus, its placement was likely an artifact of our limited dataset. It is doubtless that these three species form an evolutionary unit. Ancestral range reconstructions suggest they may have originated in the Tropical East Pacific, where *P. azaleus* is endemic. Previous studies have shown evidence for dispersal in both directions across the East Pacific barrier [cite,cite]

and further study of these lineages could determine whether *P. azaleus* originally colonized the Eastern Pacific through Hawai&#x02BB;i or merely shares a common ancestor.  and contribute to the growing literature about whether isolated islands with high levels of endemism export biodiversity to other regions [@bowenOriginsTropicalMarine2013]. On the other hand, both the *P. azaleus* and Marquesan *P. tapeinosoma* haplotypes were linked to the rest of *P. tapeinosoma* through Indian Ocean haplotypes (rather than those from the Pacific which are much closer). This may indicate that the divergence of these lineages dates back to the closure of the Tethys Sea (that divided the tropical Atlantic and Indo-Pacific realms) as observed in many other reef fish groups [@floeterAtlanticReefFish2008].

### *rhinorhynchos* species group {-}

hello sailor

<!--The taxonomy and phylogeography of fangblennies was elucidated in this study by including more species than ever before in a molecular phylogeny of *Plagiotremus*. Although only one of the Hawaiian lineages seems to merit full species status, further study of the phenotypic and genetic variation within the "*tapeinosoma*" sub-group could lead to the description of distinct species in other regions, as well as provide further insight into how endemic marine species evolve in some of the world's most isolated archipelagos.-->

# Figures and Tables {-}

![Endemic Hawaiian *Plagiotremus* species. (a) `r spp$cite('Plagiotremus goslinei')`. (b) `r spp$cite('Plagiotremus ewaensis')`. Photo credit: Keoki Stender ([https://marinelifephotography.com/](https://marinelifephotography.com/)), with permission.](data/images/goslinei-ewaensis.png){#fig:fig_plagiotremus width=75%}

\newpage

```{r fig-map, fig.cap='Map of *Plagiotremus* collection localities across the Indo-Pacific. Background shading indicates biogeographic realm, yellow: Western Indo-Pacific, green: Central Indo-Pacific, red: Eastern Indo-Pacific, orange: Tropical East Pacific. (a) *Plagiotremus goslinei* and *P. tapeinosoma*. (b) *Plagiotremus ewaensis* and *P. rhinorhynchos*.', results="asis", fig.width=8, fig.height=10, dependson="init"}
# define our fake antemeridian
ante_meridian <- 165

# define a long & slim polygon that overlaps the meridian line & set its CRS to match
# that of world
offset <- 180 - ante_meridian
polygon <- st_polygon(x = list(rbind(
  c(-0.0001 - offset, 90),
  c(0 - offset, 90),
  c(0 - offset, -90),
  c(-0.0001 - offset, -90),
  c(-0.0001 - offset, 90)
))) %>%
  st_sfc() %>%
  st_set_crs(4326)


# ocean background
offset <- ante_meridian+180

# our target crs transformation. maybe we did this twice but I'll leave it since it works
target_crs <- st_crs(str_glue("+proj=moll +x_0=0 +y_0=0 +lat_0=0 +lon_0={ante_meridian}"))


# realms we want to show on the map
map_regions <- regions %>%
  filter(str_detect(REALM,fixed("pacific",ignore_case = TRUE)) & !str_detect(REALM,fixed("temperate",ignore_case = TRUE)))

# Convert to polygon geometries, correct geometry errors, project to PCS,
# snap vertices, merge polygons by REALM, project back to original CRS
# this gets rid of the vertical line at 180, at least in theory
map_regions <- suppressWarnings(
  map_regions %>%
    st_cast("POLYGON") %>%
    st_make_valid() %>%
    st_transform(target_crs) %>%
    st_snap(x = ., y = ., tolerance = 165) %>%
    summarise(geometry = st_union(geometry), .by = REALM) %>%
    st_transform(4326) %>%
    st_shift_longitude() %>%
    st_transform(target_crs)
)

# create world map without overlaps and weird artifacts
worldrn <- suppressWarnings(
  ne_countries(scale = "medium", returnclass = "sf") %>%
    st_make_valid() %>%
    st_difference(polygon) %>%
    st_transform(crs = target_crs)
)


# make the ocean background fill
jitter = 0.01
lat <- c(seq(from=-90,to=90,by=0.25))
lon <- c(rep(offset-jitter,length(lat)),rep(offset+jitter,length(lat)))

oceans <- cbind(lon, c(lat, rev(lat))) %>%
  st_linestring() %>%
  st_cast("POLYGON") %>%
  st_sfc(crs=4326) %>%
  st_shift_longitude() %>%
  st_transform(crs=paste0("+proj=moll +lon_0=",ante_meridian," +lat_0=0 +x_0=0 +y_0=0"))

# create point data
mapping_data <- all_samples %>%
  filter(species %in% unlist(species_pairs)) %>%
  filter(source == 'sequenced') %>%
  drop_na(lat,lon) %>%
  st_as_sf(coords=c('lon','lat'),remove=FALSE,crs=4326) %>%
  dplyr::rename(Species=species) %>%
  st_transform(target_crs) %>%
  st_jitter()

# get unique species
spp <- unique(mapping_data$Species)

# make shape scale
shapos <- c(21:24) %>%
  set_names(spp)
# make fill scale
fillers <- paletteer_d('wesanderson::Darjeeling1',length(spp)) %>%
  set_names(spp)

# make two world plots with the different species pairs
plotz <- species_pairs %>%
  purrr::map(\(pair) {
    # get only species in pair
    pd <- mapping_data %>%
      filter(Species %in% pair)

    # make the pllot
    ggplot() +
      geom_sf(data = oceans, fill="lightblue", alpha=0.3) +
      geom_sf(data = worldrn, aes(group = admin), fill="lightgrey") +
      geom_sf(data = map_regions, aes(fill = REALM),show.legend = FALSE, alpha = 0.3, color="grey77") +
      scale_fill_paletteer_d('waRhol::skull_76_1',guide="none") +
      new_scale_fill() +
      geom_sf(data = pd,  aes(fill=Species, shape=Species), size=2, color="black" ) +
      scale_fill_manual(values = fillers, guide = guide_legend(title=NULL)) +
      scale_shape_manual(values=shapos, guide = guide_legend(title=NULL)) +
      theme_bw() +
      theme(
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom",
        legend.key = element_blank(),
        legend.text = element_text(size=12,face='italic')
      )
  })

# reduce the plots with `/` so they're stacked on top of one another
plotz %>%
  purrr::reduce(`/`) +
  plot_annotation(tag_levels = 'a') &
  theme(plot.tag = element_text(face='bold'))
  # theme(plot.tag = element_text(face='bold',margin = margin(r=4,b=4)))
cat("{#fig:fig_map}")
```

\newpage

```{r trees, results="asis", dependson="alignment"}
# read complete COI tree and join sample data
tree_covg <- names(alignments$coi)

trees <- tree_covg %>%
  sort() %>%
  set_names() %>%
  purrr::map(\(covg) {
    tree <- read.mrbayes(here('data','trees',str_glue('coi_{covg}.nex.con.tre'))) %>%
      left_join(all_samples,by=c(label='id'))

    # drop all but one Omobranchus tip to simplify things
    to_drop <- tree %>%
      as_tibble() %>%
      filter(
        genus == 'Omobranchus',
        label != 'Oan-AMS266-08'
      ) %>%
      pull(label)

    tree <- tree %>%
      drop.tip(to_drop)

    # figure out our one omobranchus outgroup node
    outgroup <- tree %>%
      as_tibble() %>%
      filter(genus == 'Omobranchus') %>%
      pull(node)
    if (length(outgroup) > 1) {
      outgroup <- getMRCA(tree@phylo,outgroup)
    }

    # reroot the tree on the omobranchus node
    tree@phylo <- reroot(
      tree@phylo,
      outgroup,
      position = tree@phylo$edge.length[which(tree@phylo$edge[,2]==outgroup)]/2
    )

    # probs get set to NA when tree is rerooted
    # here we're doing essentially what figtree does
    tree <- tree %>%
      mutate(
        prob = case_when(
          !isTip & is.na(prob) ~ "1",
          .default = prob
        ),
        prob_cut = cut(as.numeric(prob),c(-Inf,0.9,0.99,Inf),labels=c('<0.9','≥0.9','≥0.99'))
      )

    # create groups to color the tree sections
    genus_groups <- tree %>%
      as_tibble() %>%
      filter(genus != 'Omobranchus') %>%
      split(.$genus) %>%
      imap(\(grp,genus) {
        unique(grp$label)
      })
    tree <- tree %>%
      mutate(lbl = case_when(
        genus == 'Omobranchus' ~ 'Omobranchus',
        .default = ''
      )) %>%
      groupOTU(genus_groups,'Genus')
    # make the groups be a factor
    attr(tree@phylo,'Genus') <- factor(attr(tree@phylo,'Genus'),levels=names(genus_groups))

    tree
  })
```

```{r fig-tree-genera, fig.cap='COI Bayesian consensus trees for *Plagiotremus* and closely-related genera. (a) Full tree based on 50% coverage alignment. (b) Genus-level tree based on full alignment (all sequences). (c) Genus-level tree based on 99% coverage alignment.', results="asis", fig.width=8, fig.height=5, dependson="trees"}
# full tree
ft <- 'c50'
# subtrees
st <- grep('c50',names(trees),invert = TRUE,value = TRUE)

# calculate MRCA of genera for clade labeling, excluding Omobranchus
genera <- trees %>%
  purrr::map(\(tree) {
    tree %>%
      as_tibble() %>%
      filter(genus != 'Omobranchus') %>%
      split(.$genus) %>%
      imap(\(grp,genus) {
        if (nrow(grp) > 1) {
          node <- getMRCA(tree@phylo,grp$node)
        } else {
          node <- grp$node
        }
        tibble(id=node,text=genus)
      }) %>%
      list_rbind()
  })

onodes <- trees %>%
  purrr::map(\(tree) {
    tree %>%
      as_tibble() %>%
      filter(genus == 'Omobranchus') %>%
      pull(node)
  })

# gcol <- c(paletteer_d('lisa::MaxErnst',3),'#000000')
# gcol <- c(paletteer_d('lisa::M_C_Escher',3),'#000000')
gcol <- c(set_names(paletteer_d('lisa::WassilyKandinsky_1',3),genera$c00$text),'0'='#000000')

gnodes <- genera[[ft]] %>%
  pull(id) %>%
  combn(2) %>%
  apply(MARGIN=2,FUN=\(nodes) {
    getMRCA(trees[[ft]]@phylo,nodes)
  }) %>%
  unique()

pointcol <- c('≥0.99' = 'black','≥0.9' = 'darkgrey')

# plot the tree
full_tree <- ggtree(trees %>% pluck(ft),aes(color=Genus),size=0.7) +
  geom_cladelab(mapping=aes(node=id,label=text,color=Genus),data=genera %>% pluck(ft), align=FALSE, fontface = "bold.italic", offset = 0.005)  +
  geom_cladelabel(onodes %>% pluck(ft),'bolditalic(Omobranchus)', offset = 0.005, parse=TRUE, size=3) +
  geom_point2(aes(fill=prob_cut,subset = node %in% gnodes),shape=21,size=2,show.legend = TRUE) +
  scale_y_reverse() +
  geom_rootedge(0.005) +
  scale_color_manual(values=gcol,guide='none') +
  scale_fill_manual(values=pointcol,drop=FALSE,name='Node support\n(posterior probability)') +
  theme(
    legend.position = 'inside',
    legend.justification = 'left',
    legend.position.inside = c(0,0.1),
    legend.background = element_blank()
  ) +
  xlim(0,0.55)


subtrees <- trees[st] %>%
  imap(\(tree,covg) {
    gg <- genera %>%
      pluck(covg)

    gnodes <- gg %>%
      pull(id) %>%
      combn(2) %>%
      apply(MARGIN=2,FUN=\(nodes) {
        getMRCA(tree@phylo,nodes)
      }) %>%
      unique()

    gg$id %>%
      purrr::reduce(\(g,node) g %>% collapse(node=node),.init=ggtree(tree,aes(color=Genus))) +
      geom_cladelab(mapping=aes(node=id,label=text,color=Genus),data=gg, align=FALSE, fontface = "bold.italic", offset = 0.001)  +
      geom_point2(aes(fill=prob_cut,subset = node %in% gnodes),shape=21,size=2) +
      geom_cladelabel(onodes %>% pluck(covg),'bolditalic(Omobranchus)', parse=TRUE) +
      geom_point2(aes(subset = (node %in% gg$id)),shape=18,size=3) +
      geom_rootedge(0.005) +
      xlim(NA,0.3) +
      scale_color_manual(values=gcol,guide='none')  +
      scale_fill_manual(values=pointcol,guide='none') +
      scale_y_reverse() +
      xlim(0,0.28)
  })

# plot layout
layout <- '
  11122
  11133
'
# build the plot
c(full_tree,subtrees) %>%
  purrr::reduce(`+`) +
  plot_layout(design=layout) &
  plot_annotation(tag_levels = 'a') &
  theme(plot.tag = element_text(face='bold'))

cat("{#fig:fig_tree_genera}")
```

\newpage

```{r fig-tree-plagiotremus, fig.cap='COI Bayesian consensus tree of relationships within *Plagiotremus* based on 50% coverage alignment. Dashed magenta edges indicate individuals collected in Hawai&#x02BB;i. Node labels indicate Bayesian posterior probability.', results="asis", fig.width=9, fig.height=7, dependson="trees"}
# drop non-Plagiotremus tips
to_drop <- trees$c50 %>%
  as_tibble() %>%
  filter(genus != 'Plagiotremus') %>%
  pull(label)

# filter the tree
tree <- trees$c50 %>%
  drop.tip(to_drop)
# save the tip labels
tips <- tree@phylo$tip.label

hap <- haplotypes %>%
  pluck('pgt','coi')

# get haplotypes
# generate haplotype index map
hindex <- attr(hap$haplotypes,'index') %>%
  imap(\(idx,h) {
    tibble(
      haplotype = h,
      sequence = rownames(hap$alignment)[idx]
    )
  }) %>%
  list_rbind()

# join metadata to tree
# and try to define some geographic range info
tree <- tree %>%
  left_join(hindex,by=c(label='sequence')) %>%
  left_join(hap$samples,by=c(label='id'))

# get tips with marquesan haplotypes
marquesan_tips <- tree %>%
  as_tibble() %>%
  group_by(haplotype) %>%
  filter(province == 'Marquesas') %>%
  ungroup() %>%
  pull(label)

# get root node and all children for marquesan haplotypes
marquesan_node <- getMRCA(tree@phylo,marquesan_tips)
marquesas <- ggtree:::getSubtree(tree@phylo,marquesan_node)

# make nice-looking clades we can label
clades <- tree %>%
  as_tibble() %>%
  drop_na(label) %>%
  mutate(
    grp = case_when(
      str_detect(species,'rhinorhynchos') ~ 'P. rhinorhynchos',
      str_detect(species,'ewaensis') ~ 'P. ewaensis',
      str_detect(species,'ewaensis') ~ 'P. ewaensis',
      str_detect(species,'flavus|laudandus') ~ 'P. laudandus/flavus',
      str_detect(species,'townsendi') ~ 'P. townsendi',
      str_detect(species,'phenax') ~ 'P. phenax',
      str_detect(species,'azaleus') ~ 'P. azaleus',
      str_detect(species,'tapeinosoma|goslinei') ~ 'P. tapeinosoma/goslinei',
      .default = ''
    )
  ) %>%
  mutate(
    grp = case_when(
      node %in% marquesas & !is.na(label) ~ 'P. cf. tapeinosoma\n(Marquesas)',
      .default = grp
    ),
  ) %>%
  split(.$grp) %>%
  imap(\(grp,clade) {
    if (nrow(grp) > 1) {
      node <- getMRCA(tree@phylo,grp$node)
    } else {
      node <- grp$node
    }
    tibble(id=node,text=clade)
  }) %>%
  list_rbind() %>%
  mutate(
    offset = case_when(
      str_detect(text,'azaleus') ~ -0.001,
      str_detect(text,'goslinei') ~ -0.08,
      .default = 0
    )
  )

# get the rode node, because we're gonna set its probability to 1
rootnode <- getMRCA(tree@phylo,tree@phylo$tip.label)

# update some of the internal data for the tree
tree <- tree %>%
  mutate(
    prob = case_when(
      node == rootnode ~ 1,
      .default = as.numeric(prob)
    ),
    posterior = cut(
      prob,
      c(-Inf,0.60,0.75,0.999,Inf),
      labels=c("<0.60","0.60-0.75","0.75-0.95","1")
    )
  )

# figure out which nodes to put points on
nodepoints <- clades %>%
  pull(id) %>%
  combn(2) %>%
  apply(MARGIN=2,FUN=\(nodes) {
    getMRCA(tree@phylo,nodes)
  }) %>%
  unique()

# make colors for our node points
clr_map <- c("<0.60"="white","0.60-0.75"="gray75","0.75-0.95"="gray40","1"="black")

tree <- groupOTU(tree, tree@phylo$tip.label[grep("Pew|Pgo", tree@phylo$tip.label, invert = TRUE)])

ggtree(tree,aes(color=group,linetype=group)) +
  scale_y_reverse() +
  scale_color_manual(values = c('#C82B84','black'),guide='none') +
  scale_linetype_manual(values=c(2,1), guide='none') +
  geom_cladelab(mapping=aes(node=id,label=text,offset=offset),data=clades, parse=TRUE,fontface='bold.italic', barsize=1) +
  geom_rootedge(0.005) +
  geom_nodelab(aes(subset=(node %in% nodepoints),label=round(prob,2)),hjust=0,geom='text',fontface='bold',size=3) +#nudge_x = -0.003, nudge_y=1.1) +
  theme(legend.position = c(0.1,0.25)) +
  xlim(NA,0.25) +
  geom_treescale(y=-135,x=0,offset=1)
cat("{#fig:fig_tree_plagiotremus}")
```

\newpage

```{r fig-hap-coi-pgt, fig.cap='COI haplotype network of *Plagiotremus tapeinosoma* and *P. goslinei*. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.', results="asis", fig.width=16, fig.height=10, dependson="haplotypes"}
haplotypes$pgt$coi$plot
cat("{#fig:fig_hap_coi_pgt}")
```

\newpage

```{r fig-hap-cytb-pgt, fig.cap='cytb haplotype network of *Plagiotremus tapeinosoma* and *P. goslinei*. Abbreviations as in [@fig:fig_hap_coi_pgt].', results="asis", fig.width=16, fig.height=10, dependson="haplotypes"}
haplotypes$pgt$cytb$plot
cat("{#fig:fig_hap_cytb_pgt}")
```

\newpage

```{r fig-hap-coi-per, fig.cap='COI haplotype network of *Plagiotremus rhinorhynchos* and *P. ewaensis*. Abbreviations as in [@fig:fig_hap_coi_pgt].', results='asis', fig.width=16, fig.height=10, dependson='haplotypes'}
haplotypes$per$coi$plot
cat("{#fig:fig_hap_coi_per}")
```

\newpage

```{r fig-tree-ancestral-range, fig.cap='Maximum clade credibility tree showing ancestral range reconstructions for the *tapeinosoma* species group. Node pies indicate likelihood of reconstructed origins. Annotated nodes show most likely origin and likelihood.', results='asis', fig.width=10, fig.height=8, dependson=c('beast-tree','haplotypes')}
# figure out our color palette
origins <- tibble(
  a = provinces,
  b = provinces
) %>%
  mutate(a = if_else(str_detect(a,'Hawaii'),'Hawaii',a)) %>%
  group_by(a) %>%
  mutate( color = paletteer_d( palette_map[as.character(cur_group()$a)],direction=-1)[1:n()], n = n() ) %>%
  ungroup()
beast_pal <- origins %>%
  select(b,color) %>%
  deframe()

# create ancestral range pie subplots and color them appropriately
pies <- nodepie(ancestral_states,cols=2:10)
pies <- lapply(pies,\(x) x + scale_fill_manual(values=beast_pal))

# update tree with pretty labels
beast_tree <- beast_tree %>%
  mutate(
    label = case_when(
      !is.na(label) ~ location,
      .default = label
    ),
    loc = map2_chr(location,location.prob,\(l,p) str_glue('{province_abbrev[as.character(l)]}\n({pc(round(p,3),0.1)})')),
    height = as.numeric(height)
  )

# get color for root edge (which might not work)
rc <- beast_tree %>%
  as_tibble() %>%
  filter(node == rn) %>%
  pull(location)
rc <- beast_pal[rc]

# get legend labels
legend_labs <- beast_tree %>%
  pull(location) %>%
  levels()
legend_labs <- str_glue('{legend_labs} ({province_abbrev[legend_labs]})')

# plot tree in reverse timescale
plotz <- revts(ggtree(beast_tree,aes(color = location), key_glyph='rect',size=1)) +
  scale_color_manual(values=beast_pal, labels=legend_labs,name='Province') +
  geom_inset(pies,width = 0.05, height = 0.05) +
  geom_nodelab(
    aes(subset = (node %in% clade_nodes), label = loc),
    color='black',
    nudge_x = 0.07,
    hjust = 0
  )

# get subset of nodes for which to show pies
rm <- root_map %>%
  filter(id %in% c('pta','pgo','paz','pta_mq'))

# we have to do this because geom_cladelab doesn't support parsing plotmath for some reason
# label relevant clades
reduce2(
  rm$node,rm$group,
  \(p,x,y) p + geom_cladelabel(node=x,label=y,parse=TRUE,offset=0.1,offset.text = 0.1),
  .init = plotz
) +
  theme_tree2() +
  theme(
    legend.position = 'inside',
    legend.position.inside = c(0.15,0.85)
  ) +
  scale_x_continuous(breaks = pretty(c(-beast_tree@data$height,0),n = 5), guide = guide_axis(cap = 'both'),limits = c(-5.2,1), labels = abs) +
  xlab('Ma before present')

cat("{#fig:fig_tree_ancestral_range}")
```

\newpage

```{r tbl-collection,results="asis",dependson="init"}
tbl_collection <- original_samples %>%
  left_join(plagiotremus_data,by="id",suffix=c("","_sequenced")) %>%
  group_by(species) %>%
  summarise(
    Range = ifelse(cur_group()$species %in% hawaii_species,'Hawai&#x02BB;i','Indo-Pacific'),
    n = n(),
    `COI sequences` = sum(sequenced_coi),
    `cytb sequences` = sum(sequenced_cytb)
  ) %>%
  mutate(species = str_glue("*{species}*")) %>%
  dplyr::rename(Species = species)
print_table(tbl_collection,"tbl_collection","Specimen collection summary.",hdr_case=NA,accuracy=1)
```

\newpage

```{r tbl-stats-coi-pgt, results="asis", dependson="haplotypes"}
hap <- haplotypes$pgt$coi

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `&#x03c0;` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    pval = tajima.test(hap$alignment[id,])$Pval.beta,
    sig = case_when(
      pval < 0.001 ~ 3,
      pval < 0.01 ~ 2,
      pval < 0.05 ~ 1,
      .default = 0
    ),
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate( sig = map_chr(sig,\(s) str_c(rep('\\*',s),collapse="")) ) %>%
  mutate(
    D = case_when(
      pval < 0.05 ~ str_glue('**{round(D,3)}^{sig}^**'),
      is.nan(D) ~ NA,
      .default = as.character(round(D,3))
    )
  ) %>%
  arrange(llon) %>%
  select(-c(sig,pval,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_coi_pgt","COI population genetic summary statistics for *P. goslinei*/*tapeinosoma* across marine biogeographic provinces (n = number of samples, S = number of segregating sites, H = number of unique haplotypes, h = haplotype diversity, &#x03c0; = nucleotide diversity, D = Tajima's D; results in bold are statistically significant, \\*: *p* &lt; 0.05, \\*\\*: *p* &lt; 0.01, \\*\\*\\*: *p* &lt; 0.001).",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r amova-tbl, include=FALSE}
amova_tbl <- function(mova,roundr=3,adjust=FALSE, alpha=0.05, upper_p = FALSE) {
  phi <- mova$PhiST
  pval <- mova$p

  if (adjust != FALSE & adjust %in% p.adjust.methods) {
    pval[lower.tri(pval)] <- p.adjust(pval[lower.tri(pval)],adjust)
  }

  if (upper_p) {
    phi[upper.tri(phi)] <- t(pval)[upper.tri(t(pval))]
  }

  sig_idx <- !is.na(pval) & pval < alpha
  sig_idx <- sig_idx * 1
  sig_idx[pval < 0.01] <- 2
  sig_idx[pval < 0.001] <- 3

  phi <- round(phi,roundr)

  sig_n <- sig_idx[sig_idx > 0]
  sig_phi <- phi[sig_idx > 0]
  sig_rep <- map_chr(sig_n,\(n) str_c(rep("\\*",n),collapse=""))

  mode(phi) <- 'character'
  phi[sig_idx > 0] <- str_glue('**{sig_phi}^{sig_rep}^**')
  as.data.frame(phi)
}
```

```{r tbl-amova-pgt-coi, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$pgt$coi$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_pgt_coi","Pairwise COI &#x03A6;~st~ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where n &gt; 5) for *Plagiotremus goslinei*/*P. tapeinosoma*. Significant comparisons appear in bold. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-stats-cytb-pgt, results="asis", dependson="haplotypes"}
hap <- haplotypes$pgt$cytb

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `&#x03c0;` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    pval = tajima.test(hap$alignment[id,])$Pval.beta,
    sig = case_when(
      pval < 0.001 ~ 3,
      pval < 0.01 ~ 2,
      pval < 0.05 ~ 1,
      .default = 0
    ),
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate( sig = map_chr(sig,\(s) str_c(rep('\\*',s),collapse="")) ) %>%
  mutate(
    D = case_when(
      pval < 0.05 ~ str_glue('**{round(D,3)}^{sig}^**'),
      is.nan(D) ~ NA,
      .default = as.character(round(D,3))
    )
  ) %>%
  arrange(llon) %>%
  select(-c(sig,pval,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_cytb_pgt","cytb population genetic summary statistics for *P. goslinei*/*tapeinosoma* across marine biogeographic provinces. Column names as in [@tbl:tbl_stats_coi_pgt].",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-amova-pgt-cytb, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$pgt$cytb$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_pgt_cytb","Pairwise cytb &#x03A6;~st~ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where n &gt; 5) for *Plagiotremus goslinei*/*P. tapeinosoma*. Significant comparisons appear in bold.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-stats-coi-per, results="asis", dependson="haplotypes"}
hap <- haplotypes$per$coi

hap_stats <- hap$samples %>%
  group_by(province) %>%
  summarise(
    n = n(),
    S = length(seg.sites(hap$alignment[id,])),
    H = nrow(pegas::haplotype(hap$alignment[id,])),
    h = hap.div(hap$alignment[id,]),
    `&#x03c0;` = nuc.div(hap$alignment[id,]),
    D = tajima.test(hap$alignment[id,])$D,
    pval = tajima.test(hap$alignment[id,])$Pval.beta,
    sig = case_when(
      pval < 0.001 ~ 3,
      pval < 0.01 ~ 2,
      pval < 0.05 ~ 1,
      .default = 0
    ),
    llon = mean(llon,na.rm=T)
  ) %>%
  ungroup() %>%
  mutate( sig = map_chr(sig,\(s) str_c(rep('\\*',s),collapse="")) ) %>%
  mutate(
    D = case_when(
      pval < 0.05 ~ str_glue('**{round(D,3)}^{sig}^**'),
      is.nan(D) ~ NA,
      .default = as.character(round(D,3))
    )
  ) %>%
  arrange(llon) %>%
  select(-c(sig,pval,llon)) %>%
  dplyr::rename(Province=province)

print_table(hap_stats,id="tbl_stats_coi_per","COI population genetic summary statistics for *P. ewaensis*/*rhinorhynchos* across marine biogeographic province. Column names as in [@tbl:tbl_stats_coi_pgt].",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-amova-prh-coi, results="asis", dependson=c("amovas","amova-tbl")}
print_table(amova_tbl(amovas$prh$coi$pairwise,adjust="BY",upper_p=TRUE),id="tbl_amova_prh_coi","Pairwise COI &#x03A6;~st~ (below diagonal) and associated *p*-values (above diagonal) among biogeographic provinces (where n &gt; 5) for *Plagiotremus rhinorhynchos*. Significant comparisons appear in bold.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

# References {-}

<!-- show main bibliography -->
<div class="multi-refs"></div>

\newpage

# Supplemental material {label="S"}

```{r fig-supp-covg, fig.cap='COI alignment coverage by dataset filtering level.', results="asis", fig.width=6, fig.height=7, results="asis", dependson="alignment"}
covg_data <- alignments %>%
  pluck('coi') %>%
  imap(\(alignment,ds) {
    covg <- apply(alignment,2,\(bases) sum(bases != 4) / length(bases))
    tibble(pos = seq_along(covg), coverage = covg, dataset = ds)
  }) %>%
  list_rbind() %>%
  mutate(dataset = fct_reorder(dataset,-parse_number(dataset)))

verts <- covg_data %>%
  group_by(dataset) %>%
  summarise(min = min(pos), max = max(pos), first_cov=dplyr::first(coverage), last_cov=dplyr::last(coverage))
ggplot(covg_data) +
  geom_area(aes(x = pos, y = coverage, fill=dataset),alpha=0.7, stat="identity",position="identity") +
  geom_line(aes(x = pos, y = coverage, color = dataset)) +
  geom_segment(aes(x=min,xend=min,y=0,yend=first_cov,color=dataset),data=verts) +
  geom_segment(aes(x=max,xend=max,y=0,yend=last_cov,color=dataset),data=verts)  +
  facet_wrap(~dataset,ncol=1) +
  labs(x='Nucleotide position',y='Coverage (%)') +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    panel.grid = element_line(color="#f0f0f0")
  ) +
  guides(color='none',fill = guide_legend(override.aes = list(alpha=1))) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_paletteer_d('tvthemes::Pearl',direction = -1,labels=c('99%','50%','Full dataset'),name='Filtering') +
  scale_color_paletteer_d('tvthemes::Pearl',direction = -1,labels=c('99%','50%','Full dataset'))
cat("{#fig:fig_supp_covg}")
```

\newpage

```{r fig-supp-origin-prob, fig.cap='Likelihood of ancestral geographic origin by *tapeinosoma* species-group clade.', results='asis', fig.width=9, fig.height=7, dependson='beast-tree'}
ggplot(root_table %>%
  pivot_longer(-c(group,id),names_to = 'origin',values_to = 'prob') %>%
  mutate(origin = factor(origin,levels=provinces))) +
  geom_col(aes(x=group,y=prob,fill=origin),position='dodge') +
  scale_fill_manual(values=beast_pal,name='Origin') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45,hjust=1)) +
  labs(x='Clade',y='Likelihood') +
  scale_x_discrete(labels = \(x) parse(text=x)) +
  scale_y_continuous(labels = scales::percent)
cat("{#fig:fig_supp_origin_prob}")
```

\newpage

```{r fig-supp-meristics, fig.cap='Frequency distributions of select meristic characters in the *tapeinosoma* species group. (a) Anal-fin ray count. (b) Total dorsal-fin elements. (c) Total vertebrae.', results="asis", dependson="meristics", fig.height=4, fig.width=8}
spp_map <- c(
  'azaleus' = str_glue('italic(P.~azaleus)~(n=={mcounts$azaleus})'),
  'goslinei' = str_glue('italic(P.~goslinei)~(n=={mcounts$goslinei})'),
  'tapeinosoma' = str_glue('italic(P.~tapeinosoma)~(Indo-Pacific)~(n=={mcounts$tapeinosoma})'),
  'tapeinosoma_mq' = str_glue('italic(P.~tapeinosoma)~(Marquesas)~(n=={mcounts$tapeinosoma_mq})')
)

meristics %>%
  mutate(species = spp_map[species]) %>%
  split(.$measurement) %>%
  purrr::imap(\(vv,mm) {
    x <- str_replace(csq(x2[[mm]]),', ','<br>')
    ggplot(vv) +
      geom_density(aes(x=size,color=species),adjust=2,key_glyph='rect',linewidth = 1) +
      theme_bw() +
      theme(panel.grid = element_blank()) +
      annotation_custom(gridtext::richtext_grob(x,0.7,0.6,gp = gpar(fontsize=8),hjust = 0)) +
      guides(color = guide_legend(nrow=2,byrow=TRUE))
  }) %>%
  purrr::reduce(`+`) +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides = 'collect',axis_titles = "collect") &
  labs(x='Count',y='Frequency') &
  # scale_color_paletteer_d('ltc::fernande', name='Species',labels=scales::label_parse())
  scale_color_paletteer_d('wesanderson::Darjeeling1', name='Species',labels=scales::label_parse(),direction = -1) &
  theme(legend.position = 'bottom')
cat("{#fig:fig_supp_meristics}")
```

\newpage

```{r fig-supp-meristic-boxes, fig.cap='Box plots of meristic values by species. (a) Anal-fin ray count. (b) Total dorsal-fin elements. (c) Total vertebrae.', results="asis", dependson="meristics", fig.height=4, fig.width=8}
spp_map <- c(
  'azaleus' = str_glue('italic(P.~azaleus)'),
  'goslinei' = str_glue('italic(P.~goslinei)'),
  'tapeinosoma' = str_glue('italic(P.~tapeinosoma)~(Indo~Pacific)'),
  'tapeinosoma_mq' = str_glue('italic(P.~tapeinosoma)~(Marquesas)')
)

meristics %>%
  mutate(species = spp_map[species]) %>%
  split(.$measurement) %>%
  purrr::map(\(mm) {
    plotz <- ggplot(mm) +
      geom_boxplot(aes(species,size,fill=species)) +
      labs(x='',y='Count') +
      theme_bw() +
      theme(
        # axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.x = element_blank(),
        panel.grid = element_blank(),
      ) +
      guides(fill = guide_legend(nlabels=scales::label_parse(),col=2))
    # get categories without visible boxes
    pp <- ggplot_build(plotz)@data[[1]] %>%
      filter(ymin == ymax)
    if (nrow(pp) > 0) {
      plotz <- plotz +
        geom_segment(data=pp,aes(x=xmin,xend=xmax,y=middle,yend=middle, color=fill), size=1) +
        guides(color = guide_none())
    }
    plotz
  }) %>%
  purrr::reduce(`+`) +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides = 'collect',axis_titles = "collect") &
  scale_fill_paletteer_d('wesanderson::Darjeeling1', name='Species',labels=scales::label_parse(),direction = -1) &
  theme(legend.position = 'bottom')
cat("{#fig:fig_supp_meristic_boxes}")
```

\newpage

```{r tbl-supp-seqs,results="asis",dependson="haplotypes"}
seq_tbl <- plagiotremus_data %>%
  filter(source == 'sequenced') %>%
  mutate(species = str_glue('*{species}*')) %>%
  arrange(species,id) %>%
  mutate(
    across(
      contains('accession'),
      ~(\(a) str_glue("[{a}](https://www.ncbi.nlm.nih.gov/nuccore/{a})",.na=''))(.x)
    )
  ) %>%
  select(
    `Sequence ID`=id,species,`Province ({Kulbicki})` = kulbicki_province,ecoregion,country,
    locality,`Accession ({COI})`=coi_accession,`Accession (cytb)`=cytb_accession
  )
# seq_tbl <- plagiotremus_data %>%
#   filter(source == 'sequenced') %>%
#   mutate(accession = "") %>%
#   group_by(species,kulbicki_province,ecoregion,country,locality) %>%
#   summarise(accession="",n = n()) %>%
#   ungroup() %>%
#   mutate(species = str_glue('***{species}***')) %>%
#   group_by(species) %>%
#   mutate(across(kulbicki_province:country,~if_else(duplicated(.x),"",.x))) %>%
#   ungroup() %>%
#   mutate(species = if_else(duplicated(species),"",species)) %>%
#   dplyr::rename(`Province ({Kulbicki})` = kulbicki_province, `{GenBank} Accession` = accession)

print_table(seq_tbl,"tbl_supp_seqs","Detailed locality metadata for specimens collected in this study.",hdr_case = 'sentence')
```

\newpage

```{r tbl-supp-bold-seqs,results="asis",dependson="bold-seqs"}
bold_seq_tbl <- bold_seqs$coi$samples %>%
  select(species,processid,bin_uri,genbank_accession,country,catalognum) %>%
  dplyr::rename(
    Species=species,
    `Process ID`=processid,
    `BIN`=bin_uri,
    `GenBank Accession`=genbank_accession,
    Country=country,
    `Catalog Number`=catalognum
  ) %>%
  arrange(Species)
print_table(bold_seq_tbl,'tbl_supp_bold_seqs','Summary of external sequences used in this study.',hdr_case=NA)
```

\newpage

```{r tbl-supp-alignment-models, results="asis", dependson="init"}
print_table(here('data','tables','models.csv'),'tbl_supp_models','Summary of best-fit paritioning models by COI alignment coverage')
```

\newpage

```{r tbl-supp-dist-pgt-coi, results="asis", dependson="dna-distances"}
print_table(dna_distances$pgt$coi$tbl,id="tbl_supp_dist_pgt_coi","Mean COI Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus goslinei*/*P. tapeinosoma*. Abbreviations: RS: Red Sea, WIO: Western Indian Ocean, CIP: Central Indo-Pacific, CP: Central Pacific, PL: Polynesian Province, MHI: Main Hawaiian Islands, NWHI: Northwestern Hawaiian Islands, MQ: Marquesas.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-supp-dist-pgt-cytb, results="asis", dependson="dna-distances"}
print_table(dna_distances$pgt$cytb$tbl,id="tbl_supp_dist_pgt_cytb","Mean cytb Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus goslinei*/*P. tapeinosoma*.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```

\newpage

```{r tbl-supp-dist-per-coi, results="asis", dependson="dna-distances"}
print_table(dna_distances$per$coi$tbl,id="tbl_supp_dist_per_coi","Mean COI Kimura 2-parameter (K2P) distances among biogeographic provinces for *Plagiotremus ewaensis*/*P. rhinorhynchos*.",accuracy=0.001,hdr_case = 'none',roundr=3,na='---')
```
